### 2012-07-24 paper/paper.tex
inizio paper

\documentclass[11pt]{article}
\usepackage[latin1,utf8x]{inputenc}
\usepackage{color}

\title{Titolo da decidere}
\author{authors}
\date{date}
\begin{document}
\maketitle
\begin{abstract}
This is a very short abstract.
\end{abstract}

\section{Introduction}

\textcolor{blue}{
Process mining is a recent discipline that includes tecniques for process analysis and discovery. Today, many Information Systems that integrate the concept of business process, provide registration of events that occur during the execution of the process activities. Starting from recorded executions related on the same process, the goal of the existing process mining techniques is extracting data to discover automatically a process model or, if the model is already exists, to check the executions in term of conformance and performance.\\
\newline
The idea that ispires this work consists in developing un approach for exploiting the huge ammount of data recorded at the process activities by information systems. More precisely, our purpose is to find pattern on data that could influence the process behaviour. For this reason, the work presented in this article is based on some concepts originated to Machine Learning and Data Mining disciplines. Should be noted that using data mining for similar aims has already been explored by Van der Aalst in \cite{}, the idea here consists in discovering how data attributes may influence the routing of case. Instead, in this contribution the intention is to present an approach based on data mining techniques also, but for discovering data dependencies on the process conformance and performance. Finally, an another goal that we propose is providing a predective model to detect the conformance result and some performance metrics for new process executions.  \\}
\newline
\textcolor{blue}{In front of a very large number of process instances, all recorded in event log, it is essential to have an automatic tool for extracting usefull information in analysing the real behaviour of the process. In order to meet this requirement, in this article we present an approach based on classification and decision trees to identify rules that could influence the process in term of conformance to the existing model, and performance metrics.\\}
%\newline
%\textcolor{blue}{
%}

%Il process mining comprende una serie di tecniche che trattano processi vigenti in contesti organizzativi. Attualmente, moltissimi sistemi che integrano il concetto di processo di business a loro interno, provvedono alla registrazione degli eventi che si verificano durante l'esecuzione delle attività previste dal processo. Partendo da un insieme di esecuzioni registrate e riguardanti un unico processo, è possibile applicare una serie di tecniche volte alla scoperta del modello di processo oppure alla sua analisi in termini di conformance e performance. L'idea che ispira questo articolo è sviluppare un approccio in grado di sfruttare le enormi quantità di dati che vengono registrati in corrispondenza delle varie attività del processo.
%Si tratta di un approccio basato sulle tecniche di data mining capace di scoprire eventuali influenze dei dati sul comportamento del processo. L'impiego del data mining per il process mining è un'idea già presentata da Van deer Aalst in \cite{}, in questo articolo infatti si cerca di evidenziare le influenze dei dati sulla direzione del flusso di esecuzione in corrispondenza dei punti decisionali. La nostra idea invece è di sfruttare le tecniche di data mining per fornire agli analisti di porcesso un tool di supporto in grado di aiutare nell'analisi di conformance e performance del processo.\\

%Di fronte ad un numero molto grande di istanze di processo, tutte appositamente registrate in event log, avere uno strumento capace di estrarre informazioni utili per l'analisi del comportamento reale del processo diventa fondamentale. Cercando di rispondere a questa esigenza, in questo articolo è presentato un approccio capace di individuare eventuali regole che influenzano il processo in termini di conformance al modello prefissato, ed in termini di performance. Un ulteriore obiettivo che ci poniamo è fornire agli analisti uno strumento di predizione, ovvero un modello in grado di predire, in base ai dati delle nuove istanze di processo che si presentano, quale sarà l'esito di conformance e le misure di performance che caratterizzano la particolare esecuzione presa in esame. Per potere realizzare questi obiettivi ci basiamo sulla classificazione che è una particolare tecnica di data mining, vengono inoltre impiegati gli alberi di decisione come strumento di classificazione.\\

%L'articolo è organizzato in questo modo: dopo una presentazione nella sezione 2 dei vari formalismi adottati, nella sezione 3 viene presentato un semplice esempio a cui verrà fatto riferimento nell'articolo. Nella sezione 4 invece verranno introdotti alcuni concetti relativi alla conformance e performance di un processo. Infine, nella sezioni 5 e 6 si affronterà come impiegare la tecnica di classificazione per scoprire particolari pattern nei dati che possono influenzare conformance e performance del processo.

\textcolor{blue}{
The article is organized as follows. After an introduction to the different formalism used to carry out our purposes, the attention will be 
}

%\section{Formalismi e esempio }
\section{Conformance and Performance Analysis based on Petri nets}

The process analysis is performed by using the Petri net modelling the process and an event log recording the related executions (process instances). The basic building blocks of event logs are events. An event {\itshape e} can be seen as a pair {\itshape e = (a,t) } representing an action {\itshape a } recorded and the corresonding timestamp {\itshape t}. Action and timestamp are denoted respectively by $\alpha(e)$ and $\phi(e)$. Events that belong to the same process are grouped into {\itshape traces}. Formally, a trace $T$ is a finite sequence of events $T[1],..., T[n]$, such that $\phi(T[i]) \leq \phi(T[i+1])$ for all  $i \in [1,n)$. A log $L$ is a set of traces, recording the activities performed by a system during a finite number of process executions. We assume here the following hypothesis:
\begin{itemize}
\item All traces are instances of the same process.
\item For each action there exists a corresponding transition in the net that will be denoted, for simplicity, by the same name of the action.
\end{itemize}
The key algorithm expoloited to analyze the Petri net model with respect to the log is the {\itshape log repaly} algorithm. Given a Petri net model and an event log as intput to the algorithm, the output results can be used to check the conformance of traces and to evaluate some performance metrics. For each trace in the log, log replay starts by placing one token in the start place of the net. For each event in the trace the corresponding transition is fires in an {\itshape non blockin way} and the marking of the net is updated. 'Non blocking replay' means that if the log replay execution requires to fire an enabled transition, the missing tokens are created artificially. The output of the log replay of a trace can be represented as an orderd list $R$ of pairs $(tr, i)$, representing that the transition $tr$ has been fired to mimic the event $T[i]$.
\\

In general, there could be exist some transition in the net called {\itshape invisible}. This can happen if the transition models an internal choice that is not visible in the system, or it is used to implement a construct of a more abstract modelling language (BPMN or others). Usually, invisible transitions are considered to be lazy, i.e, when firing a visible transition $t$ corrisponding to an event of the trace, only then the invisibile transition enabling $t$ is fired. However, in this paper, we adopt an other way for handling these transitions and the firing is performed as soon as possible. This allows to carry out some performance measures as explained that ar not possible with the lazy metohode as explained in \cite.\\

The result of log repaly can be used to evaluate conformance and performance of the Petri net model. Conformance problems can be discovered by analyzing the tokens tha have been artificially created {\itshape (the missing tokens)} and the tokens tha were not consumed {\itshape (the remaining tokens)}.
\end {document}

### 2012-08-02 paper/paper.tex
analisi conformance con esempio

@@
 \usepackage{color}
+\usepackage[pdftex]{graphicx}
+\usepackage{amsmath, amsfonts, amssymb, amsthm}
 
@@
 \newline
-\textcolor{blue}{In front of a very large number of process instances, all recorded in event log, it is essential to have an automatic tool for extracting usefull information in analysing the real behaviour of the process.
+\textcolor{blue}{In front of a very large number of process instances, all recorded in event log, it is essential to have an automatic tool for extracting usefull information for the analysis of the real behaviour of the process.
 In order to meet this requirement, in this article we present an approach based on classification and decision trees to identify rules that could influence the process in term of conformance to the existing model, and performance metrics.\\}
@@
 \textcolor{blue}{
-The article is organized as follows. After an introduction to the different formalism used to carry out our purposes, the attention will be 
+The article is organized as follows. After an introduction to the different formalism used to carry out our purposes, the attention will be...
 }
 
-%\section{Formalismi e esempio }
+\section{Example}\label{example}
+L'esmepio scelto si ispira ad una procedura che trova applicazione all'interno di organizzazioni a carattere commerciale, si tratta del processo di vendita.
+In questa sede viene considerata solo una sua semplificazione rispetto a quanto avviene realmente.
+Per noi, il processo di vendita inizia con l'attività 'notify', in questa fase viene accolto un ordine da parte di un cliente.
+Dopo la ricezione dell'ordine, vengono condotte alcune attività in parallelo (visto che non presentano alcuna dipendenza tra loro) una di queste è l'attività 'FinancialCheck', in questa fase viene effettuata una valutazione dell'ordine dal punto di vista finanziario, possiamo ad esmepio pensare che vengono condotte delle indagini sulla solvibilità del cliente che ha emesso l'ordine.
+Parallelamente viene svolta l'attività 'ExternalWarehouse'...
+
 \section{Conformance and Performance Analysis based on Petri nets}
@@
 For each event in the trace the corresponding transition is fires in an {\itshape non blockin way} and the marking of the net is updated.
-'Non blocking replay' means that if the log replay execution requires to fire an enabled transition, the missing tokens are created artificially.
+``Non blocking replay'' means that if the log replay execution requires to fire an enabled transition, the missing tokens are created artificially.
 The output of the log replay of a trace can be represented as an orderd list $R$ of pairs $(tr, i)$, representing that the transition $tr$ has been fired to mimic the event $T[i]$.
@@
 The result of log repaly can be used to evaluate conformance and performance of the Petri net model.
-Conformance problems can be discovered by analyzing the tokens tha have been artificially created {\itshape (the missing tokens)} and the tokens tha were not consumed {\itshape (the remaining tokens)}.
+Conformance problems can be discovered by analyzing the tokens that have been artificially created {\itshape (the missing tokens)} and the tokens tha were not consumed {\itshape (the remaining tokens)}.\\
+
+To clarify the conformance analysis technique we exploit the Petri net presented in section \ref{example} with reference to tow traces , $T$ and $T'$, presented respectively in figures \ref{ConfLog} and \ref{NonConfLog}.
+The log replay execution of the trace $T$, which is compliant with the Petri, terminates with a marking containing one token in the end place \{${p14 \rightarrow 1}$\}, and returns the sequence:
+\begin{equation}
+\begin{split}
+R&=\{(Order\_start,1), (Order\_complete,2),(WarehouseCheck\_start,3), \\
+& (FinancialCheck\_start,4),(WarehouseCheck\_complete,5),\\
+& (FinancialCheck\_complete,6),(t10, 6), (Notify\_start,7), (Notify\_complete,8)\}
+\end{split}
+\end{equation}
+The log replay execution of the trace $T'$ terminates with remaing tokens \{${p8 \rightarrow 1,p14 \rightarrow 1}$\} and a missing token \{${p10 \rightarrow 1}$\}, the sequence returned is:
+\begin{equation}
+\begin{split}
+R'&=\{(Order\_start,1), (Order\_complete,2), (WarehouseCheck\_start), \\
+& (FinancialCheck\_start,3), (WarehouseCheck\_start,4),(t7,5), (t10,6)\\
+& (Notify\_start,6), (Notify\_complete,7)\}
+\end{split}
+\end{equation}
+
+\begin{figure}[H]\label{ConfLog}
+\centering
+\includegraphics[width=400pt,height=40pt]
+{./items/logConforme.pdf}
+\caption{Conforme log example}
+\end{figure}
+
+\begin{figure}[H]\label{NonConfLog}
+\centering
+\includegraphics[width=400pt,height=40pt]
+{./items/logNonConforme.pdf}
+\caption{Non conforme log example}
+\end{figure}
+
 \end {document}

### 2012-08-02 src/org/processmining/dataminig/performance/plugin/DMPerfPlugin.java

@@ -228,14 +228,14 @@ public class DMPerfPlugin {
 			  	//performance
 			 
 			 List<PerformanceResult> Perf = performance.getListperformance();
-			 System.out.println(Perf.size());
+			 //System.out.println(Perf.size());
 			 PerformanceResult trPerf= 	Perf.get(log.indexOf(tr));
 			 Map<Place, PerformanceData> PlaceData = trPerf.getList();
 			 List<Float> synchTimes = new Vector<Float>();
 			 for( Place p : PlaceData.keySet() ){
 				 float synch = PlaceData.get(p).getSynchTime();
 				 if (synch != 0 )
-				 	 synchTimes.add(new Float(synch));				 
+				 	 synchTimes.add(new Float(synch * 0.001));				 
 				/* if( synch < 0 )
 						 synchTimes.add( new Float(- synch));				  
 			 	*/
@@ -246,22 +246,21 @@ public class DMPerfPlugin {
 					 maxTime = time.floatValue();
 			 }
 			 
-			 System.out.println("**********TOSRING********");
+			 System.out.println("******************");
 			 System.out.println(maxTime);
-			 System.out.println("*************************");
+			 System.out.println("******************");
 			 
-			 if(maxTime > 1000 ) 
+			 if(maxTime  > 10800 ) 
 				 instance.setValue(wekaAttrindex, "HIGH");
 			 else
 				 instance.setValue(wekaAttrindex, "LOW");
 			 
-			 wekaAttrindex++;
 		  		
 		  	instances.add(instance);
-			  	/*System.out.println("**************TOSRING**********");
-			  	performance.toString();
-			  	System.out.println("************************");
-*/
+			  	
+		  	/*System.out.println("**************TOSRING**********");
+			performance.toString();
+		  	System.out.println("************************");*/
 		  }
 		 
 		 return instances;		

### 2012-08-04 paper/paper.tex
analisi perf con esempio

@@
 \section{Example}\label{example}
+
+\begin{figure}[t]\label{ConfLog}
+\includegraphics[width=360pt]
+{./items/Sales_PN.pdf}
+\caption{Petri net}
+\end{figure}
+
 L'esmepio scelto si ispira ad una procedura che trova applicazione all'interno di organizzazioni a carattere commerciale, si tratta del processo di vendita.
@@
 However, in this paper, we adopt an other way for handling these transitions and the firing is performed as soon as possible.
-This allows to carry out some performance measures as explained that ar not possible with the lazy metohode as explained in \cite.\\
+This allows to carry out some performance measures as explained that are not possible with the lazy metohode as explained in \cite.(riferimento al paper Applying Process Analysis to the italian eGiv Entreprise Architecture)\\
 
@@
 To clarify the conformance analysis technique we exploit the Petri net presented in section \ref{example} with reference to tow traces , $T$ and $T'$, presented respectively in figures \ref{ConfLog} and \ref{NonConfLog}.
-The log replay execution of the trace $T$, which is compliant with the Petri, terminates with a marking containing one token in the end place \{${p14 \rightarrow 1}$\}, and returns the sequence:
+The log replay execution of the trace $T$, which is compliant with the Petri net, terminates with a marking containing one token in the end place \{${p14 \rightarrow 1}$\}, and returns the sequence:
 \begin{equation}
@@
 \end{equation}
-The log replay execution of the trace $T'$ terminates with remaing tokens \{${p8 \rightarrow 1,p14 \rightarrow 1}$\} and a missing token \{${p10 \rightarrow 1}$\}, the sequence returned is:
+The log replay execution of the trace $T'$ terminates with remaing tokens \{${p8 \rightarrow 1,p14 \rightarrow 1}$\} and a missing token \{${p10 \rightarrow 1}$\}.
+The missing token is created artificially and this fact records a wrong execution of the event $Notify\_start$.
+In fact, notice that in the trace $T'$ this event is executed before the termination of the activity ``FinancialCheck'', and this is interpreted as a non conformance to the process model.
+The sequence returned by log replay is:
 \begin{equation}
@@
 
-\begin{figure}[H]\label{ConfLog}
+\begin{figure}[t]\label{ConfLog}
 \centering
@@
 
-\begin{figure}[H]\label{NonConfLog}
+\begin{figure}[t]\label{NonConfLog}
 \centering
@@
 
+Since the logs contain timestamps, the log replay can be used to compute performance measures of the process.
+The idea is to calculate the time interval between production and consumption of tokens in each place.
+This tecniques can be applied only to traces that do not present missing token during the replay, because such tokens cannot have time information.
+During the log replay the following metrics can be computed for catch trace and each place:
+\begin{itemize}
+\item sejour time $(tsj)$ : the time interval between arrival and depature of tokens;
+\item synchronization time $(tsc)$: the time interval between arrival of a token in the place and enabling of a transition in the post-set of the place;
+\item waiting time $(tw)$:  the time interval between enabling of a transition in the post-set of the place and token departure (thus $tsj=tsc+tw $).
+\end{itemize}
+
+To clarify the metrics evaluated by this tecnique, as done with conformance analysis, we exploit the Petri net  presented in section \ref{example} and the trace T presented in figure \ref{ConfLog}.
+The replay starts with firing the transition $Order\_start$ at time $0s$, thus a token arrives at $p2$ at time $0s$.
+After $45 min$ the token in $p2$ is consumed and the transition $Order\_complete$ is fired, so $tw(p2)=tsj(p2)=45min$, and a token arrives at $p3$ and $p4$.
+At the time $1h:5min:25s$ the transition $WarhouseCheck\_start$ is fired, thus $tw(p4)=20min+25sec$ and a token is produced at $p6$.
+At the time $1h:10min:20s$ the transition $FinancialCheck\_start$ is fired and a token is produced at $p5$, so $tw(p3)=25min+20s$.
+After $2h:30min:20s$ from the replay start the transition $Financial\_complete$ is fired, hence we have $tw(p5)=1h+20min$ and a token is produced at $p11$.
+At time $3h:10min:20s$ the firing of the transition $WarehouseCheck\_complete$ is executed, so $tw(p6)=2h+4min+55s$ and token is produced at $p7$ and $p8$.
+At the same time the invisible transition $t10$ is fired and a token is produced at $p10$.
+At this point the transition $t10$ is enabled, so firing it is now possible and a token is produced at $p12$.
+Notice that $tsc(p11)=40min, tsc(p10)=0s$.
+At the time $3h:25min:20s$ the transition $Notify\_start$ is fired and a token is produced in $p13$, after $4min+4sec$ the trasition $Notify\_complete$ is fired and $tw(p13) = 4min+40s$.
+
 \end {document}

### 2012-09-20 paper/paper.tex
aggiunta petri net
aggiornamenti paper

@@
 
-\textcolor{blue}{
 Process mining is a recent discipline that includes tecniques for process analysis and discovery.
@@
 Finally, an another goal that we propose is providing a predective model to detect the conformance result and some performance metrics for new process executions.
-\\}
-\newline
-\textcolor{blue}{In front of a very large number of process instances, all recorded in event log, it is essential to have an automatic tool for extracting usefull information for the analysis of the real behaviour of the process.
-In order to meet this requirement, in this article we present an approach based on classification and decision trees to identify rules that could influence the process in term of conformance to the existing model, and performance metrics.\\}
+\\
+
+In front of a very large number of process instances, all recorded in event log, it is essential to have an automatic tool for extracting usefull information for the analysis of the real behaviour of the process.
+In order to meet this requirement, in this article we present an approach based on classification and decision trees to identify rules that could influence the process in term of conformance to the existing model, and performance metrics.\\
 %\newline
@@
 
-\textcolor{blue}{
 The article is organized as follows. After an introduction to the different formalism used to carry out our purposes, the attention will be...
-}
 
 \section{Example}\label{example}
-
-\begin{figure}[t]\label{ConfLog}
+\begin{figure}[h]\label{pnet}
 \includegraphics[width=360pt]
@@
 
-\section{Conformance and Performance Analysis based on Petri nets}
+\section{Conformance and Performance Analysis based on Petri nets}\label{Background}
 
@@
 However, in this paper, we adopt an other way for handling these transitions and the firing is performed as soon as possible.
-This allows to carry out some performance measures as explained that are not possible with the lazy metohode as explained in \cite.(riferimento al paper Applying Process Analysis to the italian eGiv Entreprise Architecture)\\
+This allows to carry out some performance measures as explained that are not possible with the lazy metohode as explained in \cite.(riferimento al paper Applying Process Analysis to the italian eGov Entreprise Architecture)\\
 
@@
 
-\begin{figure}[t]\label{NonConfLog}
+\begin{figure}[h]\label{NonConfLog}
 \centering
@@
 The replay starts with firing the transition $Order\_start$ at time $0s$, thus a token arrives at $p2$ at time $0s$.
-After $45 min$ the token in $p2$ is consumed and the transition $Order\_complete$ is fired, so $tw(p2)=tsj(p2)=45min$, and a token arrives at $p3$ and $p4$.
+After $45 min$ the token in $p2$ is consumed and the transition $Order\_complete$ is fired, so $tw(p2)=45min$, and a token arrives at $p3$ and $p4$.
 At the time $1h:5min:25s$ the transition $WarhouseCheck\_start$ is fired, thus $tw(p4)=20min+25sec$ and a token is produced at $p6$.
@@
 
+An important performance measure is the activity execution time.
+This can be deducted from the waiting time in the places between the start and the activity complete transitions.
+For example, time execution for ``FinancialCheck'' is given by the waiting time computed for the place $p5: tw(p5)=1h+20min$.\\
+
+It is worth noting that the synchronization time is greater than zero only for places which contain on their post-set transitions depending from other places.
+For the Petri net in figure \ref{pnet}, the only places which can have a synchronization time not null are $p10$ and $p7$.
+For instance, the sycronization time is greater than zero for $p10$ in the trace in figure \ref{ConfLog}.
+That means that, for this particular instance of the process, the branch with the occurance of the activity ``FinancialCheck'' is faster than the branch with activities relating the warehouse.
+
+\section{An approach based on classification for conformance analysis}
+%Supponiamo di avere un numero altissimo di esecuzioni relative al medesimo processo registrate in event log.
+Supponiamo anche che tutte queste esecuzioni siano state analizzate dal punto di vista della conformance per mezzo del log replay.
+L'algoritmo di analisi log replay si limita ad individuare come mostrato in sezione \ref{Background} quali sono i punti in cui si verificano errori di conformance, tuttavia, esso non fornisce alcuna informazione sulle cause che generano queste anomalie.\\
+
+%Riuscire ad individuare anomalie è uno degli obiettivo fondamentali nell'analisi dei processi, avere un'idea su quali sono le cause che li generano lo è ancora di più.
+Questo infatti permette di capire anche come agire per apportare eventuali correzioni.
+
+%Perciò l'idea che ci ha guidato in questo lavoro consiste nel riuscire ad individuare nei dati contenuti negli event log, eventuali pattern che possono influenzare gli scostamenti rispetto al modello prestabilito.
+Questo infatti potrebbe aiutare nell'individuare le cause che stanno dietro agli scostamenti del comportamento reale del processo rispetto a quello atteso.
+
+Employing data mining techniques for process analysis was already experimented by Van Deer Aalst in \cite{decision mining in business process}.
+In this article \cite{}, the idea consists in discovering how data attributes may influence the routing of case.
+The motivation that encourages this work consists in the presence of a huge ammount of data recorded in the event logs in the form of attributes that could contain implicit information.
+In this contribution, the same methodology is adopted to offer an additional tool for process analysts.\\
+
+With the approach presented here, the goal proposed is discovering patterns or rules in the event logs data in corrispondence of which conformance errors, discussed in section \ref{Background}, occur.
+The goal is to conduct an investigation into data to discover the cause of conformance anomalies.
+Given the enormous quantity of data, the analysis is carried out with the classification, one of data mining techniques.
+In a classification problem, data is represented by a collection of records (called instances too), each of them is caracterized by a tuple $(\mathbf{x},y)$, where $\mathbf{x}$ is an attribute set, while $y$ is a special attribute called ``target attribute'' denoting the belowing class of the record.
+With this technique, the idea is learning a classification model that maps each attribute set $x$ to one of the predefined class labels $y$.
+See \cite{} for more information about classification.\\
+
+Analysing data arising from event logs can be transformed into a classification problem.
+The data set can be derived from the process data, in particular, each instance data determines a record of the data set.
+For the process presented in section \ref{example}, a record is charaterized by the following attributes set: an instance identifier, a client identifier, the client typology with two possible values: new client and consolidated client, the sales manager responsabile for the order, the chief financial officier name who conducts the financial evaluation activities, the warehouseman name for the warehouse check activities, a supplier name in case of an external warehouse acitivity, and finally the result comunicated to the client for the order issued.
+The conformance result for the process instance provides an additional and important data which takes the role of the target attribute.
+All these attributes are discrete and contribute to formulate the data set presented in table \ref{tab:SaleData} of the classification problem.\\
+
+\begin{table}[!h]
+\scriptsize
+\centering
+\begin{tabular}{|p{0,8cm}|p{0,7cm}|p{1,5cm}|p{1,2cm}|p{1,2cm}|p{1,2cm}|p{0.8cm}|p{1,2cm}|p{0,5cm}|}
+\hline OrdIde & CltIde & CltType & SalMan & FinOff & WrhsMan & SplIde & OrdResut & Conf\\
+\hline
+1 & 20 & consolidate & Marc & Alexander & Alex & 15 & positive & yes\\
+\hline
+2 & 700 & new & Johann & Mario & Alex & 04 & positive & yes\\
+\hline
+3 & 10 &consolidate & Mary & Robert & Alex & 15 & negative & yes\\
+\hline
+... & ... & ... & ... & ... & ... & ... & .... & .... \\
+\hline
+... & ... & ... & ... & ... & ... & ... & .... & ...  \\
+\hline
+\end{tabular}
+\scriptsize
+\caption{Data set}
+\label{tab:SaleData}
+\end{table}
+\normalsize
+
+Many classification models could be used for a classification problem.
+For our purpose we choose the decision tree model because of its simplicity and diffusion.
+In a decision tree, each leaf node is assigned a class label (a values of the attribute target).
+The non-terminal nodes, which include the root and other internal nodes, contain attribute test conditions to separate records that have different characteristics.
+In principle, there are exponentially many decision trees that can be constructed from a given set of attributes.
+While some of the trees are more accurate than others, finding the optimal tree is computationally infeasible because of the exponential size of the search space.
+However, efficient algoritms have been developed to induce a reasonable accurate decision tree in a a reasonable ammount of time.
+These algorithm usually employ a greedy strategy that grows a decision tree by making a series of decisions about which attribute to use for partioning the data.
+One is Hunt's algorithm, which is the basic of many existing desicion tree induction algorithms, including $C4.5$ used for our purpose.\\
+
+Using existing data mining tools for classification, it is possible to build a decistion tree as a classification model for the example presented here, based on data produced artificially just for examplary purposes: from the event log recorded, the attributes that characterized the process are extracted, and all the resources needed for the classification algorithm employed are produced.
+The decision tree resulted is presented in figure \ref{salesDecTree}.\\
+
+\begin{figure}[h]\label{salesDecTree}
+\centering
+\includegraphics[width=230pt,height=110pt]
+{./items/Sales_tree.pdf}
+\caption{decision tree}
+\end{figure}
+The decision tree returned descibes a data pattern in corrispondence of which a process instance could present conformance errors: order managed by the sales manager ``Mary'' and received from consolidate clients of the organization may not respect the standard sales procedure.
+However, to get information most significant for our analysis, it is usefull to relate what is noticed by the decision tree and the log replay results.
+In figure \ref{replayResult} is presented a Petri net that resumes the conformance analysis conducted on instances recorded in the event log taken in exam.
+Arcs are labled by the number of activations done, red colored places presented missing or remaning tokens and colored transitions notice a wrong execution.\\
+\begin{figure}[h]\label{replayResult}
+\includegraphics[width=390pt]
+{./items/Sales_PN_result.pdf}
+\caption{Petri net}
+\end{figure}
+
+The Petri net reporting the log replay results notice that there are 37 of 250 process executions with conformance erros, that is proved by the 37 missing tokens on the place $p12$ and a wrong execution of the transition ``Notify\_start''.
+In order to give an interpretation for this results, it must be taken into account the non blocking behaviour of the log replay.
+In this case, the 37 missing tokens in $p12$ notice that during the replay the algorithm needs to mimic the event ``Notify\_start'' but the transition associated to the event is not enabled in the marking reached by the net, so the algorithm creates artificial tokens for completing the trace repaly.
+This situation is possible only if the invisible transition $t10$ is not enabled.
+Now, if $t10$ is not enabled that means that there are no tokens in the place $p10$ and $p7$.
+But notice that there are 21 remaining tokens in the place $p10$, so the transition $t10$ is not enabled for those $37$ instances because the place $p7$ does not present tokens.
+Moreover, the Petri net in figure highlights 37 remaing tokens in the place $p3$, so we can deduce that in those instances financial activities were not performed.
+From this facts, we can conclude that the conformance problems presented in  37 instances are due to a missing execution of the Financial Activities check.
+On the other hand, the decision tree reports that the instances with conformance problems are made by consolidate clients.
+In this case, it might be reasonable that, for the orders made by the consolidate clients of the organization, could not be necessary a financial valutation of their situation.\\
+
+Analysing the log replay results tanking into consideration patterns discoverd with classification technique leads to discover new cases to include in the process model by an extension.
+In other cases, if the process model is a representation of a fixed procedure to be respected, what discovered with such analysis can be used to take corrective measures.
+In fact, through a combination of the log replay results and the data patterns rules it may be possible to discover conformance errors cause, and this is fundamental to bring the right corrections.
+For instance, in the example presented here, if the model process represents a sales procedure that must be respected in any case, an appropriate correction can be taken to correct the wrong behaviour of the staff about consolidate clients.
+
 \end {document}

### 2012-09-20 logenerator/genSalesConf.py

@@ -1,0 +1,178 @@
+#!/usr/bin/python
+# -*- coding: latin-1 -*-
+
+from generator import Sequence, Choice, Entry, Par
+from generator import generate
+from generator import defaultActivityHook
+
+import random
+
+
+# def docActivityHook(startTime, attrs):
+#     t = startTime
+
+
+#     # delta1 = random.randint(1, 30*60)		#sceglie un tempo da 0 a 30 minuti (convertito in secondi). Tempo di inizio
+#     delta1 = 15
+   
+#     #delta2 = random.randint(delta1+1, 59*60)	sceglie un tempo tra delta1 e un'ora dopo (convertito in secondi). Tempo di fine
+
+#     if( attrs["destinazione"] == "EXTRAEU" ):
+# 	delta2 = (delta1+1) + (45*60)
+#     else:
+# 	delta2 = (delta1+1) + (25*60)
+
+#     t0 = t+delta1				# tempo di inizio dato dal startTime più delta1
+#     t1 = t+delta2				# tempo di fine dato dal starTime più delta2.
+
+#     return t0, t1
+
+# def bagActivityHook(startTime,attrs):
+	
+# 	t = startTime
+
+#     	#delta1 = random.randint(1, 30*60)		#sceglie un tempo da 0 a 30 minuti (convertito in secondi). Tempo di inizio
+# 	delta1 = 15
+	
+# 	delta2 = (delta1+1) + (20*60)  
+
+# 	t0 = t+delta1				# tempo di inizio dato dal startTime più delta1
+#     	t1 = t+delta2				# tempo di fine dato dal starTime più delta2.
+
+#     	return t0, t1
+
+# def docHook(t, attrs):
+
+# 	t0, t1 = docActivityHook(t,attrs)
+
+# 	return t0,t1
+
+# def bagHook(t,attrs):
+
+# 	t0, t1 = bagActivityHook(t,attrs)
+	
+# 	return t0, t1
+
+def orderHook(t, attrs):
+    products = ["bottles", "boxes", "equipment"]
+    attrs["product"] = products[random.randint(0, len(products)-1)]
+    customers = ["A", "B", "C"]
+    clientTypes = ["consolidate", "new"]
+    salesmanagers = ["Marc", "Johann", "Mary"]
+    financialOfficiers = ["Alexander" , "Robert", "Mario"]
+    warehousemen = ["Alex"]
+    orderResults = ["positive", "negative"]
+
+
+    attrs["customer"] = customers[random.randint(0, len(customers)-1)]
[…]

### 2012-09-28 paper/paper.tex
mod paper

@@
 \section{Example}\label{example}
+To clarify what presented in this article the example illustrated in this section will be taken into exam whenever is needed.\\
+
+Commercial organizations usually follow a specific procedure to manage orders received form various clients.
+In general this procedure represents a business process that includes many differents activities, each of theme can involves diffent departments of the organization.
+Let us consider a semplifacation of this procedure.
+For us, the sale process begins with receving an order notification from a client, so the first process activity is simply called ``Order''.
+After the order receipt the sale process continues with some activities that can be done in a parallel way (since they do not present any dependencies).
+One of thes is the ``FianancialCheck'' activity during which the order is analysed from a financial point of view (fo example verifying the client situation and solvency).
+In parallel an activity called ``WarehouseCheck'' is performed.
+In this phase a check regarding the merchandise requested by the order is done, and in case of a shortage an external order is issued for a supplying, this is done through the activity ``ExternalWarehouse''.
+After the completion of the  parallel activities, a synchronization is needed before the continuation of the sale procedure.
+In the semplification considered here the sale process terminates with the activity ``Notify'' in which a result regarding the acceptance of the order is comunicated to the client.\\
+
+The are many differents formalisms for modeling business process, the most popolar are based on flowcharts because of their expressivness (BPMN fo instance).
+For our purposes, a process is represented by a Petri net which is a matematical model so rigorous and better suited to process mining analysis.
+It is worth noting that they exist some techniques that allow traduction of a model espressed in a flowchart to a model espressed with a Petri net, more information can be found in \cite{}.
+The petri net representing the sale process presented in this section is in figure \ref{example}.\\
+
 \begin{figure}[h]\label{pnet}
@@
 \end{figure}
-
-L'esmepio scelto si ispira ad una procedura che trova applicazione all'interno di organizzazioni a carattere commerciale, si tratta del processo di vendita.
-In questa sede viene considerata solo una sua semplificazione rispetto a quanto avviene realmente.
-Per noi, il processo di vendita inizia con l'attività 'notify', in questa fase viene accolto un ordine da parte di un cliente.
-Dopo la ricezione dell'ordine, vengono condotte alcune attività in parallelo (visto che non presentano alcuna dipendenza tra loro) una di queste è l'attività 'FinancialCheck', in questa fase viene effettuata una valutazione dell'ordine dal punto di vista finanziario, possiamo ad esmepio pensare che vengono condotte delle indagini sulla solvibilità del cliente che ha emesso l'ordine.
-Parallelamente viene svolta l'attività 'ExternalWarehouse'...
 
@@
 
-\section{An approach based on classification for conformance analysis}
+\section{An approach based on classification for conformance analysis}\label{ClassConformance}
 %Supponiamo di avere un numero altissimo di esecuzioni relative al medesimo processo registrate in event log.
@@
 
+COMMENTO: manca un pezzo che spiega come il classificatore puo' essere usato in senso predittivo aumentando la potenza del metodo. Mostrare a titolo di esempio che vantaggio ne possiamo trarre.
+
+\section{An approach based on classification for performance analysis}\label{ClassPerf}
+
+The approach presented in section \ref{ClassConformance} can be used also for performance analysis.
+Using classification tecnique for discovering how data attribute can influence process performance provides usefull information in analysing and optimazing the process services.
+For example, discovering data patterns in corrispondence of which some activities need more time for completion than others, helps in making decision about resources distribution to various process activities or in scheduling activities.\\
+
+In addition to the execution time, classification can be used to discover information about more complex performance metrics such as the syncronization time.
+Processes with parallel activities and syncronizations can often present activities that represent a bottle neck, this leads to increase waiting time of some activities and consequently the completion time of the entire process.
+In this context, with the approach presented in this article, the objective is discovering rules in the process data that influence the syncronization time.
+The model builded by classification techinque can be used also in a predicting way, and this is certainly very important in order to optimize performance execution of the process.\\
+
+With reference to the example taken in exam for conformance analysis, in this paragraph the approach based on classification is illustrated for the performance analysis.
 \end {document}

### 2012-12-27 paper/paper.tex
last mod to paper

--- 2012-08-04 paper/paper.tex
@@
-\documentclass[11pt]{article}
+%\documentclass[11pt]{article}
+\documentclass{llncs}
 \usepackage[latin1,utf8x]{inputenc}
@@
 \usepackage[pdftex]{graphicx}
-\usepackage{amsmath, amsfonts, amssymb, amsthm}
+%\usepackage{amsmath, amsfonts, amssymb, amsthm}
+\usepackage{amsmath}
 
@@
 
-\textcolor{blue}{
-Process mining is a recent discipline that includes tecniques for process analysis and discovery.
-Today, many Information Systems that integrate the concept of business process, provide registration of events that occur during the execution of the process activities.
-Starting from recorded executions related on the same process, the goal of the existing process mining techniques is extracting data to discover automatically a process model or, if the model is already exists, to check the executions in term of conformance and performance.\\
-\newline
-The idea that ispires this work consists in developing un approach for exploiting the huge ammount of data recorded at the process activities by information systems.
+Process mining is a recent discipline that includes techniques for process analysis and discovery.
+Today, many Information Systems that integrate the concept of business process provide registration of events that occur during the process activities.
+Starting from recorded executions related to the same process, the goal of the process mining techniques is extracting data to discover automatically a process model or, if the model already exists, to check the executions in term of conformance and performance.\\
+
+The idea of this work consists in developing an approach for exploiting the huge ammount of data recorded in the process activities by the information systems.
 More precisely, our purpose is to find pattern on data that could influence the process behaviour.
-For this reason, the work presented in this article is based on some concepts originated to Machine Learning and Data Mining disciplines.
-Should be noted that using data mining for similar aims has already been explored by Van der Aalst in \cite{}, the idea here consists in discovering how data attributes may influence the routing of case.
-Instead, in this contribution the intention is to present an approach based on data mining techniques also, but for discovering data dependencies on the process conformance and performance.
-Finally, an another goal that we propose is providing a predective model to detect the conformance result and some performance metrics for new process executions.
-\\}
-\newline
-\textcolor{blue}{In front of a very large number of process instances, all recorded in event log, it is essential to have an automatic tool for extracting usefull information for the analysis of the real behaviour of the process.
-In order to meet this requirement, in this article we present an approach based on classification and decision trees to identify rules that could influence the process in term of conformance to the existing model, and performance metrics.\\}
+For this reason, the work presented in this paper is based on some concepts of Machine Learning and Data Mining.
+Should be noted that using data mining for similar aims has already been explored by Van der Aalst in \cite{}, the idea there consists in discovering how data attributes may influence the routing of cases.
+Instead, in this contribution the intention is to present an approach based on data mining techniques also, but for discovering dependencies between process data and its conformance and performance.
+Finally, another goal that we pursue is providing a predective model to check the conformance result and some performance metrics for the future process executions.
+\\
+
+%In front of a very large number of process instances, all recorded in event log, it is essential to have an automatic tool for extracting usefull information for the analysis of the real behaviour of the process.
+In order to meet this requirement, in this article we present an approach based on classification and decision trees to identify rules that could influence the process in term of conformance to the existing model, and performance metrics.\\
 %\newline
@@
 
-\textcolor{blue}{
-The article is organized as follows. After an introduction to the different formalism used to carry out our purposes, the attention will be...
-}
+%The article is organized as follows. After an introduction to the different formalism used to carry out our purposes, the attention will be...
 
 \section{Example}\label{example}
+To clarify what presented in this paper the example illustrated in this section will be taken into exam whenever is needed.\\
 
-\begin{figure}[t]\label{ConfLog}
+Commercial organizations usually follow a specific procedure to manage orders received from various clients.
+In general this procedure represents a business process that includes many different activities, each of them can involve diffent departments of the organization.
+Let us consider a semplification of this procedure.
+For us, the sale process begins with receving an order notification from a client, so the first process activity is simply called ``Order''.
+After the order is received the sale process continues with some activities that can be done in parallel (since they do not present any dependencies).
+One of these is the ``FinancialCheck'' activity during which the order is analysed from a financial point of view (for example verifying the client situation and solvency).
+In parallel an activity called ``WarehouseCheck'' is performed.
+In this phase a check regarding the merchandise requested by the order is done, and in case of a shortage an external order is issued for a supplying, this is done through the activity ``ExternalWarehouse''.
+After the completion of the  parallel activities, a synchronization is needed before continuing the sale procedure.
+In the semplification considered here the sale process terminates with the activity ``Notify'' in which a result regarding the acceptance of the order is comunicated to the client.\\
+
+There are many differents formalisms for modeling business processes, the most popolar are based on flowcharts because of their expressivness (BPMN for instance).
+For our purposes, a process is represented by a Petri net which is a mathematical model characterized by rigor and suitable for process mining analysis.
+It is worth noting that there exist some techniques that allow transforming a model expressed in a flowchart to a model expressed with a Petri net, more information can be found in \cite{}.
+The petri net representing the sale process discussed in this section is shown in figure \ref{pnet}.
+The Petri net is obtained by transforming a BPMN model in which each activity can have only tow possibile state: start state, denoting the begin of the activity, and the end state.
+
+\begin{figure}[h]
 \includegraphics[width=360pt]
@@
 \caption{Petri net}
+\label{pnet}
 \end{figure}
 
-L'esmepio scelto si ispira ad una procedura che trova applicazione all'interno di organizzazioni a carattere commerciale, si tratta del processo di vendita.
-In questa sede viene considerata solo una sua semplificazione rispetto a quanto avviene realmente.
-Per noi, il processo di vendita inizia con l'attività 'notify', in questa fase viene accolto un ordine da parte di un cliente.
-Dopo la ricezione dell'ordine, vengono condotte alcune attività in parallelo (visto che non presentano alcuna dipendenza tra loro) una di queste è l'attività 'FinancialCheck', in questa fase viene effettuata una valutazione dell'ordine dal punto di vista finanziario, possiamo ad esmepio pensare che vengono condotte delle indagini sulla solvibilità del cliente che ha emesso l'ordine.
-Parallelamente viene svolta l'attività 'ExternalWarehouse'...
+\section{Process Analysis}\label{Background}
 
-\section{Conformance and Performance Analysis based on Petri nets}
-
-The process analysis is performed by using the Petri net modelling the process and an event log recording the related executions (process instances).
+The process analysis is performed by using the Petri net modeling the process and an event log recording the related executions (process instances).
 The basic building blocks of event logs are events.
@@
 \end{itemize}
-The key algorithm expoloited to analyze the Petri net model with respect to the log is the {\itshape log repaly} algorithm.
-Given a Petri net model and an event log as intput to the algorithm, the output results can be used to check the conformance of traces and to evaluate some performance metrics.
+
+\subsection{Log replay algorithm}
+
+The key algorithm expoloited to analyze the Petri net model with respect to the log is the {\itshape log replay} algorithm \cite{} .Given a Petri net model and an event log as intput to the algorithm, the output results can be used to check the conformance of traces and to evaluate some performance metrics.
 For each trace in the log, log replay starts by placing one token in the start place of the net.
@@
 
-In general, there could be exist some transition in the net called {\itshape invisible}.
+In general, there could exist some transition in the net called {\itshape invisible}.
 This can happen if the transition models an internal choice that is not visible in the system, or it is used to implement a construct of a more abstract modelling language (BPMN or others).
-Usually, invisible transitions are considered to be lazy, i.e, when firing a visible transition $t$ corrisponding to an event of the trace, only then the invisibile transition enabling $t$ is fired.
-However, in this paper, we adopt an other way for handling these transitions and the firing is performed as soon as possible.
-This allows to carry out some performance measures as explained that are not possible with the lazy metohode as explained in \cite.(riferimento al paper Applying Process Analysis to the italian eGiv Entreprise Architecture)\\
+Usually, invisible transitions are considered to be lazy, i.e, when firing a visible transition $t$ corrisponding to an event of the trace is needed, only then the invisibile transition enabling $t$ is fired.
+However, in this paper, we adopt an other way for handling these transitions: the firing is performed as soon as possible.
+This allows to carry out some performance measures   that are not possible with the lazy methode as explained in \cite.(riferimento al paper Applying Process Analysis to the italian eGov Entreprise Architecture)\\
 
 The result of log repaly can be used to evaluate conformance and performance of the Petri net model.
-Conformance problems can be discovered by analyzing the tokens that have been artificially created {\itshape (the missing tokens)} and the tokens tha were not consumed {\itshape (the remaining tokens)}.\\
+Conformance problems can be discovered by analyzing the tokens that have been artificially created {\itshape (the missing tokens)} and the tokens that were not consumed {\itshape (the remaining tokens)}.\\
 
-To clarify the conformance analysis technique we exploit the Petri net presented in section \ref{example} with reference to tow traces , $T$ and $T'$, presented respectively in figures \ref{ConfLog} and \ref{NonConfLog}.
+\subsection{Conformance Analysis}
+To clarify the conformance analysis technique we exploit the Petri net presented in section \ref{example} with reference to tow traces, $T$ and $T'$, presented respectively in figures \ref{ConfLog} and \ref{NonConfLog}.
 The log replay execution of the trace $T$, which is compliant with the Petri net, terminates with a marking containing one token in the end place \{${p14 \rightarrow 1}$\}, and returns the sequence:
@@
 R&=\{(Order\_start,1), (Order\_complete,2),(WarehouseCheck\_start,3), \\
-& (FinancialCheck\_start,4),(WarehouseCheck\_complete,5),\\
-& (FinancialCheck\_complete,6),(t10, 6), (Notify\_start,7), (Notify\_complete,8)\}
+& (FinancialCheck\_start,4),(FinancialCheck\_complete,5),\\
+& (WarehouseCheck\_complete,6),(t5,7),(t7,7),(t10, 7), (Notify\_start,7), (Notify\_complete,8)\}
 \end{split}
 \end{equation}
+
 The log replay execution of the trace $T'$ terminates with remaing tokens \{${p8 \rightarrow 1,p14 \rightarrow 1}$\} and a missing token \{${p10 \rightarrow 1}$\}.
@@
 \begin{split}
-R'&=\{(Order\_start,1), (Order\_complete,2), (WarehouseCheck\_start), \\
-& (FinancialCheck\_start,3), (WarehouseCheck\_start,4),(t7,5), (t10,6)\\
+R'&=\{(Order\_start,1), (Order\_complete,2), (WarehouseCheck\_start,3), \\
+& (FinancialCheck\_start,4), (WarehouseCheck\_complete,5),(t5,6),(t7,6), (t10,6)\\
 & (Notify\_start,6), (Notify\_complete,7)\}
@@
 
-\begin{figure}[t]\label{ConfLog}
+\begin{figure}[t]
 \centering
@@
 \caption{Conforme log example}
+\label{ConfLog}
 \end{figure}
 
-\begin{figure}[t]\label{NonConfLog}
+\begin{figure}[h]
 \centering
@@
 \caption{Non conforme log example}
+\label{NonConfLog}
 \end{figure}
 
+\subsection{Performance analysis}
+A performance analysis of a process can also be done based on the log replay results.
 Since the logs contain timestamps, the log replay can be used to compute performance measures of the process.
 The idea is to calculate the time interval between production and consumption of tokens in each place.
-This tecniques can be applied only to traces that do not present missing token during the replay, because such tokens cannot have time information.
-During the log replay the following metrics can be computed for catch trace and each place:
+Analysing performance can be applied only to traces that do not present missing token during the replay, because such tokens cannot have time information.
+During the log replay the following metrics can be computed for each trace and each place:
 \begin{itemize}
@@
 After $2h:30min:20s$ from the replay start the transition $Financial\_complete$ is fired, hence we have $tw(p5)=1h+20min$ and a token is produced at $p11$.
-At time $3h:10min:20s$ the firing of the transition $WarehouseCheck\_complete$ is executed, so $tw(p6)=2h+4min+55s$ and token is produced at $p7$ and $p8$.
-At the same time the invisible transition $t10$ is fired and a token is produced at $p10$.
+At time $3h:10min:20s$ the firing of the transition $WarehouseCheck\_complete$ is executed, so $tw(p6)=2h+4min+55s$ and token is produced at $p7$.
+After $15 min$ the invisible transition $t7$ is fired and a token is produced at $p10$.
 At this point the transition $t10$ is enabled, so firing it is now possible and a token is produced at $p12$.
 Notice that $tsc(p11)=40min, tsc(p10)=0s$.
-At the time $3h:25min:20s$ the transition $Notify\_start$ is fired and a token is produced in $p13$, after $4min+4sec$ the trasition $Notify\_complete$ is fired and $tw(p13) = 4min+40s$.
+At the time $3h:25min:20s$ the transition $Notify\_start$ is fired and a token is produced in $p13$, after $4min+4sec$ the transition $Notify\_complete$ is fired and $tw(p13) = 4min+40s$.\\
+
+An important performance measure is the activity execution time.
+This can be deducted from the waiting time in the places between the start and the complete transitions of each activity.
+For example, time execution for ``FinancialCheck'' is given by the waiting time computed for the place $p5: tw(p5)=1h+20min$.\\
+
+It is worth noting that the synchronization time is greater than zero only for places which contain on their post-set transitions depending from other places.
+For the Petri net in figure \ref{pnet}, the only places which can have a synchronization time not null are $p10$ and $p11$.
+For instance, the sycronization time is greater than zero for $p11$ in the trace in figure \ref{ConfLog}.
+That means that, for this particular instance of the process, the branch with the activity related to the Financial checking is faster than the branch with the warehouse activities.
+
+\section{An approach based on classification for Process Analysis}\label{ClassConformance}
+%Supponiamo di avere un numero altissimo di esecuzioni relative al medesimo processo registrate in event log.
+Supponiamo anche che tutte queste esecuzioni siano state analizzate dal punto di vista della conformance per mezzo del log replay.
+L'algoritmo di analisi log replay si limita ad individuare come mostrato in sezione \ref{Background} quali sono i punti in cui si verificano errori di conformance, tuttavia, esso non fornisce alcuna informazione sulle cause che generano queste anomalie.\\
+
+%Riuscire ad individuare anomalie è uno degli obiettivo fondamentali nell'analisi dei processi, avere un'idea su quali sono le cause che li generano lo è ancora di più.
+Questo infatti permette di capire anche come agire per apportare eventuali correzioni.
+
+%Perciò l'idea che ci ha guidato in questo lavoro consiste nel riuscire ad individuare nei dati contenuti negli event log, eventuali pattern che possono influenzare gli scostamenti rispetto al modello prestabilito.
+Questo infatti potrebbe aiutare nell'individuare le cause che stanno dietro agli scostamenti del comportamento reale del processo rispetto a quello atteso.
+
+Employing data mining techniques for process analysis was already experimented by Van Deer Aalst in \cite{decision mining in business process}.
+In that paper the idea consists in discovering how data attributes may influence the routing of cases.
+This work is motivated by the presence of a huge ammount of data recorded in the event logs in the form of attributes that could contain implicit information.
+In this contribution, the same methodology is adopted to offer an additional tool for process analysis.\\
+
+With the approach presented here, our goal is to discover patterns or rules in the event logs data in corrispondence of which conformance errors, discussed in section \ref{Background}, occur.
+The goal is to conduct an investigation into data to discover the cause of conformance anomalies.
+Given the enormous quantity of data, the analysis is carried out with \emph{classification}, a classical data mining technique.
+
+\subsection{Classification: basic concepts}
+
+In a classification problem, data is represented by a collection of records (called instances), each of them is characterized by a tuple $(\mathbf{x},y)$, where $\mathbf{x}$ is an attribute set, while $y$ is a special attribute called ``target attribute'' denoting the class to which the record belongs.
+With this technique, the idea is learning a classification model that maps each attribute set $x$ to one of the predefined class labels $y$.\\
+
+A classification model is useful for the following purposes:
+\begin{itemize}
+\item Descriptive Modeling: A calssification model can serve as an explanatory tool to distinguish between objects of different classes.
+\item Predictive Modeling:  A classification model can also be used to predict the class label of unknown records.
+In fact, a classification model can be treated as a black box that automatically assigns a class label when presented with an attribute set of an unknown record.
+\end{itemize}
+
+Many classification models could be used for a classification problem.
+For our purpose we choose the decision tree model because of its simplicity and diffusion.
+In a decision tree, to each leaf node is assigned a class label (a possible value of the attribute target).
+The non-terminal nodes, which include the root and other internal nodes, contain attribute test conditions to separate records that have different characteristics.
+In principle, there are exponentially many decision trees that can be constructed from a given set of attributes.
+While some of the trees are more accurate than others, finding the optimal tree is computationally infeasible because of the exponential size of the search space.
+However, efficient algoritms have been developed to induce a reasonably accurate decision tree in a reasonable amount of time.
+These algorithms usually employ a greedy strategy that grows a decision tree by making a series of decisions about which attribute to use for partioning the data.
+For instance Hunt's algorithm is one of theme and it represents the basis of many existing decision tree induction algorithms, including $C4.5$ algorithm used for our purposes.\\
+See \cite{} for more information about classification.\\
+
+\subsection{Classification for Conformance Checking}
+Analysing data arising from event logs can be transformed into a classification problem.
+The data set can be derived from the process data, In particular, each instance data determines a record of the data set.
+For the process presented in section \ref{example}, a record is charaterized by the following attributes set: an instance identifier, a client identifier, the client typology with two possible values: new client and consolidated client, the sales manager responsabile for the order, the chief financial officier name who conducts the financial evaluation activities, the warehouseman name for the warehouse check activities, a supplier name in case of an external warehouse activity, and finally the result communicated to the client for the order issued.
+The conformance result for the process instance provides an additional and important data which takes the role of the target attribute.
+All these attributes are discrete and contribute to formulate the data set presented in table \ref{tab:SaleData} of the classification problem.\\
+
+\begin{table}[!h]
+\scriptsize
+\centering
+\begin{tabular}{|p{1cm}|p{1cm}|p{1,5cm}|p{1,2cm}|p{1,4cm}|p{1,2cm}|p{1cm}|p{1,4cm}|p{0,8cm}|}
+\hline OrdIde & CltIde & CltType & SalMan & FinOff & WrhsMan & Supplier & OrdResut & Conf\\
+\hline
+1 & 20 & consolidate & Marco & Alessandro & Alex & Gianni & positive & yes\\
+\hline
+2 & 15 & new & Anna & Mario & Alessio & Mario & positive & yes\\
+\hline
+3 & 10 &consolidate & Maria & Roberto & Alessio & Gianni & negative & yes\\
+\hline
+... & ... & ... & ... & ... & ... & ... & .... & .... \\
+\hline
+... & ... & ... & ... & ... & ... & ... & .... & ...  \\
+\hline
+\end{tabular}
+\scriptsize
+\caption{Data set}
+\label{tab:SaleData}
+\end{table}
+\normalsize
+
+Using existing data mining tools for classification, it is possible to build a decistion tree as a classification model for the example presented here.
+We do this based on data produced artificially just for exemplary purposes: from the event log recorded, the attributes that characterize the process are extracted, and all the resources needed for the classification algorithm employed are produced.
+The resulting decision tree is presented in figure \ref{salesDecTree}.\\
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=230pt,height=110pt]
+{./items/Sales_tree.pdf}
+\caption{decision tree}
+\label{salesDecTree}
+\end{figure}
+The decision tree returned describes a data pattern in correspondence of which a process instance could present conformance errors: order managed by the sales manager ``Mary'' and received from consolidate clients of the organization may not respect the standard sales procedure.
+However, to get information most significant for our analysis, it is useful to relate what is noticed by the decision tree with the log replay results.
+In figure \ref{replayResult} is presented a Petri net that resumes the conformance analysis conducted on instances recorded in the event log taken in exam.
+Arcs are labled by the number of activations done.
+Places, whenever presents some remaining or missing tokens, are labled by the number of these tokens.\\
+\begin{figure}[h]
+\includegraphics[width=390pt]
+{./items/Sales_PN_result.pdf}
+\caption{Petri net}
+\label{replayResult}
+\end{figure}
+
+The Petri net reporting the log replay results shows that there are 37 of 250 process executions with conformance erros.
+In fact, the presence of 37 missing tokens in $p12$ signals a forced execution of the transition ``Notify\_start''.
+In order to give an interpretation for this results, it must be taken into account the non blocking behaviour of the log replay.
+In this case, the 37 missing tokens in $p12$ notice that during the replay the algorithm needs to mimic the event ``Notify\_start'' but the transition associated to the event is not enabled in the marking reached by the net, so the algorithm creates artificial tokens for completing the trace replay.
+This situation is possible only if the invisible transition $t10$ is not enabled.
+Now, if $t10$ is not enabled that means that there are no tokens in the place $p10$ or $p7$.
+But notice that there are 21 remaining tokens in the place $p10$, so the transition $t10$ is not enabled for those $37$ instances because the place $p7$ does not present tokens.
+Moreover, the Petri net in figure highlights 37 remaing tokens in the place $p3$, so we can deduce that in those instances financial activities were not performed.
+From this facts, we can conclude that the conformance problems presented in  37 instances are due to a missing execution of the financial activities check.
+On the other hand, the decision tree reports that the instances with conformance problems are made by consolidate clients.
+In this case, it might be reasonable that, for the orders made by the consolidate clients of the organization, could not be necessary a financial valutation of their situation.\\
+
+The analysis based on a combination of the log replay results and the pattern data described by the classification, may allow to discover new cases of the business procedure to take into consideration during a possible process extension.
+For instance, with the analysis done before, in order to extend the model presented in section \ref{example} we can include what it was discovered: for the orders done by the consolidate clients of  the organization, the financial check activity is not needed.
+In other cases, whenever  a process extension is not possible (i.e.
+a fixed procedure to be strictly respected), such kind of analysis can help in taking corrective measures in order to avoid conformance errors.
+In particular, through a combination of the log replay results and the data patterns rules it may be possible to discover (and localize) conformance errors causes.
+This fact can be fundamental to bring the right corrections.
+For instance, in our sale process, if the procedure described by the model must be respected in any case, a revision on the staff behaviour about consolidate clients can be done in order to avoid conformance errors.\\
+
+The classifier constructed with this approach can be used not only in a descriptive sense but also in predictive way.
+In fact, given a new event log, in order to perform the conformance checking to the process model, the decision tree obtained can be used to predict the conformance result of the new process instances.
+This can allow some advantage in term of the execution time since the log replay algorithm can take more time than the one needed by a decision tree to classify a new instance.
+Also, it is worth noting that, whenever the set of attributes needed for classification is known before the execution of the process instance, it can be interesting to predict the conformance result of the instance and avoid the errors that could happen during the execution.
+
+\subsection{Classification for Perfomance Analysis}\label{ClassPerf}
+
+The approach presented in section \ref{ClassConformance} can be extended for performance analysis.
+Using classification technique for discovering how data attribute can influence process performance provides useful information in analysing and optimizing the process services.
+For example, discovering data patterns in correspondence of which some activities need more time for completion than others, helps in making decision about resources distribution to various process activities or in scheduling activities.\\
+
+In addition to the execution or completion time, classification can be used to discover information about more complex performance metrics such as the syncronization time.
+Processes with parallel activities and syncronizations can often present activities that represent a bottleneck, this leads to increase the waiting time of some activities and consequently the completion time of the entire process.
+In this context, with the approach presented in this paper, the goal is to discover rules in the process data that influence the syncronization time.
+The model built by classification techinque can be used also in a predicting way, and this is certainly very important in order to optimize performance execution of the process.\\
+
+\begin{table}[!h]
+\scriptsize
+\centering
+\begin{tabular}{|p{1cm}|p{1cm}|p{1,5cm}|p{1cm}|p{1,3cm}|p{1,4cm}|p{1,1cm}|p{1,2cm}|p{1,5cm}|}
+\hline OrdIde & CltIde & CltType & SalMan & FinOff & WrhsMan & Supplier & OrdResut & Synch\\
+\hline
+1 & 20 & consolidate & Marco & Alessandro & Alex & Gianni & positive & high\_wareh.\\
+\hline
+2 & 15 & new & Anna & Mario & Alessio & Mario & positive & normal\\
+\hline
+3 & 10 &consolidate & Maria & Roberto & Alessio & Mario & negative & normal\\
+\hline
+2 & 15 & new & Anna & Mario & Roberto & Gianni & positive & high\_wareh.\\
+\hline
+... & ... & ... & ... & ... & ... & ... & .... & .... \\
+\hline
+... & ... & ... & ... & ... & ... & ... & .... & ...  \\
+\hline
+\end{tabular}
+\caption{Data set}
+\label{tab:SaleDataPerf}
+\end{table}
+\normalsize
+
+With reference to the example in section \ref{example} and to the performance metrics explained in section \ref{Background}, in this paragraph the approach based on classification is illustrated for the performance analysis.
+From the Petri net in figure \ref{pnet} two parallel branches are identified, one presents financial activities and the other the warehouse activities.
+We are interested in analysing the syncronization between the two branches.
+Indeed, the synchronization time of interest are: $tsc(p10)$, $tsc(p7)$ that are one of the results returned by the log replay algorithm (section \ref{Background}).
+As explained for the conformance analysis, to formulate the classification problem the dataset is in part extracted from the event logs.
+Target attribute must be something that characterizes the performance in term of synchronization.
+The idea chosen in this illustration considers simply for each instance the difference between syncronization time of $p10$ and $p7$: $tsc(p7) - tsc(p10)$.
+It should be noted that classification deals with discrete or categorical target attribute.
+Since the performance metrics (syncronization time in this case) are expressed as continous values, a discretization phase is required  during the formulation of the classification problem.
+Let us choose the target attribute as a categorical attribute.
+Given a treshold $S$, if $tsc(p7)- tsc(p10)$ exceeds the value $S$, then the target attribute indicates the branch with the high syncronization time, otherwise the target indicates that the instance presents a normale syncronization time.
+Therefore, the target attribute can assume three possible values: \emph{high\_financial, high\_warehouse} and \emph{normal}.
+Based on these information (process data extracted form event logs and target attribute desumed from log replay results) the data set for the classification problem is formulated and presented in table \ref{tab:SaleDataPerf}.
+As done for conformance analysis, the data set is used as an input for the classification algorithm that returns the decision tree presented in figure \ref{salesPerfDecTree}.\\
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=140pt,height=80pt]
+{./items/Sales_perf_tree.pdf}
+\caption{decision tree}
+\label{salesPerfDecTree}
+\end{figure}
+
+The decision tree shows that the syncronization time is too high, so not acceptable, for instances in which the external warehousing is done by the supplier ``Gianni''.
+In other general terms, the rule detected by the decision tree correlates the process performance with something regarding the warehousing.
+So the process efficiency dependes on the provision policy of the organization.
+
+\section{Implementation with ProM Framework}
+In order to do experiments with some business process prototype, the approach presented in this paper has been implemented as a set of plugins that integrates the \emph{ProM6}, a process mining framework \cite{}, and \emph{Weka}, a datamining framework providing tools about classification and other techniques.
+We can divide the plugins in three classes:
+\begin{itemize}
+\item Plugins for managing the event log: with these plugins the goal is parsing the event log which traces the business process, extracting significant attributes and preparing the data set.
+There are three plugin in this class:\\
+\begin{itemize}
+\item \emph{Generate Instances With Conformance:} this plugin takes as input an event log and returns as output a complete data set with the target attribute as shown in figure \ref{tab:SaleData}.
+The data set returned respects the format needed by Weka classification tool.
+\item \emph{Generate Instances With Performance}: this plugin is analogous to the previous one, but the data set generated in this case has as target attribute something characterizing the process performance as discussed in section \ref{ClassPerf}.
+\item \emph{Generate Instances to Classify}: starting from an event log, with this plugin the goal is to generate a set of instances for which the conformance (or performance) result is unknown.
+The instances generated can be classified using an existing classification model.\\
+\end{itemize}
+\item Plugins for managing the classification: with these plugins the goal is to generate a classification model based on a data set and to use it for classifying unknowm records.
+In fact, in this class we find two plugins:\\
+\begin{itemize}
+\item \emph{Generate Classifier}: given a data set in which target attribute is known, this plugin uses tools provided by \emph{Weka} library to generate a classifier.
+In particular, the model generated is a decision tree built using $J48$ algorithm, the implementation of $C4.5$ developed by \emph{Weka}.
+\item \emph{Classify Instances}: given a set of not classified instances and a classifier model, this plugin simply classify the instances according to the classifier. \\
+\end{itemize}
+\item Plugins managing resources: in this class we find a set of plugins useful for the serialization (and deserialization) of the resources used by the previous plugins, such as data set and classifiers.
+There are also some plugins allowing the visualization of these resources into \emph{ProM6} framework.
+\end{itemize}
 

### 2013-09-17 paper/paper2.tex
last versions

\documentclass{llncs}
\usepackage[latin1,utf8x]{inputenc}
\usepackage{color}
\usepackage[pdftex]{graphicx}
\usepackage{amsmath}
\usepackage{verbatim}
\usepackage{marginnote} 
%\usepackage[outer=5cm, inner=2cm, heightrounded, marginparwidth=4cm, marginparsep=0.5cm]{geometry}

\newcommand\margin[1]{\marginnote{\scriptsize{\bfseries #1}}}

\pagestyle{plain}

\title{Classification Techniques for Conformance and Performance checking in Process Analysis}
\author{Hind Chfouka, Andrea Corradini, Roberto Guanciale}
\institute{Department of Computer Science, Università di Pisa}
\date{today}
\begin{document}
\maketitle
\begin{abstract}
Standard process analysis techniques like conformance checking or performance evaluation are enabled by the event logs that trace the process executions and by the presence of a model that formally represents the process. Such analysis techniques use only part of the huge amount of data recorded in the event logs. In this paper the goal is to transform this data into useful information for the conformance checking and the performance analysis. Through classification technique, we present an approach to investigate how the process data influences its behavior in term of conformance and performance. 
\end{abstract}

\section{Introduction}

Today, many Information Systems that integrate the concept of business process record all events occurring during the process execution in an event log. An event log represents a trace of the process behavior that can be observed and analyzed in order to tune it with the business objectives pursued. Process Mining provides a set of techniques for process discovery and analysis. Through the knowledge extracted from event logs, process discovery allows the construction of a process model. Process analysis instead assumes the existence of a model and it consists of checking the conformance and performance of the process executions with respect to it.

The motivation of this paper is based on the observation that existing event logs are rich of data, and this data is used only in part by the existing process analysis techniques. Since data is considered a valid source of knowledge, developing a technique based on data analysis can provide advantages for  process analysis.
Therefor, our goal consists of finding a way for transforming the data recorded during the process activities in knowledge useful for the analysis. To this aim, one possible approach is to try to understand how data may influence the process behavior. In particular, it is interesting to understand how may the data influence the conformance and performance of the process executions. Exploring this influence can provide qualitative information about the causes of possible anomalies in the process behavior, and this will facilitate the task of taking corrective measures.

In order to analyze data arising from event logs, in this paper we present an approach based on \emph{data mining} \cite{5}. In particular, we exploit \emph{classification} techniques to find patterns on data in presence of which conformance errors occur. Next, the same approach is extended to consider the process behavior in terms of performance to discover how data influences the process efficiency. Classification is a well-known \emph{machine learning} technique, and it consists of identifying to which category, among a given set, a new observation belongs. This is done on the basis of a training set of data containing observations whose category membership is known.

Classification techniques in the Process Mining field have already been explored \cite{1}\cite{14}. In \cite{1} the idea is exploring how data influences the case routing of process flow execution. In this contribution the goal is to strengthen the use of data analysis techniques for the benefit of process analysis. This is done by studying how data may influence, in general, the overall process behavior.

After the presentation of a simple business process in Section \ref{example}, Section \ref{Background} introduces some background concepts about process analysis. First the algorithm log replay is briefly described, then conformance analysis is presented and illustrated through the business process of Section \ref{example}. In Section \ref{ClassifApproach} our approach to process analysis based on classification is presented. First a few basic concepts about the classification technique are described, next the conformance analysis based on classification is presented in detail through an illustration with the process of Section \ref{example}, moreover a possible extension of the approach for performance analysis is presented. Finally, Section \ref{implementation} describes the extension of \emph{ProM6}\cite{6} framework by some plug-ins realizing the approach presented in Section \ref{ClassifApproach}, as well the methodology to follow in the process analysis based on classification.

\section{An example of business process model for a sale process}\label{example}
A business process can be seen as a collection of activities occurring within an organization that lead to a specific goal. There are several formalisms for representing business processes. In this paper we represent them using Petri nets \cite{8}\cite{10} since this make possible exploiting some existing process analysis techniques that are based on Petri nets. However, in the business management context, Petri nets can be not expressive enough and usually process models are presented as flowcharts, in particular using the standard for the business process modeling: BPMN (Business Process Model and Notation) \cite{9}. Transforming a BPMN model into a Petri net is possible thanks to a mapping technique presented in \cite{12}\cite{2}.
\begin{figure}[h]
\includegraphics[width=360pt]
{./items/Sales_PN.pdf}
\caption{Petri net for a a sale process}
\label{pnet}
\end{figure}

Figure \ref{pnet} presents the Petri net model for a sale process. This is the business process representing in an abstract way a fragment of the procedure followed in a commercial organization for managing orders received from clients. In general this procedure includes different activities and each of them involves a specific department of the organization. We assume that the sale process begins with receiving an order notification from a client, so the first activity is simply called \emph{Order}. After the order is received, the sale process continues with some activities that can be done in parallel, since they do not present any dependencies. One of these is the \emph{FinancialCheck} activity during which the order is analyzed from a financial point of view (i.e verifying the financial situation of the client). In parallel to this, an activity called \emph{WarehouseCheck} is performed. In this phase the availability of merchandise requested by the order is checked and, in case of a shortage, a supplying procedure is started. This is done through the activity \emph{ExternalWarehouse}. After the completion of the  parallel activities, a synchronization is needed before continuing the sale procedure with the final activity called \emph{Notify}, by which a result regarding the acceptance of the order is communicated to the client.

Note that the Petri net of Figure \ref{pnet} is obtained by transforming a BPMN model using the algorithm of \cite{2}, and mapping each activity to a pair of start/end transitions. The resulting net has been simplified by removing the unnecessary \emph{invisible transitions} used to encode e.g. the \emph{Join} or \emph{Fork} gateway \cite{2}.

\section{Process Analysis}\label{Background}

Process analysis is performed using the Petri net representing the process and an event log recording the related executions, also called \emph{process instances}. The basic building blocks of event logs are events. An event {\itshape e} can be seen as a pair {\itshape e = (a,t) } representing an action {\itshape a } recorded and the corresponding timestamp {\itshape t}. Action and timestamp are denoted respectively by $\alpha(e)$ and $\phi(e)$. Events that belong to the same process are grouped into {\itshape traces}. Formally, a trace $T$ is a finite sequence of events $T[1],..., T[n]$, such that $\phi(T[i]) \leq \phi(T[i+1])$ for all  $i \in [1,n)$. A log $L$ is a set of traces, recording the activities performed by a system during a finite number of process executions. In this paper the following assumptions are made:
\begin{itemize}
\item All traces are instances of the same process.
\item For each action there exist two corresponding transitions in the net that will be denoted, for simplicity, by the same name of the action.
\end{itemize}

\subsection{Log replay algorithm}\label{logreplayAlg}

The key algorithm exploited to analyze a Petri net model with respect to the log is the {\itshape log replay} algorithm \cite{13}\cite{3}\cite{4}. Given a Petri net model and an event log as input to the algorithm, the output results can be used to check the conformance of traces and to evaluate some performance metrics. For each trace in the log, the algorithm starts by placing one token in the start place of the net. For each event in the trace the corresponding transition is fired assuming a {\itshape non-blocking} behavior of the algorithm, then the marking of the net is updated. Non-blocking replay means that whenever a firing of a disabled transition is needed, the algorithm enables the transition either creating artificial tokens in the pre-set or, if possible, firing some invisible transitions; the non-determinism in this procedure is resolved with a suitable cost function. The output of a log replay execution on a trace can be represented as an ordered list $R$ of pairs $(tr, i)$, representing that the transition $tr$ has been fired to mimic the event $T[i]$.

The log replay result can be used to evaluate both the conformance and performance of a business process.

\subsection{Conformance Analysis}\label{ConformanceAnalysis}
The goal of the conformance analysis is to check if a trace complies with the Petri net modeling the business process. Conformance problems can be discovered by analyzing tokens artificially created during the replay {\itshape (the missing tokens)} and tokens not consumed {\itshape (the remaining tokens)}. Figures \ref{ConfLog} and \ref{NonConfLog} present two different traces, $T$ and $T'$, of the business process presented in Section \ref{example}. The log replay of the trace $T$, which is compliant with the Petri net, terminates with a marking containing one token in the end place \{${p13 \rightarrow 1}$\} and without reporting any missing tokens.

\begin{comment}
\begin{equation}
\begin{split}
R&=\{(Order\_start,1), (Order\_complete,2),(WarehouseCheck\_start,3), \\
& (FinancialCheck\_start,4),(FinancialCheck\_complete,5),\\
& (WarehouseCheck\_complete,6),(t2, 7), (Notify\_start,7), \\
&(Notify\_complete,8)\}
\end{split}
\end{equation}
\end{comment}

\begin{figure}[h]
\centering
\includegraphics[width=300pt,height=60pt]
{./items/logConforme.pdf}
\caption{A conforme trace: $T$}
\label{ConfLog}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=300pt,height=60pt]
{./items/logNonConforme.pdf}
\caption{Non-conforme trace: $T'$.}
\label{NonConfLog}
\end{figure}

The log replay of the trace $T'$ terminates with remaining tokens \{${p5 \rightarrow 1,p13 \rightarrow 1}$\} and one missing token \{${p8 \rightarrow 1}$\}. The missing token is created artificially and this fact witness a wrong execution of the event $Notify\_start$. In fact, in $T'$ this event is executed before the termination of the activity \emph{FinancialCheck}: this is interpreted as a non-conformance to the process model.

\begin{comment}
The sequence returned by log replay is:
\begin{equation}
\begin{split}
R'&=\{(Order\_start,1), (Order\_complete,2), (WarehouseCheck\_start,3), \\
& (FinancialCheck\_start,4), (WarehouseCheck\_complete,5),(t1,6),(t2,6),\\
& (Notify\_start,6), (Notify\_complete,7)\}
\end{split}
\end{equation}
\end{comment}

\begin{comment}

\subsection{Performance analysis}\label{PerformanceAnalysis}
A performance analysis of a process can be performed by exploiting the log replay results. Since the logs contain timestamps, the log replay can be used to compute performance measures of the process. This is done based on the elapsed time between production and consumption of tokens. Performance analysis can be applied only to traces that do not generate missing tokens during the replay since such tokens cannot have time information \cite{2} . During the log replay the following metrics can be computed for each trace and each place:
\begin{itemize}
\item sejour time $(tsj)$ : the time interval between arrival and departure of tokens;
\item synchronization time $(tsc)$: the time interval between arrival of a token in the place and enabling of a transition in the post-set of the place;
\item waiting time $(tw)$:  the time interval between enabling of a transition in the post-set of the place and token departure (thus $tsj=tsc+tw $).
\end{itemize}

The duration of an activity is an important performance measure as well, and it is given by the waiting time of the only place between the start and the complete transition of the activity.
\end{comment}

\begin{comment}
To clarify the metrics evaluated by this technique, as done with conformance analysis, we exploit the Petri net  presented in Section \ref{example} and the trace T presented in Figure \ref{ConfLog}. The replay starts with firing the transition $Order\_start$ at time $0s$, so a token arrives at $p2$ at time $0s$. After $45 min$ the token in $p2$ is consumed and the transition $Order\_complete$ is fired, so $tw(p2)=45min$, and a token arrives at $p3$ and $p4$. At the time $1h:5min:25s$ the transition $WarhouseCheck\_start$ is fired, thus $tw(p4)=20min+25sec$ and a token is produced at $p6$. At the time $1h:10min:20s$ the transition $FinancialCheck\_start$ is fired and a token is produced at $p5$, so $tw(p3)=25min+20s$. After $2h:30min:20s$ from the replay start the transition $Financial\_complete$ is fired, hence we have $tw(p5)=1h+20min$ and a token is produced at $p8$. At time $3h:10min:20s$ the firing of the transition $WarehouseCheck\_complete$ is executed, so $tw(p6)=2h+4min+55s$ and a token is produced at $p7$. After $15 min$ the invisible transition $t1$ is fired and a token is produced at $p10$. At this point the transition $t2$ is enabled, so its firing is now possible and a token is produced at $p11$. Notice that $tsc(p8)=40min, tsc(p10)=0s$. At the time $3h:25min:20s$ the transition $Notify\_start$ is fired and a token is produced in $p12$, after $4min+4sec$ the transition $Notify\_complete$ is fired and $tw(p12) = 4min+40s$.
\end{comment}

It is worth noting that the synchronization time is greater than zero only for places which contain in their post-set transitions depending from other places. For the Petri net in Figure \ref{pnet}, the only places which can have a non-null synchronization time are $p8$ and $p10$. For example, the synchronization time is greater than zero for $p8$ in the trace $T$ in Figure \ref{ConfLog}. That means that, for this particular instance of the process, the branch with the activity related to the financial checking is faster than the branch with the warehouse activities.

\section{An approach based on classification for Process Analysis}\label{ClassifApproach}
Employing data mining techniques for process analysis is encouraged by the presence of a huge amount of data recorded in event logs under \emph{attribute} format. The implicit information contained in those data could be significant for the process behavior analysis. In fact, this potential information could contain an explanation for the deviations discovered during conformance analysis, and for the performance level provided by the process. For this reason, we are interested in discovering how the process data may influence the conformance and the performance of the process executions. This is done with an approach based on \emph{classification}, a classical data mining technique able to detect patterns on data in correspondence to which the process assumes a specific behavior.

\subsection{Classification: basic concepts}\label{classifBasicConcept}

In a classification problem \cite{5}, data is represented by a collection of records (called instances), and each of them is characterized by a tuple $(\mathbf{x},y)$, where $\mathbf{x}$ is a set of attributes, while $y$ is a special attribute called \emph{target attribute}. The target denotes the class to which a record belongs. With this technique, the goal is to learn a classification model that maps each attribute set $\mathbf{x}$ to one of the predefined class labels $y$.

A classification model can be useful for a descriptive purpose: it provides an explanatory tool to distinguish between objects of different classes, and for a predictive purpose: the model can be used to predict the class label of new records. In this way the model acts as a black box that automatically assigns a class label to an attribute set of a new record.

There is a general approach to solve a classification problem from a dataset, that is independent from the specific classification model. Each classification technique employs a learning algorithm to identify a model that best fit the relationship between the attribute set and class label of the input data. The model generated should both fit the input data well, and correctly predict the class labels of new records. Finding the right trade\_off between this two purposes is the most delicate part of a classification technique, and in general of any machine learning task.

\begin{figure}[h]
\centering
\includegraphics[width=200pt]
{./items/classifApproach.pdf}
\caption{General approach for solving a classification problem.}
\label{figclassApproach}
\end{figure}

Figure \ref{figclassApproach} summarizes the general approach for solving a classification problem \cite{5}. A \emph{Training set} consisting of records whose class labels are known must be provided to a learning algorithm, and it is used to build a classification model. This model is then applied to a \emph{test set}, which consists of records with unknown class labels. This application allows the estimation of the \emph{model accuracy} that can be done in different ways. The simple one consists in computing the \emph{performance} as the number of the records correctly predicted over the total number of records in the test set.

Many classification models could be used for a classification problem. For our purpose we choose the \emph{decision tree} model since the rules detected by the classification are shown explicitly. This fact is fundamental for giving an explanation to the process behavior based on the process data values. Furthermore, this choice is encouraged by the simplicity of decision trees. In a decision tree, a class label (possible value of the attribute target) is assigned to each leaf node. The non-terminal nodes, which include the root and the internal nodes, contain attribute test conditions to filter records that have different characteristics.
\begin{comment}
In principle, there are exponentially many decision trees that can be constructed from a given set of attributes. While some of the trees are more accurate than others, finding the optimal tree is computationally infeasible because of the exponential size of the search space. However, efficient algoritms have been developed to induce a reasonably accurate decision tree in a reasonable amount of time. These algorithms usually employ a greedy strategy that grows a decision tree by making a series of decisions about which attribute to use for partioning the data. For instance Hunt's algorithm is one of them and it represents the basis of many existing decision tree induction algorithms, including $C4.5$ algorithm used for our purposes.
\end{comment}
\subsection{Classification for Conformance Checking}\label{ClassConformance}
In order to identify the possible causes of non-conformance, a classification problem can be formulated. The classification dataset is extracted from the process data: for each process instance, one will extract some data that will be included as a record in the dataset. For the process presented in Section \ref{example}, a record is characterized by the following attribute set: an instance identifier, a client identifier, the client typology (new or consolidated client), the sales manager responsible for the order, the financial officer who conducts the financial evaluation activities, the warehouseman responsible for the warehouse checking, the supplying responsible name in case of a provision, and finally the result communicated to the client for the order issued. The conformance result for the process instance provides an additional and important data which takes the role of the target attribute. All these attributes are discrete and contribute to build the dataset presented in table \ref{tab:SaleData} of the classification problem. Note that, since the goal of this paper is just to illustrate an approach for exploiting process data in the process analysis, the data presented here is just a synthetic data that was generated with some noise to emulate a real situation.

\begin{table}[!h]
\scriptsize{
\centering
\begin{tabular}{|p{1cm}|p{1cm}|p{1,5cm}|p{1,2cm}|p{1,4cm}|p{1,2cm}|p{1,5cm}|p{1,4cm}|p{0,8cm}|}
\hline OrdIde & CltIde & CltType & SalMan & FinOff & WrhsMan & SupplyResp & OrdResut & Conf\\
\hline
1 & 20 & consolidate & Marco & Mary & Alex & Gianni & positive & no\\
\hline
2 & 15 & new & Anna & Johann & Roberto & Mario & positive & yes\\
\hline
3 & 10 &consolidate & Maria & Mary & Alessio & Gianni & negative & no\\
\hline
10 & 18 & consolidate & Johann & Mary & Roberto & Gianni & positive & yes \\
\hline
... & ... & ... & ... & ... & ... & ... & .... & ...  \\
\hline
\end{tabular}
}
\caption{Dataset for the conformance analysis.}
\label{tab:SaleData}
\end{table}
\normalsize
Through existing data mining tools for classification, it is possible to build a decision tree as a classification model for the sale process example. Starting from a given event log, the attributes characterizing the process are extracted and all the resources needed for the classification algorithm employed are produced. The resulting decision tree is presented in Figure \ref{salesDecTree}.

\begin{figure}[h]
\centering
\includegraphics[width=210pt,height=100pt]
{./items/Sales_tree.pdf}
\caption{Decision tree for the conformance analysis.}
\label{salesDecTree}
\end{figure}
The resulting decision tree describes a data pattern in correspondence to which a process instance could present conformance errors: order managed by the sales manager \emph{Mary} and received from consolidate clients of the organization may not respect the standard sales procedure. To get more significant information for our analysis, it is useful to relate what is noticed by the decision tree with the log replay results. Figure \ref{replayResult} presents a Petri net that summarizes the conformance analysis conducted on instances recorded in the event log taken in exam. Arcs are labeled with the number of activations done. Places, whenever they present some remaining or missing tokens, are labeled with the number of these tokens.
\begin{figure}[h]
\includegraphics[width=360pt]
{./items/Sales_PN_resultMod.pdf}
\caption{Petri net: conformance results.}
\label{replayResult}
\end{figure}

The Petri net in Figure \ref{replayResult} shows that there are 37 out of 250 process executions with conformance errors. The presence of 37 missing tokens in $p11$ signals a forced execution of  transition \emph{Notify\_start}. In order to provide an interpretation for this results, we have to take into account the non-blocking behavior of the log replay algorithm. In this case, the 37 missing tokens in $p11$ certifies that during the replay the algorithm needs to mimic the event \emph{Notify\_start} but the transition associated to the event is not enabled in the marking reached by the net, so the algorithm creates artificial tokens for completing the trace replay. This situation is possible only if the invisible transition $t2$ is not enabled. Now, if $t2$ is not enabled that means that there are no tokens in the places $p8$ or $p10$. But notice that there are 21 remaining tokens in the place $p10$, so the transition $t2$ is not enabled for those $37$ instances because the place $p8$ does not have any token. Moreover, the Petri net in Figure \ref{replayResult} shows 37 remaining tokens in place $p3$, so we can deduce that in those instances financial activities were not performed. From these facts, we can conclude that the conformance problems presented in  37 instances are due to a missing execution of the financial activity checking. On the other hand, the decision tree reports that the instances with conformance problems are made by consolidate clients.

From the analysis done for the sales process based on both classification and log replay results, a new scenario of the sale procedure has emerged: orders done by the consolidate clients of the organization are not checked from a financial point of view. This fact can be accepted as a possible and reasonable behavior in a sale procedure, consequently an extension of the business process model is needed in order to include the new scenario (\emph{model extension activity}). Alternatively, the new scenario can be considered as an anomaly in the sales procedure. In this case, since the analysis done provides accurate information about the conformance error, corrective measures can be taken in order to avoid errors in the next orders.

The classifier constructed with this approach can be used also in a predictive way. Given a new trace recording a process execution, in order to perform the conformance checking to the process model, the decision tree can be used to predict the conformance result. This bring an advantage in terms of the time required by the analysis since the log replay algorithm takes more time than the one needed by a decision tree to classify a new instance. Moreover, it is worth noting that whenever the set of attributes needed for the classification is known before the completion of a process execution, it can be interesting to predict the conformance result of the execution and avoid the errors that could possibly happen.

\subsection{Extension of the approach for Performance Analysis}\label{ClassPerf}

Performance analysis of a process can be performed exploiting log replay. Since a log contains timestamps, during the replay of a trace it is possible to compute (for each place of the Petri net model) some performance measures such as the \emph{synchronization time} (i.e. time interval between arrival of a token in the place and enabling of a transition in the post-set of the place).

Performing an analysis based on the measures computed by log replay gives just a \emph{quantitative} information about the process performance. In order to understand the \emph{cause} of performance anomalies that can affect the process behavior, we have to explore the process data. Discovering how data attribute can influence process performance provides useful information in analyzing and optimizing the process services. For example, discovering data patterns in correspondence to which some activities need more time for completion than others, helps in making decisions about resources distribution to the process activities or in scheduling activities. In addition to the completion time, one could analyze a process under more complex performance metrics such as the \emph{synchronization time}. A process with parallel branches and synchronizations can present bottleneck activities that lead to increase the execution time of other activities and consequently the completion time of the entire process. To find out the possible data influences on synchronization time, the approach presented in Section \ref{ClassConformance} for conformance analysis through classification can be easily extended to performance analysis. For example, a classification problem can be formulated for each synchronization point of the process. In this way, the model obtained in correspondence to a synchronization point classifies the process instances based on their attribute value and regarding the synchronization time of the point taken into exam.

\begin{comment}
\begin{table}[!h]
\scriptsize
\centering
\begin{tabular}{|p{1cm}|p{1cm}|p{1,5cm}|p{1cm}|p{1,3cm}|p{1,4cm}|p{1,1cm}|p{1,2cm}|p{1,5cm}|}
\hline OrdIde & CltIde & CltType & SalMan & FinOff & WrhsMan & Supplier & OrdResut & Synch\\
\hline
1 & 20 & consolidate & Marco & Alessandro & Alex & Gianni & positive & high\_wareh.\\
\hline
2 & 15 & new & Anna & Mario & Alessio & Mario & positive & normal\\
\hline
3 & 10 &consolidate & Maria & Roberto & Alessio & Mario & negative & normal\\
\hline
2 & 15 & new & Anna & Mario & Roberto & Gianni & positive & high\_wareh.\\
\hline
... & ... & ... & ... & ... & ... & ... & .... & .... \\
\hline
\end{tabular}
\caption{Dataset for the performance analysis.}
\label{tab:SaleDataPerf}
\end{table}
\normalsize

To discover how the process data may influence its performance in term of synchronization time, a classification problem is formulated for each synchronization point of the process. In this way, the model obtained in correspondence to a synchronization point classifies the process instances based on their attribute value and regarding the synchronization time of the point taken into exam. For the sales process presented in Section \ref{example}, a single point of synchronization is identified. In fact, the Petri net in Figure \ref{pnet} has two parallel branches, one for the financial activity and the other for the warehouse activities. We are interested in analyzing the synchronization between these branches. Indeed, the synchronization time of interest is measured in correspondence to the places $p(8)$ and $p(10)$: $tsc(p8)$, $tsc(p10)$ which are computed by the log replay algorithm as explained in Section \ref{Background}.

To formulate the classification problem the dataset is extracted from the event log, whereas the target attribute is one of the log replay performance results. The target must be something that characterizes the performance in term of synchronization. An idea can be to simply consider the difference between synchronization time of $p8$ and $p10$: $tsc(p8) - tsc(p10)$. It should be noted that classification deals with discrete or categorical target attribute. Since the performance metrics, the synchronization time in this case, are expressed as continous values, a discretization phase is required  during the formulation of the classification problem. Let us choose the attribute target as a categorical attribute: given a treshold $S$, if $|tsc(p8)- tsc(p10)|$ exceeds the value $S$, then the target attribute indicates the branch with the high synchronization time, otherwise the target indicates that the instance presents a normal synchronization time. Therefore, the target attribute can assume three possible values: \emph{high\_financial, high\_warehouse} and \emph{normal}. Based on the process data extracted from the event log and target attribute computed by a discretization of the log replay results, the dataset for the classification problem is built and presented in table \ref{tab:SaleDataPerf}. As done for conformance analysis, the dataset is used as input of the classification algorithm that returns the decision tree presented in Figure \ref{salesPerfDecTree}.

\begin{figure}[h]
\centering
\includegraphics[width=120pt,height=60pt]
{./items/Sales_perf_tree.pdf}
\caption{Decision tree for the performance analysis.}
\label{salesPerfDecTree}
\end{figure}

The decision tree shows that the synchronization time is too high, so not acceptable, for those instances in which the external warehousing is done by the supplier \emph{Gianni}. In other  terms, the rule detected by the decision tree correlates the process performance, measured as the difference of two synchronization times with the external warehousing activities. In this case, the analysis through classification identified such activities as a bottleneck in the \emph{specific case} of the supplier Gianni. This result can be taken as a start point for more investigations that can be done in a \emph{targeted manner} thanks to the pattern identified by the classification analysis. 
\end{comment}

\section{Implementation with ProM}\label{implementation}
In order to experiment with some business process prototypes, the approach presented in this paper has been implemented as a set of plug-ins that integrates \emph{ProM6} \cite{6}, an open-source framework implementing Process Mining tools, and \emph{Weka} \cite{7}, a data mining framework providing machine learning tools.
We can classify the plug-ins developed in three categories. The first one includes plug-ins for building the dataset needed for the classification:
\begin{itemize}
\item \texttt{Generate Instances With Conformance}: this plug-in takes as input an event log and its conformance results computed by log replay. It returns as output a training set as shown in Table \ref{tab:SaleData} in the format needed by Weka classification tool. Note that the plug-in extracts \emph{all} the attributes data recorded in the event log. In case, the dataset resulting can be subject of some \emph{feature selection} techniques before using it as a training set for the classification task.

\begin{comment}
\item \texttt{Generate Instances With Performance}: the plug-in takes as input an event log, an indication of the synchronization point to examine and the performance result computed by the log replay. The output is a training set as presented in table \ref{tab:SaleDataPerf}.
\end{comment}

\item \texttt{Generate Instances to Classify}: given an event log, the plug-in generates a dataset of instances that can be classified using a classification model.
\end{itemize}
The second category includes plug-ins for the classification model generation and utilization for the prediction purpose:
\begin{itemize}
\item \texttt{Generate Classifier}: given a training set, the plug-in generates a decision tree model using $J48$ algorithm, an implementation of the algorithm $C4.5$ developed by Weka.
\item \texttt{Classify Instances}: given a set of non-classified instances and a classifier model, this plug-in simply classify the instances according to the classifier.
\end{itemize}
Finally, the last category includes a set of plug-ins useful for the serialization, deserialization and visualization of the resources used by the previous plug-ins, such as dataset and classifiers.

\subsection{Methodology}
This last paragraph is dedicated to a description of the methodology to be followed to analyze a business process under the approach presented in this paper.

It should be clear that the analysis presented on the previous Section assumes the presence of a Petri net model of the business process to be analyzed. The model can be expressed in BPMN language as well, but in this case a transformation into a Petri net can be done through the techniques described in \cite{2}. The classification analysis  needs also an event log tracing the process behavior over multiple instances. Given a process model $M$ and an event log $L$, the methodology to be followed for the conformance analysis with the approach presented in Section \ref{ClassConformance} can be summarized by these steps:
\begin{enumerate}

\item \label{step1} \textbf{Log replay analysis}: this phase consists in an execution of the log replay with  the model $M$ and the log $L$ as inputs. The results consists in two structures characterizing both the conformance and the performance of the traces in $L$. Let us call them $ConfResult$, and $PerfResult$.

\item \label{step2} \textbf{Training set construction}: this phase aims to build the training set for the classification technique. This is done through an execution of the plug-in \texttt{Generate Instances With Conformance} with the event log $L$ and the conformance result $ConfResult$ as inputs. Let us call the training set obtaind $TS$.

\item \textbf{Classifier building}: in this step the classification model (decision tree) for the conformance analysis is built using the plug-in \texttt{Generate Classifier} with the training set $TS$ as input. Let us call the decision tree obtained $DT$.

\item \textbf{Using the classifier}: in this step the goal is to exploit $DT$ and the log replay result $ConfResult$ to establish the corrective measures in case of anomalies. Note that both $DT$ and $ConfResult$ can be visualized as in Figures \ref{salesDecTree} and \ref{replayResult} into ProM6 thanks to the plug-ins for resources managing that integrate the framework. $DT$ can also be used in a predictive way to establish the conformance of new process instances without needing a log replay execution. Given an event log containing new process instances, a dataset can be built through the plug-in \texttt{Generate Instances to Classify} and then classified by the plug-in \texttt{Classify Instances} according to the model $DT$.
\end{enumerate}

To perform a performance analysis based on the extension presented in Section \ref{ClassPerf}, the same methodology can be followed modifying the step \ref{step2}. The construction of the training set is done through a plug-in (analogous to \texttt{Generate Instances With Conformance}) that takes as input $PerfResult$ obtained by the step \ref{step1} and an indication about the synchronization point chosen for the analysis.

\addcontentsline{toc}{chapter}{Bibliography} 
\bibliography{biblio}
\bibliographystyle{abbrv}

\end{document}

### 2013-09-17 paper/paper2.tex
minor changes

@@
 Since data is considered a valid source of knowledge, developing a technique based on data analysis can provide advantages for  process analysis.
-Therefor, our goal consists of finding a way for transforming the data recorded during the process activities in knowledge useful for the analysis.
+Therefore, our goal consists of finding a way for transforming the data recorded during the process activities in knowledge useful for the analysis.
 To this aim, one possible approach is to try to understand how data may influence the process behavior.
 In particular, it is interesting to understand how may the data influence the conformance and performance of the process executions.
-Exploring this influence can provide qualitative information about the causes of possible anomalies in the process behavior, and this will facilitate the task of taking corrective measures.
+Exploring this influence can provide qualitative information about the causes of possible anomalies in the process behavior, facilitating the task of taking corrective measures.
 
@@
 Next, the same approach is extended to consider the process behavior in terms of performance to discover how data influences the process efficiency.
-Classification is a well-known \emph{machine learning} technique, and it consists of identifying to which category, among a given set, a new observation belongs.
+Classification is a well-known \emph{machine learning} technique, and it consists of identifying to which category among a given set a new observation belongs.
 This is done on the basis of a training set of data containing observations whose category membership is known.
 
-Classification techniques in the Process Mining field have already been explored \cite{1}\cite{14}.
+Classification techniques in the Process Mining field have already been explored.
 In \cite{1} the idea is exploring how data influences the case routing of process flow execution.
@@
 
-After the presentation of a simple business process in Section \ref{example}, Section \ref{Background} introduces some background concepts about process analysis.
-First the algorithm log replay is briefly described, then conformance analysis is presented and illustrated through the business process of Section \ref{example}.
-In Section \ref{ClassifApproach} our approach to process analysis based on classification is presented.
-First a few basic concepts about the classification technique are described, next the conformance analysis based on classification is presented in detail through an illustration with the process of Section \ref{example}, moreover a possible extension of the approach for performance analysis is presented.
-Finally, Section \ref{implementation} describes the extension of \emph{ProM6}\cite{6} framework by some plug-ins realizing the approach presented in Section \ref{ClassifApproach}, as well the methodology to follow in the process analysis based on classification.
+After the presentation of a case study in Section \ref{example},
+Section \ref{Background} introduces some background concepts about
+process analysis. First the algorithm log replay is briefly described,
+then conformance and performance analysis are presented. Both kind of
+analysis are illustrated through the business process of Section
+\ref{example}. In Section \ref{ClassifApproach} our approach to
+process analysis based on classification is presented. First a few
+basic concepts about the classification technique are described, next
+the conformance analysis based on classification is presented in
+detail through an illustration with our case study, moreover a
+possible extension of the approach for performance analysis is
+presented. Finally, Section \ref{implementation} describes the
+implementation of our approach provided as \emph{ProM6}\cite{6} plug-ins, as well the methodology to follow in the process analysis based on classification.
 
-\section{An example of business process model for a sale process}\label{example}
+\section{The case study: a sale business process}\label{example}
 A business process can be seen as a collection of activities occurring within an organization that lead to a specific goal.
@@
 
-Figure \ref{pnet} presents the Petri net model for a sale process.
-This is the business process representing in an abstract way a fragment of the procedure followed in a commercial organization for managing orders received from clients.
-In general this procedure includes different activities and each of them involves a specific department of the organization.
-We assume that the sale process begins with receiving an order notification from a client, so the first activity is simply called \emph{Order}.
-After the order is received, the sale process continues with some activities that can be done in parallel, since they do not present any dependencies.
-One of these is the \emph{FinancialCheck} activity during which the order is analyzed from a financial point of view (i.e verifying the financial situation of the client).
-In parallel to this, an activity called \emph{WarehouseCheck} is performed.
-In this phase the availability of merchandise requested by the order is checked and, in case of a shortage, a supplying procedure is started.
+Figure \ref{pnet} presents the Petri net model for a sale
+process.
+\footnote{
+The Petri net of Figure \ref{pnet} is obtained by
+transforming a BPMN model using the algorithm of \cite{2}, and mapping
+each activity to a pair of start/end transitions. The resulting net
+has been simplified by removing the unnecessary \emph{invisible
+  transitions} used to encode the \emph{Join} or \emph{Fork} gateway \cite{2}.
+}
+This business process abstractly represents a fragment of the
+procedure followed in a commercial organization for managing orders
+received from clients. In general this procedure includes several
+activities and each of them involves a specific department of the
+organization. We assume that the sale process begins with the
+notification of an order from a client, so the first activity is
+simply called \emph{Order}. After the order is received, the sale
+process continues with some activities that can be done in parallel,
+since they do not present any dependencies. The \emph{FinancialCheck}
+activity represents the financial analysis of the order  (i.e
+verifying the financial situation of the client). Concurrently, the
+\emph{WarehouseCheck} activity checks the availability of merchandise
+requested by the order and, in case of a shortage, starts a supplying procedure.
 This is done through the activity \emph{ExternalWarehouse}.
@@
 
-Note that the Petri net of Figure \ref{pnet} is obtained by transforming a BPMN model using the algorithm of \cite{2}, and mapping each activity to a pair of start/end transitions.
-The resulting net has been simplified by removing the unnecessary \emph{invisible transitions} used to encode e.g.
-the \emph{Join} or \emph{Fork} gateway \cite{2}.
-
 \section{Process Analysis}\label{Background}
 
-Process analysis is performed using the Petri net representing the process and an event log recording the related executions, also called \emph{process instances}.
-The basic building blocks of event logs are events.
-An event {\itshape e} can be seen as a pair {\itshape e = (a,t) } representing an action {\itshape a } recorded and the corresponding timestamp {\itshape t}.
-Action and timestamp are denoted respectively by $\alpha(e)$ and $\phi(e)$.
-Events that belong to the same process are grouped into {\itshape traces}.
-Formally, a trace $T$ is a finite sequence of events $T[1],..., T[n]$, such that $\phi(T[i]) \leq \phi(T[i+1])$ for all  $i \in [1,n)$.
-A log $L$ is a set of traces, recording the activities performed by a system during a finite number of process executions.
-In this paper the following assumptions are made:
-\begin{itemize}
-\item All traces are instances of the same process.
-\item For each action there exist two corresponding transitions in the net that will be denoted, for simplicity, by the same name of the action.
-\end{itemize}
+Process analysis is performed using the Petri net representing the
+process and an event log recording the related executions, also called
+\emph{process instances}. The basic building blocks of event logs are
+events. An event {\itshape e} can be seen as a pair {\itshape e =
+  (a,t) } representing an action {\itshape a } observer at the
+timestamp {\itshape t}. Action and timestamp are denoted respectively
+by $\alpha(e)$ and $\phi(e)$. Events that belong to the same process
+are grouped into {\itshape traces}. Formally, a trace $T$ is a finite
+sequence of events $T[1],..., T[n]$, such that $\phi(T[i]) \leq
+\phi(T[i+1])$ for all  $i \in [1,n)$. A log $L$ is a set of traces,
+recording the activities performed by a system during a finite number
+of process executions. In this paper we assume that
+(i) all traces are instances of the same process
+and
+(ii) for each action there exists a corresponding transition (labeled
+with the same name of the action) in the Petri net.
 
@@
 
-The log replay result can be used to evaluate both the conformance and performance of a business process.
+% The log replay result can be used to evaluate both the conformance and performance of a business process.
 
@@
 Conformance problems can be discovered by analyzing tokens artificially created during the replay {\itshape (the missing tokens)} and tokens not consumed {\itshape (the remaining tokens)}.
-Figures \ref{ConfLog} and \ref{NonConfLog} present two different traces, $T$ and $T'$, of the business process presented in Section \ref{example}.
+Figure \ref{ConfLog}  presents two different traces, $T$ and $T'$, of the business process presented in Section \ref{example}.
 The log replay of the trace $T$, which is compliant with the Petri net, terminates with a marking containing one token in the end place \{${p13 \rightarrow 1}$\} and without reporting any missing tokens.
@@
 \centering
-\includegraphics[width=300pt,height=60pt]
+\includegraphics[width=0.9\linewidth]
 {./items/logConforme.pdf}
-\caption{A conforme trace: $T$}
+\caption{Conform ($T$) and non-conform ($T'$) traces}
 \label{ConfLog}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=300pt,height=60pt]
-{./items/logNonConforme.pdf}
-\caption{Non-conforme trace: $T'$.}
-\label{NonConfLog}
 \end{figure}
@@
 
-\begin{comment}
-
 \subsection{Performance analysis}\label{PerformanceAnalysis}
-A performance analysis of a process can be performed by exploiting the log replay results.
-Since the logs contain timestamps, the log replay can be used to compute performance measures of the process.
-This is done based on the elapsed time between production and consumption of tokens.
+The performance of a process can be analyzed by exploiting the log
+replay results. Since the logs contain timestamps, the log replay can
+be used to compute performance measures of the process my measuring the elapsed time between production and consumption of tokens.
 Performance analysis can be applied only to traces that do not generate missing tokens during the replay since such tokens cannot have time information \cite{2} .
@@
 \begin{itemize}
-\item sejour time $(tsj)$ : the time interval between arrival and departure of tokens;
+\item sojourn time $(tsj)$ : the time interval between arrival and departure of tokens;
 \item synchronization time $(tsc)$: the time interval between arrival of a token in the place and enabling of a transition in the post-set of the place;
@@
 The duration of an activity is an important performance measure as well, and it is given by the waiting time of the only place between the start and the complete transition of the activity.
-\end{comment}
 
@@
 
-It is worth noting that the synchronization time is greater than zero only for places which contain in their post-set transitions depending from other places.
-For the Petri net in Figure \ref{pnet}, the only places which can have a non-null synchronization time are $p8$ and $p10$.
-For example, the synchronization time is greater than zero for $p8$ in the trace $T$ in Figure \ref{ConfLog}.
-That means that, for this particular instance of the process, the branch with the activity related to the financial checking is faster than the branch with the warehouse activities.
+It is worth noting that the synchronization time is greater than zero
+only for places which contain in their post-set transitions depending
+from other places. For the Petri net in Figure~\ref{pnet}, the only
+places which can have a non-null synchronization time are $p8$ and
+$p10$. For example, the synchronization time is greater than zero for
+$p8$ in the trace $T$ in Figure \ref{ConfLog}, representing that, for
+this particular process instance, the branch performing the financial
+checking is faster than the warehouse branch.
 
@@
 
-In a classification problem \cite{5}, data is represented by a collection of records (called instances), and each of them is characterized by a tuple $(\mathbf{x},y)$, where $\mathbf{x}$ is a set of attributes, while $y$ is a special attribute called \emph{target attribute}.
+In a classification problem \cite{5}, data is represented by a
+collection of records (called instances), and each of them is
+characterized by a tuple $(\mathbf{x},y)$, where $\mathbf{x}$ is a set
+of attributes and $y$ is a special attribute called \emph{target attribute}.
 The target denotes the class to which a record belongs.
-With this technique, the goal is to learn a classification model that maps each attribute set $\mathbf{x}$ to one of the predefined class labels $y$.
+The goal is to learn a classification model that maps each attribute set $\mathbf{x}$ to one of the predefined class labels $y$.
 
@@
 The model generated should both fit the input data well, and correctly predict the class labels of new records.
-Finding the right trade\_off between this two purposes is the most delicate part of a classification technique, and in general of any machine learning task.
+Finding the right trade-off between this two purposes is the most delicate part of a classification technique, and in general of any machine learning task.
 
@@
 \subsection{Extension of the approach for Performance Analysis}\label{ClassPerf}
-
-Performance analysis of a process can be performed exploiting log replay.
-Since a log contains timestamps, during the replay of a trace it is possible to compute (for each place of the Petri net model) some performance measures such as the \emph{synchronization time} (i.e.
-time interval between arrival of a token in the place and enabling of a transition in the post-set of the place).
-
-Performing an analysis based on the measures computed by log replay gives just a \emph{quantitative} information about the process performance.
-In order to understand the \emph{cause} of performance anomalies that can affect the process behavior, we have to explore the process data.
 Discovering how data attribute can influence process performance provides useful information in analyzing and optimizing the process services.

### 2013-09-17 paper/paper2.tex
semplificata tutta la performance

--- 2013-09-17 paper/paper2.tex
@@
 \section{Introduction}
-
 Today, many Information Systems that integrate the concept of business process record all events occurring during the process execution in an event log.
@@
 
-Note that the Petri net of Figure \ref{pnet} is obtained by transforming a BPMN model using the algorithm of \cite{2}, and mapping each activity to a pair of start/end transitions.
+Note that the Petri net of Figure \ref{pnet} is obtained by transforming a BPMN model using the algorithm described in \cite{2}, and mapping each activity to a pair of start/end transitions.
 The resulting net has been simplified by removing the unnecessary \emph{invisible transitions} used to encode e.g.
@@
 \section{Process Analysis}\label{Background}
-
-Process analysis is performed using the Petri net representing the process and an event log recording the related executions, also called \emph{process instances}.
+%Process analysis is performed using the Petri net representing the process and an event log recording the related executions, also called \emph{process instances}.
 The basic building blocks of event logs are events.
@@
 Formally, a trace $T$ is a finite sequence of events $T[1],..., T[n]$, such that $\phi(T[i]) \leq \phi(T[i+1])$ for all  $i \in [1,n)$.
+A log $L$ is a set of traces, recording the activities performed by a system during a finite number of process executions.
+Process analysis is performed using the Petri net representing the process and an event log recording the related executions, also called \emph{process instances}.
+The basic building blocks of event logs are events.
+An event $e$ can be seen as a pair $e = (a,t)$ representing an action $a$ recorded and the corresponding timestamp $t$.
+Events that belong to the same process are grouped into {\itshape traces}.
+A trace $T$ is a finite sequence of events ordered based on timestamp.
 A log $L$ is a set of traces, recording the activities performed by a system during a finite number of process executions.
@@
 \subsection{Log replay algorithm}\label{logreplayAlg}
-
 The key algorithm exploited to analyze a Petri net model with respect to the log is the {\itshape log replay} algorithm \cite{13}\cite{3}\cite{4}.
@@
 Non-blocking replay means that whenever a firing of a disabled transition is needed, the algorithm enables the transition either creating artificial tokens in the pre-set or, if possible, firing some invisible transitions; the non-determinism in this procedure is resolved with a suitable cost function.
-The output of a log replay execution on a trace can be represented as an ordered list $R$ of pairs $(tr, i)$, representing that the transition $tr$ has been fired to mimic the event $T[i]$.
+%The output of a log replay execution on a trace can be represented as an ordered list $R$ of pairs $(tr, i)$, representing that the transition $tr$ has been fired to mimic the event $T[i]$.
 
@@
 
-The log replay of the trace $T'$ terminates with remaining tokens \{${p5 \rightarrow 1,p13 \rightarrow 1}$\} and one missing token \{${p8 \rightarrow 1}$\}.
+The log replay of the trace $T'$ terminates with remaining tokens \{$p5 \rightarrow 1$, $p13 \rightarrow 1$\} and one missing token \{${p8 \rightarrow 1}$\}.
 The missing token is created artificially and this fact witness a wrong execution of the event $Notify\_start$.
@@
 At the time $3h:25min:20s$ the transition $Notify\_start$ is fired and a token is produced in $p12$, after $4min+4sec$ the transition $Notify\_complete$ is fired and $tw(p12) = 4min+40s$.
-\end{comment}
 
@@
 That means that, for this particular instance of the process, the branch with the activity related to the financial checking is faster than the branch with the warehouse activities.
+
+\end{comment}
 
@@
 This application allows the estimation of the \emph{model accuracy} that can be done in different ways.
-The simple one consists in computing the \emph{performance} as the number of the records correctly predicted over the total number of records in the test set.
+The simple one consists in computing the accuracy as the number of the records correctly predicted over the total number of records in the test set.
 
@@
 The conformance result for the process instance provides an additional and important data which takes the role of the target attribute.
-All these attributes are discrete and contribute to build the dataset presented in table \ref{tab:SaleData} of the classification problem.
+All these attributes are discrete and contribute to build the dataset presented in Table \ref{tab:SaleData} of the classification problem.
 Note that, since the goal of this paper is just to illustrate an approach for exploiting process data in the process analysis, the data presented here is just a synthetic data that was generated with some noise to emulate a real situation.
@@
 
-Performance analysis of a process can be performed exploiting log replay.
-Since a log contains timestamps, during the replay of a trace it is possible to compute (for each place of the Petri net model) some performance measures such as the \emph{synchronization time} (i.e.
-time interval between arrival of a token in the place and enabling of a transition in the post-set of the place).
+Performance analysis of a process can be carried exploiting log replay.
+Since a log contains timestamps, during the replay of a trace it is possible to compute (for each place of the Petri net model) some performance measures such as: \emph{synchronization time} (i.e.
+time interval between arrival of a token in the place and enabling of a transition in the post-set of the place), \emph{sejour time} (i.e.
+the time interval between arrival and departure of tokens) and \emph{waiting time} (i.e.
+the time interval between enabling of a transition in the post-set of the place and token departure).
 
 Performing an analysis based on the measures computed by log replay gives just a \emph{quantitative} information about the process performance.
-In order to understand the \emph{cause} of performance anomalies that can affect the process behavior, we have to explore the process data.
+In order to understand the \emph{causes} of performance anomalies that can affect the process behavior, one can explore the process data.
 Discovering how data attribute can influence process performance provides useful information in analyzing and optimizing the process services.
 For example, discovering data patterns in correspondence to which some activities need more time for completion than others, helps in making decisions about resources distribution to the process activities or in scheduling activities.
-In addition to the completion time, one could analyze a process under more complex performance metrics such as the \emph{synchronization time}.
+In addition to the completion time, one could analyze a process under more complex performance metrics such as the synchronization time.
 A process with parallel branches and synchronizations can present bottleneck activities that lead to increase the execution time of other activities and consequently the completion time of the entire process.
@@
 For example, a classification problem can be formulated for each synchronization point of the process.
-In this way, the model obtained in correspondence to a synchronization point classifies the process instances based on their attribute value and regarding the synchronization time of the point taken into exam.
+In this way, the classifier obtained in correspondence to a synchronization point classifies the process instances based on their attribute value and regarding the synchronization time of the point taken into exam.
 
@@
 Therefore, the target attribute can assume three possible values: \emph{high\_financial, high\_warehouse} and \emph{normal}.
-Based on the process data extracted from the event log and target attribute computed by a discretization of the log replay results, the dataset for the classification problem is built and presented in table \ref{tab:SaleDataPerf}.
+Based on the process data extracted from the event log and target attribute computed by a discretization of the log replay results, the dataset for the classification problem is built and presented in Table \ref{tab:SaleDataPerf}.
 As done for conformance analysis, the dataset is used as input of the classification algorithm that returns the decision tree presented in Figure \ref{salesPerfDecTree}.
@@
 Note that the plug-in extracts \emph{all} the attributes data recorded in the event log.
-In case, the dataset resulting can be subject of some \emph{feature selection} techniques before using it as a training set for the classification task.
+In case, the dataset resulting can be subject to some \emph{feature selection} techniques before using it as a training set for the classification task.
 
@@
 \item \texttt{Generate Instances With Performance}: the plug-in takes as input an event log, an indication of the synchronization point to examine and the performance result computed by the log replay.
-The output is a training set as presented in table \ref{tab:SaleDataPerf}.
+The output is a training set as presented in Table \ref{tab:SaleDataPerf}.
 \end{comment}
@@
 \end{itemize}
-The second category includes plug-ins for the classification model generation and utilization for the prediction purpose:
+The second category includes plug-ins for the classifier generation and utilization as a predictive model:
 \begin{itemize}

### 2013-09-17 logenerator/genSalesConf.py
commit su master

@@ -141,11 +141,18 @@ def externalHook(t, attrs):
  # return l[0]
 
 def choiceHook(l, attrs):
-  if  ( attrs["salesmanager"] == "Mary" and attrs["clientType"] == "consolidate"):
+  noise = random.randint(1,3)
+  if( attrs["salesmanager"] == "Mary" and attrs["clientType"] == "consolidate" and noise!=3):
         return l[1]
-  return l[0]
-
-
+  #return l[0]
+  else:
+   noise = random.randint(1,9)
+   if( noise != 5 ):
+    return l[0]
+   else: 
+    return l[1]
+  
+	
 process = Sequence([ 
 		Entry("Order", orderHook),
 		Choice([
@@ -175,4 +182,4 @@ process = Sequence([
 		])
 	
 
-generate(process, 250)
+generate(process, 300)

### 2013-09-17 logenerator/genViaggi.py

@@ -54,7 +54,7 @@ import random
 # 	return t0, t1
 
 def orderHook(t, attrs):
-    products = ["bottles", "boxes"]
+    products = ["bottles", "boxes", "equipment"]
     attrs["product"] = products[random.randint(0, len(products)-1)]
     customers = ["A", "B", "C"]
     attrs["customer"] = customers[random.randint(0, len(customers)-1)]
@@ -119,6 +119,14 @@ def externalHook(t, attrs):
     delta1 = random.randint(min_t * 60*60, max_t * 60*60)
     return t0, t1 + delta1
 
+#def choiceHook(l, attrs):
+    #if  attrs["urg"] == "Yes":
+    #    return l[1]
+    #return l[0]
+#  if  ( attrs["product"] == "equipment" ):
+  #      return l[1]
+ # return l[0]
+
 process = Sequence([ 
 		Entry("Order", orderHook),		
 		Par([ Entry("FinancialCheck", financialHook),
@@ -135,4 +143,4 @@ process = Sequence([
 
 		
 
-generate(process, 1)
+generate(process, 250)

### 2013-09-17 src/org/processmining/dataminig/performance/plugin/DMPerfPlugin.java

@@ -131,9 +131,10 @@ public class DMPerfPlugin {
 		  }
 		    
 		  Vector<String> values = new Vector<String>();
+		  values.add("NEGLIGIBLE");
+		  values.add("OK");
 		  values.add("HIGH");
 		  values.add("LOW");
-		  values.add("OK");
 		  values.add("?");
 		  Attribute wekaAttr = new Attribute("Sync", values);
 		  wekaAttrList.add(wekaAttr);
@@ -236,10 +237,7 @@ public class DMPerfPlugin {
 				 float synch = PlaceData.get(p).getSynchTime();
 				 if (synch != 0 )
 				 	 synchTimes.add(new Float(synch * 0.001));				 
-				/* if( synch < 0 )
-						 synchTimes.add( new Float(- synch));				  
-			 	*/
-			 }
+			 	}
 			 float maxTime = 0;
 			 for(Float time : synchTimes){
 				 if (time.floatValue() > maxTime )
@@ -250,10 +248,25 @@ public class DMPerfPlugin {
 			 System.out.println(maxTime);
 			 System.out.println("******************");
 			 
-			 if(maxTime  > 10800 ) 
-				 instance.setValue(wekaAttrindex, "HIGH");
-			 else
+			 //Lineare
+			 /*if(maxTime < 10800 ) 
+				 instance.setValue(wekaAttrindex, "NEGLIGIBLE");
+			 else if (maxTime < 21600)
 				 instance.setValue(wekaAttrindex, "LOW");
+			 else if (maxTime < 32400)
+				 instance.setValue(wekaAttrindex, "OK");
+			 else
+				 instance.setValue(wekaAttrindex, "HIGH"); */
+			 
+			 //Logaritmico
+			 if(maxTime < 8100)
+				 instance.setValue(wekaAttrindex, "LOW");
+			 else if(maxTime < 18900)
+				 instance.setValue(wekaAttrindex, "OK");
+			 else 
+				 instance.setValue(wekaAttrindex, "HIGH");
+
+
 			 
 		  		
 		  	instances.add(instance);

### 2013-09-17 paper/paper2.tex
Merge with guancio branch

--- 2013-09-17 paper/paper2.tex
@@
 \date{today}
+
 \begin{document}
@@
 Next, the same approach is extended to consider the process behavior in terms of performance to discover how data influences the process efficiency.
-Classification is a well-known \emph{machine learning} technique, and it consists of identifying to which category among a given set a new observation belongs.
+Classification is a well-known \emph{machine learning} technique, and it consists of identifying to which category, among a given set, a new observation belongs.
 This is done on the basis of a training set of data containing observations whose category membership is known.
 
-Classification techniques in the Process Mining field have already been explored.
+Classification techniques in the Process Mining field have already been explored \cite{1}\cite{14}.
 In \cite{1} the idea is exploring how data influences the case routing of process flow execution.
@@
 
-After the presentation of a case study in Section \ref{example},
-Section \ref{Background} introduces some background concepts about
-process analysis. First the algorithm log replay is briefly described,
-then conformance and performance analysis are presented. Both kind of
-analysis are illustrated through the business process of Section
-\ref{example}. In Section \ref{ClassifApproach} our approach to
-process analysis based on classification is presented. First a few
-basic concepts about the classification technique are described, next
-the conformance analysis based on classification is presented in
-detail through an illustration with our case study, moreover a
-possible extension of the approach for performance analysis is
-presented. Finally, Section \ref{implementation} describes the
-implementation of our approach provided as \emph{ProM6}\cite{6} plug-ins, as well the methodology to follow in the process analysis based on classification.
+After the presentation of a case study in Section \ref{example}, Section \ref{Background} introduces some background concepts about process analysis.
+First the algorithm log replay is briefly described, then conformance  analysis is presented.
+This analysis is illustrated through the business process of Section \ref{example}.
+In Section \ref{ClassifApproach} our approach to process analysis based on classification is presented.
+First a few basic concepts about the classification technique are described, next the conformance analysis based on classification is presented in detail through an illustration with our case study, moreover a possible extension of the approach for performance analysis is presented.
+Finally, Section \ref{implementation} describes the implementation of our approach provided as \emph{ProM6}\cite{6} plug-ins, as well the methodology to follow in the process analysis based on classification.
 
@@
 Figure \ref{pnet} presents the Petri net model for a sale
-process.
-\footnote{
+process
+\footnote{\small{
 The Petri net of Figure \ref{pnet} is obtained by
@@
   transitions} used to encode the \emph{Join} or \emph{Fork} gateway \cite{2}.
-}
+}}.
 This business process abstractly represents a fragment of the
@@
 
+%Note that the Petri net of Figure \ref{pnet} is obtained by transforming a BPMN model using the algorithm described in \cite{2}, and mapping each activity to a pair of start/end transitions.
+The resulting net has been simplified by removing the unnecessary \emph{invisible transitions} used to encode e.g.
+the \emph{Join} or \emph{Fork} gateway \cite{2}.
+
 \section{Process Analysis}\label{Background}
-
-Process analysis is performed using the Petri net representing the
-process and an event log recording the related executions, also called
-\emph{process instances}. The basic building blocks of event logs are
-events. An event {\itshape e} can be seen as a pair {\itshape e =
-  (a,t) } representing an action {\itshape a } observer at the
-timestamp {\itshape t}. Action and timestamp are denoted respectively
-by $\alpha(e)$ and $\phi(e)$. Events that belong to the same process
-are grouped into {\itshape traces}. Formally, a trace $T$ is a finite
-sequence of events $T[1],..., T[n]$, such that $\phi(T[i]) \leq
-\phi(T[i+1])$ for all  $i \in [1,n)$. A log $L$ is a set of traces,
-recording the activities performed by a system during a finite number
-of process executions. In this paper we assume that
-(i) all traces are instances of the same process
-and
-(ii) for each action there exists a corresponding transition (labeled
-with the same name of the action) in the Petri net.
+%Process analysis is performed using the Petri net representing the process and an event log recording the related executions, also called \emph{process instances}.
+The basic building blocks of event logs are events.
+An event {\itshape e} can be seen as a pair {\itshape e = (a,t) } representing an action {\itshape a } recorded and the corresponding timestamp {\itshape t}.
+Action and timestamp are denoted respectively by $\alpha(e)$ and $\phi(e)$.
+Events that belong to the same process are grouped into {\itshape traces}.
+Formally, a trace $T$ is a finite sequence of events $T[1],..., T[n]$, such that $\phi(T[i]) \leq \phi(T[i+1])$ for all  $i \in [1,n)$.
+A log $L$ is a set of traces, recording the activities performed by a system during a finite number of process executions.
+Process analysis is performed using the Petri net representing the process and an event log recording the related executions, also called \emph{process instances}.
+The basic building blocks of event logs are events.
+An event $e$ can be seen as a pair $e = (a,t)$ representing an action $a$ recorded and the corresponding timestamp $t$.
+Events that belong to the same process are grouped into {\itshape traces}.
+A trace $T$ is a finite sequence of events ordered based on timestamp.
+A log $L$ is a set of traces, recording the activities performed by a system during a finite number of process executions.
+In this paper the following assumptions are made:
+\begin{itemize}
+\item All traces are instances of the same process.
+\item For each action there exist two corresponding transitions in the net that will be denoted, for simplicity, by the same name of the action.
+\end{itemize}
 
 \subsection{Log replay algorithm}\label{logreplayAlg}
-
 The key algorithm exploited to analyze a Petri net model with respect to the log is the {\itshape log replay} algorithm \cite{13}\cite{3}\cite{4}.
@@
 
+\begin{comment}
 \subsection{Performance analysis}\label{PerformanceAnalysis}
@@
 The duration of an activity is an important performance measure as well, and it is given by the waiting time of the only place between the start and the complete transition of the activity.
-
+\end{comment}
 \begin{comment}
@@
 
+\end{comment}
+
 \section{An approach based on classification for Process Analysis}\label{ClassifApproach}
@@
 
-In a classification problem \cite{5}, data is represented by a
-collection of records (called instances), and each of them is
-characterized by a tuple $(\mathbf{x},y)$, where $\mathbf{x}$ is a set
-of attributes and $y$ is a special attribute called \emph{target attribute}.
+In a classification problem \cite{5}, data is represented by a collection of records (called instances), and each of them is characterized by a tuple $(\mathbf{x},y)$, where $\mathbf{x}$ is a set of attributes and $y$ is a special attribute called \emph{target attribute}.
 The target denotes the class to which a record belongs.
@@
 The non-terminal nodes, which include the root and the internal nodes, contain attribute test conditions to filter records that have different characteristics.
+
 \begin{comment}
@@
 \subsection{Extension of the approach for Performance Analysis}\label{ClassPerf}
+Performance analysis of a process can be carried exploiting log replay.
+Since a log contains timestamps, during the replay of a trace it is possible to compute (for each place of the Petri net model) some performance measures such as: \emph{synchronization time} (i.e.
+time interval between arrival of a token in the place and enabling of a transition in the post-set of the place), \emph{sejour time} (i.e.
+the time interval between arrival and departure of tokens) and \emph{waiting time} (i.e.
+the time interval between enabling of a transition in the post-set of the place and token departure).
+
+Performing an analysis based on the measures computed by log replay gives just a \emph{quantitative} information about the process performance.
+In order to understand the \emph{causes} of performance anomalies that can affect the process behavior, one can explore the process data.
 Discovering how data attribute can influence process performance provides useful information in analyzing and optimizing the process services.
 For example, discovering data patterns in correspondence to which some activities need more time for completion than others, helps in making decisions about resources distribution to the process activities or in scheduling activities.
-In addition to the completion time, one could analyze a process under more complex performance metrics such as the \emph{synchronization time}.
+In addition to the completion time, one could analyze a process under more complex performance metrics such as the synchronization time.
 A process with parallel branches and synchronizations can present bottleneck activities that lead to increase the execution time of other activities and consequently the completion time of the entire process.

### 2013-09-19 paper/paper2.tex
aggiunti: conclusioni e future works

@@
 
-\title{Classification Techniques for Conformance and Performance checking in Process Analysis}
-\author{Hind Chfouka, Andrea Corradini, Roberto Guanciale}
-\institute{Department of Computer Science, Università di Pisa}
+\title{Classification Techniques for Conformance and Performance Checking in Process Analysis}
+\author{Hind Chfouka\inst{1}, Andrea Corradini\inst{1} and Roberto Guanciale\inst{2}}
+\institute{Department of Computer Science, University of Pisa, Italy\and School of Computer Science and Communication,
+KTH Royal Institute of Technology, Stockholm, Sweden}
 \date{today}
@@
 \begin{abstract}
-Standard process analysis techniques like conformance checking or performance evaluation are enabled by the event logs that trace the process executions and by the presence of a model that formally represents the process.
-Such analysis techniques use only part of the huge amount of data recorded in the event logs.
-In this paper the goal is to transform this data into useful information for the conformance checking and the performance analysis.
-Through classification technique, we present an approach to investigate how the process data influences its behavior in term of conformance and performance.
+Standard process analysis techniques, like conformance checking or performance evaluation, are enabled by the existence of event logs that trace the process executions and by the presence of a model that formally represents the process.
+Such analysis techniques use only part of the huge amount of data recorded in event logs.
+In this paper the goal is to exploit this data to extract useful information for  conformance checking and  performance analysis.
+We present an approach that using standard classification technique, explores how data influence process behaviors by affecting its conformance or performance.
 \end{abstract}
@@
 \section{Introduction}
-Today, many Information Systems that integrate the concept of business process record all events occurring during the process execution in an event log.
+Today, many Information Systems that support  the concept of business process record all events occurring during the process execution in  event logs.
 An event log represents a trace of the process behavior that can be observed and analyzed in order to tune it with the business objectives pursued.
@@
 
-In order to analyze data arising from event logs, in this paper we present an approach based on \emph{data mining} \cite{5}.
+In order to analyze data arising from event logs, in this paper we present an approach based on \emph{data mining} \cite{tan2005introduction}.
 In particular, we exploit \emph{classification} techniques to find patterns on data in presence of which conformance errors occur.
-Next, the same approach is extended to consider the process behavior in terms of performance to discover how data influences the process efficiency.
+The same approach has been extended to evaluate how data influences  process efficiency, by combining classification problems with performance evaluation: this is just summarized briefly in the paper due to size constraints.
 Classification is a well-known \emph{machine learning} technique, and it consists of identifying to which category, among a given set, a new observation belongs.
@@
 
-Classification techniques in the Process Mining field have already been explored \cite{1}\cite{14}.
-In \cite{1} the idea is exploring how data influences the case routing of process flow execution.
+%% Classification techniques in the Process Mining field have already been explored \cite{DBLP:conf/bpm/RozinatA06}\cite{DBLP:conf/sac/LeoniA13}.
+In \cite{DBLP:conf/bpm/RozinatA06} the idea is exploring how data influences the case routing of process flow execution.
 In this contribution the goal is to strengthen the use of data analysis techniques for the benefit of process analysis.
@@
 First a few basic concepts about the classification technique are described, next the conformance analysis based on classification is presented in detail through an illustration with our case study, moreover a possible extension of the approach for performance analysis is presented.
-Finally, Section \ref{implementation} describes the implementation of our approach provided as \emph{ProM6}\cite{6} plug-ins, as well the methodology to follow in the process analysis based on classification.
+Finally, Section \ref{implementation} describes the implementation of our approach provided as \emph{ProM6}\cite{DBLP:conf/bpm/AalstDGRVW09} plug-ins, as well the methodology to follow in the process analysis based on classification.
 
@@
 There are several formalisms for representing business processes.
-In this paper we represent them using Petri nets \cite{8}\cite{10} since this make possible exploiting some existing process analysis techniques that are based on Petri nets.
-However, in the business management context, Petri nets can be not expressive enough and usually process models are presented as flowcharts, in particular using the standard for the business process modeling: BPMN (Business Process Model and Notation) \cite{9}.
-Transforming a BPMN model into a Petri net is possible thanks to a mapping technique presented in \cite{12}\cite{2}.
+In this paper we represent them using Petri nets \cite{DBLP:conf/ifip/Petri62}\cite{DBLP:journals/jcsc/Aalst98} since this make possible exploiting some existing process analysis techniques that are based on Petri nets.
+However, in the business management context, Petri nets can be not expressive enough and usually process models are presented as flowcharts, in particular using the standard for the business process modeling: BPMN (Business Process Model and Notation) \cite{BPMN}.
+Transforming a BPMN model into a Petri net is possible thanks to a mapping technique presented in \cite{DBLP:conf/wsfm/BruniCFFGS11}.
 \begin{figure}[h]
@@
 The Petri net of Figure \ref{pnet} is obtained by
-transforming a BPMN model using the algorithm of \cite{2}, and mapping
+transforming a BPMN model using the algorithm of \cite{DBLP:conf/wsfm/BruniCFFGS11}, and mapping
 each activity to a pair of start/end transitions. The resulting net
 has been simplified by removing the unnecessary \emph{invisible
-  transitions} used to encode the \emph{Join} or \emph{Fork} gateway \cite{2}.
+  transitions} used to encode the \emph{Join} or \emph{Fork} gateway \cite{DBLP:conf/wsfm/BruniCFFGS11}.
 }}.
@@
 
-%Note that the Petri net of Figure \ref{pnet} is obtained by transforming a BPMN model using the algorithm described in \cite{2}, and mapping each activity to a pair of start/end transitions.
+%Note that the Petri net of Figure \ref{pnet} is obtained by transforming a BPMN model using the algorithm described in \cite{DBLP:conf/wsfm/BruniCFFGS11}, and mapping each activity to a pair of start/end transitions.
 The resulting net has been simplified by removing the unnecessary \emph{invisible transitions} used to encode e.g.
-the \emph{Join} or \emph{Fork} gateway \cite{2}.
+the \emph{Join} or \emph{Fork} gateway \cite{DBLP:conf/wsfm/BruniCFFGS11}.
 
@@
 \subsection{Log replay algorithm}\label{logreplayAlg}
-The key algorithm exploited to analyze a Petri net model with respect to the log is the {\itshape log replay} algorithm \cite{13}\cite{3}\cite{4}.
+The key algorithm exploited to analyze a Petri net model with respect to the log is the {\itshape log replay} algorithm \cite{DBLP:journals/widm/AalstAD12}\cite{DBLP:journals/is/RozinatA08}\cite{DBLP:conf/bpm/Munoz-GamaC10}.
 Given a Petri net model and an event log as input to the algorithm, the output results can be used to check the conformance of traces and to evaluate some performance metrics.
@@
 be used to compute performance measures of the process my measuring the elapsed time between production and consumption of tokens.
-Performance analysis can be applied only to traces that do not generate missing tokens during the replay since such tokens cannot have time information \cite{2} .
+Performance analysis can be applied only to traces that do not generate missing tokens during the replay since such tokens cannot have time information \cite{DBLP:conf/wsfm/BruniCFFGS11} .
 During the log replay the following metrics can be computed for each trace and each place:
@@
 
-In a classification problem \cite{5}, data is represented by a collection of records (called instances), and each of them is characterized by a tuple $(\mathbf{x},y)$, where $\mathbf{x}$ is a set of attributes and $y$ is a special attribute called \emph{target attribute}.
+In a classification problem \cite{tan2005introduction}, data is represented by a collection of records (called instances), and each of them is characterized by a tuple $(\mathbf{x},y)$, where $\mathbf{x}$ is a set of attributes and $y$ is a special attribute called \emph{target attribute}.
 The target denotes the class to which a record belongs.
@@
 
-Figure \ref{figclassApproach} summarizes the general approach for solving a classification problem \cite{5}.
+Figure \ref{figclassApproach} summarizes the general approach for solving a classification problem \cite{tan2005introduction}.
 A \emph{Training set} consisting of records whose class labels are known must be provided to a learning algorithm, and it is used to build a classification model.
@@
 \section{Implementation with ProM}\label{implementation}
-In order to experiment with some business process prototypes, the approach presented in this paper has been implemented as a set of plug-ins that integrates \emph{ProM6} \cite{6}, an open-source framework implementing Process Mining tools, and \emph{Weka} \cite{7}, a data mining framework providing machine learning tools.
+In order to experiment with some business process prototypes, the approach presented in this paper has been implemented as a set of plug-ins that integrates \emph{ProM6} \cite{DBLP:conf/bpm/AalstDGRVW09}, an open-source framework implementing Process Mining tools, and \emph{Weka} \cite{DBLP:journals/sigkdd/HallFHPRW09}, a data mining framework providing machine learning tools.
 We can classify the plug-ins developed in three categories. The first one includes plug-ins for building the dataset needed for the classification:
@@
 It should be clear that the analysis presented on the previous Section assumes the presence of a Petri net model of the business process to be analyzed.
-The model can be expressed in BPMN language as well, but in this case a transformation into a Petri net can be done through the techniques described in \cite{2}.
+The model can be expressed in BPMN language as well, but in this case a transformation into a Petri net can be done through the techniques described in \cite{DBLP:conf/wsfm/BruniCFFGS11}.
 The classification analysis  needs also an event log tracing the process behavior over multiple instances.
@@
 
+\section{Conclusions and Future Work}
+
+We presented some preliminary results of a research activity aimed at applying machine learning techniques, i.e.~ classification algorithms, in the realm of process analysis.
+The idea is to try to identify some correlation between the data value of certain variables in the individual cases of event logs, and the results of conformance checking.
+A successful identification of such values could allow to predict the conformance result  of a case during execution.
+The same technique was applied to performance checking as well, but only briefly summarized here because of space constraints.
+The approach is implemented with a combination of plug-ins of the Process Mining framework ProM and of the classification engine WEKA.
+Some guidelines about their use are presented in a concluding section.
+The approach has been successfully tested on synthetic event logs only.
+An obvious continuation of this work includes experimentations with real event logs: this will allow us to evaluate the scalability of the approach and its robustness in presence of noise.
+
+Classification techniques in the Process Mining field have already been explored in \cite{DBLP:conf/bpm/RozinatA06} and more recently in \cite{DBLP:conf/sac/LeoniA13}.
+In those papers the  idea is to  explore how data influences the case routing of process flow execution, by assigning a classification problem with each decision point in the model.
+In particular, \cite{DBLP:conf/sac/LeoniA13} improves significantly the results of \cite{DBLP:conf/bpm/RozinatA06} by exploiting some recently developed \emph{alignment techniques} between event log cases and the process model.
+We intend to explore how far these techniques can be applied fruitfully in our framework as well.
+
 \addcontentsline{toc}{chapter}{Bibliography} 

### 2013-09-19 paper/paper2.tex
restructured the conformance section

--- 2013-09-17 paper/paper2.tex
@@
 \date{today}
-
 \begin{document}
@@
 \section{Introduction}
+
 Today, many Information Systems that integrate the concept of business process record all events occurring during the process execution in an event log.
@@
 
-After the presentation of a case study in Section \ref{example}, Section \ref{Background} introduces some background concepts about process analysis.
-First the algorithm log replay is briefly described, then conformance  analysis is presented.
-This analysis is illustrated through the business process of Section \ref{example}.
-In Section \ref{ClassifApproach} our approach to process analysis based on classification is presented.
-First a few basic concepts about the classification technique are described, next the conformance analysis based on classification is presented in detail through an illustration with our case study, moreover a possible extension of the approach for performance analysis is presented.
-Finally, Section \ref{implementation} describes the implementation of our approach provided as \emph{ProM6}\cite{6} plug-ins, as well the methodology to follow in the process analysis based on classification.
+After the presentation of a case study in Section \ref{example},
+Section \ref{Background} introduces some background concepts about
+process analysis. First the algorithm log replay is briefly described,
+then conformance and performance analysis are presented. Both kind of
+analysis are illustrated through the business process of Section
+\ref{example}. In Section \ref{ClassifApproach} our approach to
+process analysis based on classification is presented. First a few
+basic concepts about the classification technique are described, next
+the conformance analysis based on classification is presented in
+detail through an illustration with our case study, moreover a
+possible extension of the approach for performance analysis is
+presented. Finally, Section \ref{implementation} describes the
+implementation of our approach provided as \emph{ProM6}\cite{6} plug-ins, as well the methodology to follow in the process analysis based on classification.
 
@@
 Figure \ref{pnet} presents the Petri net model for a sale
-process
-\footnote{\small{
+process.
+\footnote{
 The Petri net of Figure \ref{pnet} is obtained by
@@
   transitions} used to encode the \emph{Join} or \emph{Fork} gateway \cite{2}.
-}}.
+}
 This business process abstractly represents a fragment of the
@@
 
-%Note that the Petri net of Figure \ref{pnet} is obtained by transforming a BPMN model using the algorithm described in \cite{2}, and mapping each activity to a pair of start/end transitions.
-The resulting net has been simplified by removing the unnecessary \emph{invisible transitions} used to encode e.g.
-the \emph{Join} or \emph{Fork} gateway \cite{2}.
+\section{Process Analysis}\label{Background}
 
-\section{Process Analysis}\label{Background}
-%Process analysis is performed using the Petri net representing the process and an event log recording the related executions, also called \emph{process instances}.
-The basic building blocks of event logs are events.
-An event {\itshape e} can be seen as a pair {\itshape e = (a,t) } representing an action {\itshape a } recorded and the corresponding timestamp {\itshape t}.
-Action and timestamp are denoted respectively by $\alpha(e)$ and $\phi(e)$.
-Events that belong to the same process are grouped into {\itshape traces}.
-Formally, a trace $T$ is a finite sequence of events $T[1],..., T[n]$, such that $\phi(T[i]) \leq \phi(T[i+1])$ for all  $i \in [1,n)$.
-A log $L$ is a set of traces, recording the activities performed by a system during a finite number of process executions.
-Process analysis is performed using the Petri net representing the process and an event log recording the related executions, also called \emph{process instances}.
-The basic building blocks of event logs are events.
-An event $e$ can be seen as a pair $e = (a,t)$ representing an action $a$ recorded and the corresponding timestamp $t$.
-Events that belong to the same process are grouped into {\itshape traces}.
-A trace $T$ is a finite sequence of events ordered based on timestamp.
-A log $L$ is a set of traces, recording the activities performed by a system during a finite number of process executions.
-In this paper the following assumptions are made:
-\begin{itemize}
-\item All traces are instances of the same process.
-\item For each action there exist two corresponding transitions in the net that will be denoted, for simplicity, by the same name of the action.
-\end{itemize}
+Process analysis is performed using the Petri net representing the
+process and an event log recording the related executions, also called
+\emph{process instances}. The basic building blocks of event logs are
+events. An event {\itshape e} can be seen as a pair {\itshape e =
+  (a,t) } representing an action {\itshape a } observer at the
+timestamp {\itshape t}. Action and timestamp are denoted respectively
+by $\alpha(e)$ and $\phi(e)$. Events that belong to the same process
+are grouped into {\itshape traces}. Formally, a trace $T$ is a finite
+sequence of events $T[1],..., T[n]$, such that $\phi(T[i]) \leq
+\phi(T[i+1])$ for all  $i \in [1,n)$. A log $L$ is a set of traces,
+recording the activities performed by a system during a finite number
+of process executions. In this paper we assume that
+(i) all traces are instances of the same process
+and
+(ii) for each action there exists a corresponding transition (labeled
+with the same name of the action) in the Petri net.
 
 \subsection{Log replay algorithm}\label{logreplayAlg}
+
 The key algorithm exploited to analyze a Petri net model with respect to the log is the {\itshape log replay} algorithm \cite{13}\cite{3}\cite{4}.
@@
 Non-blocking replay means that whenever a firing of a disabled transition is needed, the algorithm enables the transition either creating artificial tokens in the pre-set or, if possible, firing some invisible transitions; the non-determinism in this procedure is resolved with a suitable cost function.
-%The output of a log replay execution on a trace can be represented as an ordered list $R$ of pairs $(tr, i)$, representing that the transition $tr$ has been fired to mimic the event $T[i]$.
+The output of a log replay execution on a trace can be represented as an ordered list $R$ of pairs $(tr, i)$, representing that the transition $tr$ has been fired to mimic the event $T[i]$.
 
@@
 
-The log replay of the trace $T'$ terminates with remaining tokens \{$p5 \rightarrow 1$, $p13 \rightarrow 1$\} and one missing token \{${p8 \rightarrow 1}$\}.
+The log replay of the trace $T'$ terminates with remaining tokens \{${p5 \rightarrow 1,p13 \rightarrow 1}$\} and one missing token \{${p8 \rightarrow 1}$\}.
 The missing token is created artificially and this fact witness a wrong execution of the event $Notify\_start$.
@@
 
-\begin{comment}
 \subsection{Performance analysis}\label{PerformanceAnalysis}
@@
 The duration of an activity is an important performance measure as well, and it is given by the waiting time of the only place between the start and the complete transition of the activity.
-\end{comment}
+
 \begin{comment}
@@
 At the time $3h:25min:20s$ the transition $Notify\_start$ is fired and a token is produced in $p12$, after $4min+4sec$ the transition $Notify\_complete$ is fired and $tw(p12) = 4min+40s$.
+\end{comment}
 
@@
 
-\end{comment}
-
 \section{An approach based on classification for Process Analysis}\label{ClassifApproach}
@@
 
-In a classification problem \cite{5}, data is represented by a collection of records (called instances), and each of them is characterized by a tuple $(\mathbf{x},y)$, where $\mathbf{x}$ is a set of attributes and $y$ is a special attribute called \emph{target attribute}.
+In a classification problem \cite{5}, data is represented by a
+collection of records (called instances), and each of them is
+characterized by a tuple $(\mathbf{x},y)$, where $\mathbf{x}$ is a set
+of attributes and $y$ is a special attribute called \emph{target attribute}.
 The target denotes the class to which a record belongs.
@@
 
-Figure \ref{figclassApproach} summarizes the general approach for solving a classification problem \cite{5}.
-A \emph{Training set} consisting of records whose class labels are known must be provided to a learning algorithm, and it is used to build a classification model.
-This model is then applied to a \emph{test set}, which consists of records with unknown class labels.
-This application allows the estimation of the \emph{model accuracy} that can be done in different ways.
-The simple one consists in computing the accuracy as the number of the records correctly predicted over the total number of records in the test set.
+Figure \ref{figclassApproach} summarizes the general approach for
+solving a classification problem \cite{5}. A \emph{Training set}
+consisting of records whose class labels are known must be provided to
+a learning algorithm, and it is used to build a classification
+model. This model is then applied to a \emph{test set}, which consists
+of records with unknown class labels.
+This step allows to  estimate the \emph{model accuracy} using one of
+the standard approaches: e.g. the accuracy can be represented by the
+fraction of number the records correctly predicted and the total number of records in the test set.
 
-Many classification models could be used for a classification problem.
-For our purpose we choose the \emph{decision tree} model since the rules detected by the classification are shown explicitly.
-This fact is fundamental for giving an explanation to the process behavior based on the process data values.
+Many classification models can be applied to a classification
+problem. For our purpose we choose the \emph{decision tree} model
+since the rules detected by the classification are shown explicitly,
+providing an overview of how process data affects the
+process behavior.
 Furthermore, this choice is encouraged by the simplicity of decision trees.
@@
 The non-terminal nodes, which include the root and the internal nodes, contain attribute test conditions to filter records that have different characteristics.
-
 \begin{comment}
@@
 \subsection{Classification for Conformance Checking}\label{ClassConformance}
-In order to identify the possible causes of non-conformance, a classification problem can be formulated.
-The classification dataset is extracted from the process data: for each process instance, one will extract some data that will be included as a record in the dataset.
-For the process presented in Section \ref{example}, a record is characterized by the following attribute set: an instance identifier, a client identifier, the client typology (new or consolidated client), the sales manager responsible for the order, the financial officer who conducts the financial evaluation activities, the warehouseman responsible for the warehouse checking, the supplying responsible name in case of a provision, and finally the result communicated to the client for the order issued.
-The conformance result for the process instance provides an additional and important data which takes the role of the target attribute.
-All these attributes are discrete and contribute to build the dataset presented in Table \ref{tab:SaleData} of the classification problem.
-Note that, since the goal of this paper is just to illustrate an approach for exploiting process data in the process analysis, the data presented here is just a synthetic data that was generated with some noise to emulate a real situation.
+In order to identify the possible causes of non-conformance, a
+classification problem can be formulated. The classification dataset
+is extracted from the process data: for each process instance,
+relevant information is  extract and included as a record in the
+dataset.
+For the process presented in Section \ref{example}, a record is
+characterized by the following attributes: an instance identifier, a
+client identifier, the client typology (new or consolidated client),
+the sales manager responsible for the order, the financial officer who
+conducts the financial evaluation activities, the warehouseman
+responsible for the warehouse checking, the supplying responsible name
+in case of a provision, and finally the result communicated to the
+client for the order issued. The conformance result for the process
+instance provides an additional and important data, which takes the
+role of the target attribute. 
+All these attributes are discrete and contribute to build the dataset
+presented in table \ref{tab:SaleData} of the classification
+problem. Notice that, since our goal is to illustrate the application
+of existing data mining techniques to process analysis rather then the
+evaluation of a new data mining algorithm, we use synthetic data that
+are generated to emulate a realistic process executions.
+
+In order to build the classification problem, the attributes characterizing the
+process are extracted from the log, while the conformance attribute is
+computed using the conformance checking algorithm.
+The Petri net in Figure \ref{replayResult}  summarizes the
+conformance analysis executed on the event log:
+edges are labeled with the number of activations done,
+places are labeled with the number of remaining and missing tokens.
+\begin{figure}[h]
+\includegraphics[width=360pt]
+{./items/Sales_PN_resultMod.pdf}
+\caption{Petri net: conformance results.}
+\label{replayResult}
+\end{figure}
+
+The resulting data highlight  that there are 37 out of 250 process
+executions with conformance errors.
+The presence of 37 missing tokens in $p11$ represents a forced
+execution of  transition \emph{Notify\_start}.
+In order to provide an interpretation for this results, we have to
+take into account the non-blocking behavior of the log replay
+algorithm. In this case, the 37 missing tokens in $p11$ represents the
+number of times that that the replay the algorithm failed to mimic the
+event \emph{Notify\_start} since the transition associated to the
+event was not enabled. 
+Moreover, 37 remaining tokens in place $p3$ indicates the number of
+times that the instances financial activities were not performed.
+Taking into account the application scenario, we can conclude that 
+37 instances of the process are not compliant with sale policy since
+they did not executed the financial activity.
+
+% This situation is possible only if the invisible transition $t2$ is not enabled.
+Now, if $t2$ is not enabled that means that there are no tokens in the places $p8$ or $p10$.
+But notice that there are 21 remaining tokens in the place $p10$, so the transition $t2$ is not enabled for those $37$ instances because the place $p8$ does not have any token.
+Moreover, the Petri net in Figure \ref{replayResult} shows 37 remaining tokens in place $p3$, so we can deduce that in those instances financial activities were not performed.
+From these facts, we can conclude that the conformance problems presented in  37 instances are due to a missing execution of the financial activity checking.
+On the other hand, the decision tree reports that the instances with conformance problems are made by consolidate clients.
 
@@
 \normalsize
-Through existing data mining tools for classification, it is possible to build a decision tree as a classification model for the sale process example.
-Starting from a given event log, the attributes characterizing the process are extracted and all the resources needed for the classification algorithm employed are produced.
-The resulting decision tree is presented in Figure \ref{salesDecTree}.
 
@@
 \end{figure}
+
+The results of the conformance checking enable the usage of the
+existing data mining tools, allowing us to mine the decision tree in Figure \ref{salesDecTree} as classification model for the sale process.
 The resulting decision tree describes a data pattern in correspondence to which a process instance could present conformance errors: order managed by the sales manager \emph{Mary} and received from consolidate clients of the organization may not respect the standard sales procedure.
-To get more significant information for our analysis, it is useful to relate what is noticed by the decision tree with the log replay results.
-Figure \ref{replayResult} presents a Petri net that summarizes the conformance analysis conducted on instances recorded in the event log taken in exam.
-Arcs are labeled with the number of activations done.
-Places, whenever they present some remaining or missing tokens, are labeled with the number of these tokens.
-\begin{figure}[h]
-\includegraphics[width=360pt]
-{./items/Sales_PN_resultMod.pdf}
-\caption{Petri net: conformance results.}
-\label{replayResult}
-\end{figure}
+To get more significant information for our analysis,
+From the analysis done for the sales process based on both
+classification and log replay results, a new scenario of the sale
+procedure has emerged: orders done by the consolidate clients of the
+organization are not checked from a financial point of view. This fact
+can be accepted as a valid  behavior in a sale procedure, consequently
+an extension of the business process model is needed to include the
+new scenario (\emph{model extension activity}). Alternatively, the
+misbehaving scenario can be considered as an anomaly in the sales
+procedure. In this case, since the analysis provided accurate
+information about the conformance error, corrective measures can be
+taken in order to avoid errors in the subsequent orders.
 
-The Petri net in Figure \ref{replayResult} shows that there are 37 out of 250 process executions with conformance errors.
-The presence of 37 missing tokens in $p11$ signals a forced execution of  transition \emph{Notify\_start}.
-In order to provide an interpretation for this results, we have to take into account the non-blocking behavior of the log replay algorithm.
-In this case, the 37 missing tokens in $p11$ certifies that during the replay the algorithm needs to mimic the event \emph{Notify\_start} but the transition associated to the event is not enabled in the marking reached by the net, so the algorithm creates artificial tokens for completing the trace replay.
-This situation is possible only if the invisible transition $t2$ is not enabled.
-Now, if $t2$ is not enabled that means that there are no tokens in the places $p8$ or $p10$.
-But notice that there are 21 remaining tokens in the place $p10$, so the transition $t2$ is not enabled for those $37$ instances because the place $p8$ does not have any token.
-Moreover, the Petri net in Figure \ref{replayResult} shows 37 remaining tokens in place $p3$, so we can deduce that in those instances financial activities were not performed.
-From these facts, we can conclude that the conformance problems presented in  37 instances are due to a missing execution of the financial activity checking.
-On the other hand, the decision tree reports that the instances with conformance problems are made by consolidate clients.
-
-From the analysis done for the sales process based on both classification and log replay results, a new scenario of the sale procedure has emerged: orders done by the consolidate clients of the organization are not checked from a financial point of view.
-This fact can be accepted as a possible and reasonable behavior in a sale procedure, consequently an extension of the business process model is needed in order to include the new scenario (\emph{model extension activity}).
-Alternatively, the new scenario can be considered as an anomaly in the sales procedure.
-In this case, since the analysis done provides accurate information about the conformance error, corrective measures can be taken in order to avoid errors in the next orders.
-
-The classifier constructed with this approach can be used also in a predictive way.
-Given a new trace recording a process execution, in order to perform the conformance checking to the process model, the decision tree can be used to predict the conformance result.
-This bring an advantage in terms of the time required by the analysis since the log replay algorithm takes more time than the one needed by a decision tree to classify a new instance.
-Moreover, it is worth noting that whenever the set of attributes needed for the classification is known before the completion of a process execution, it can be interesting to predict the conformance result of the execution and avoid the errors that could possibly happen.
+The classifier constructed with this approach can be used also in a
+predictive way. Given a trace recording a process execution the
+decision tree can be used to predict the conformance result. This
+bring an advantage in terms of the time required by the analysis since
+the log replay algorithm takes more time than execution of a decision
+tree. Moreover, it is worth noting that whenever the set of attributes
+needed for the classification is known before the completion of a
+process execution, the non-conformances predicted by the decision tree can
+be used to promptly alert the stakeholders and activate proper
+countermeasures.
 
 \subsection{Extension of the approach for Performance Analysis}\label{ClassPerf}
-Performance analysis of a process can be carried exploiting log replay.
-Since a log contains timestamps, during the replay of a trace it is possible to compute (for each place of the Petri net model) some performance measures such as: \emph{synchronization time} (i.e.
-time interval between arrival of a token in the place and enabling of a transition in the post-set of the place), \emph{sejour time} (i.e.
-the time interval between arrival and departure of tokens) and \emph{waiting time} (i.e.
-the time interval between enabling of a transition in the post-set of the place and token departure).
-
-Performing an analysis based on the measures computed by log replay gives just a \emph{quantitative} information about the process performance.
-In order to understand the \emph{causes} of performance anomalies that can affect the process behavior, one can explore the process data.
 Discovering how data attribute can influence process performance provides useful information in analyzing and optimizing the process services.
 For example, discovering data patterns in correspondence to which some activities need more time for completion than others, helps in making decisions about resources distribution to the process activities or in scheduling activities.
-In addition to the completion time, one could analyze a process under more complex performance metrics such as the synchronization time.
+In addition to the completion time, one could analyze a process under more complex performance metrics such as the \emph{synchronization time}.
 A process with parallel branches and synchronizations can present bottleneck activities that lead to increase the execution time of other activities and consequently the completion time of the entire process.
@@
 For example, a classification problem can be formulated for each synchronization point of the process.
-In this way, the classifier obtained in correspondence to a synchronization point classifies the process instances based on their attribute value and regarding the synchronization time of the point taken into exam.
+In this way, the model obtained in correspondence to a synchronization point classifies the process instances based on their attribute value and regarding the synchronization time of the point taken into exam.
 
@@
 Therefore, the target attribute can assume three possible values: \emph{high\_financial, high\_warehouse} and \emph{normal}.
-Based on the process data extracted from the event log and target attribute computed by a discretization of the log replay results, the dataset for the classification problem is built and presented in Table \ref{tab:SaleDataPerf}.
+Based on the process data extracted from the event log and target attribute computed by a discretization of the log replay results, the dataset for the classification problem is built and presented in table \ref{tab:SaleDataPerf}.
 As done for conformance analysis, the dataset is used as input of the classification algorithm that returns the decision tree presented in Figure \ref{salesPerfDecTree}.
@@
 Note that the plug-in extracts \emph{all} the attributes data recorded in the event log.
-In case, the dataset resulting can be subject to some \emph{feature selection} techniques before using it as a training set for the classification task.
+In case, the dataset resulting can be subject of some \emph{feature selection} techniques before using it as a training set for the classification task.
 
@@
 \item \texttt{Generate Instances With Performance}: the plug-in takes as input an event log, an indication of the synchronization point to examine and the performance result computed by the log replay.
-The output is a training set as presented in Table \ref{tab:SaleDataPerf}.
+The output is a training set as presented in table \ref{tab:SaleDataPerf}.
 \end{comment}
@@
 \end{itemize}
-The second category includes plug-ins for the classifier generation and utilization as a predictive model:
+The second category includes plug-ins for the classification model generation and utilization for the prediction purpose:
 \begin{itemize}

### 2013-09-19 paper/paper2.tex

--- 2013-09-19 paper/paper2.tex
@@
 
-Figure \ref{figclassApproach} summarizes the general approach for solving a classification problem \cite{tan2005introduction}.
-A \emph{Training set} consisting of records whose class labels are known must be provided to a learning algorithm, and it is used to build a classification model.
-This model is then applied to a \emph{test set}, which consists of records with unknown class labels.
-This application allows the estimation of the \emph{model accuracy} that can be done in different ways.
-The simple one consists in computing the accuracy as the number of the records correctly predicted over the total number of records in the test set.
+Figure \ref{figclassApproach} summarizes the general approach for
+solving a classification problem \cite{tan2005introduction}. A \emph{Training set}
+consisting of records whose class labels are known must be provided to
+a learning algorithm, and it is used to build a classification
+model. This model is then applied to a \emph{test set}, which consists
+of records with unknown class labels.
+This step allows to  estimate the \emph{model accuracy} using one of
+the standard approaches: e.g. the accuracy can be represented by the
+fraction of number the records correctly predicted and the total number of records in the test set.
 
-Many classification models could be used for a classification problem.
-For our purpose we choose the \emph{decision tree} model since the rules detected by the classification are shown explicitly.
-This fact is fundamental for giving an explanation to the process behavior based on the process data values.
+Many classification models can be applied to a classification
+problem. For our purpose we choose the \emph{decision tree} model
+since the rules detected by the classification are shown explicitly,
+providing an overview of how process data affects the
+process behavior.
 Furthermore, this choice is encouraged by the simplicity of decision trees.
@@
 \end{figure}
+
+The results of the conformance checking enable the usage of the
+existing data mining tools, allowing us to mine the decision tree in Figure \ref{salesDecTree} as classification model for the sale process.
 The resulting decision tree describes a data pattern in correspondence to which a process instance could present conformance errors: order managed by the sales manager \emph{Mary} and received from consolidate clients of the organization may not respect the standard sales procedure.
-To get more significant information for our analysis, it is useful to relate what is noticed by the decision tree with the log replay results.
-Figure \ref{replayResult} presents a Petri net that summarizes the conformance analysis conducted on instances recorded in the event log taken in exam.
-Arcs are labeled with the number of activations done.
-Places, whenever they present some remaining or missing tokens, are labeled with the number of these tokens.
-\begin{figure}[h]
-\includegraphics[width=360pt]
-{./items/Sales_PN_resultMod.pdf}
-\caption{Petri net: conformance results.}
-\label{replayResult}
-\end{figure}
+To get more significant information for our analysis,
+From the analysis done for the sales process based on both
+classification and log replay results, a new scenario of the sale
+procedure has emerged: orders done by the consolidate clients of the
+organization are not checked from a financial point of view. This fact
+can be accepted as a valid  behavior in a sale procedure, consequently
+an extension of the business process model is needed to include the
+new scenario (\emph{model extension activity}). Alternatively, the
+misbehaving scenario can be considered as an anomaly in the sales
+procedure. In this case, since the analysis provided accurate
+information about the conformance error, corrective measures can be
+taken in order to avoid errors in the subsequent orders.
 
-The Petri net in Figure \ref{replayResult} shows that there are 37 out of 250 process executions with conformance errors.
-The presence of 37 missing tokens in $p11$ signals a forced execution of  transition \emph{Notify\_start}.
-In order to provide an interpretation for this results, we have to take into account the non-blocking behavior of the log replay algorithm.
-In this case, the 37 missing tokens in $p11$ certifies that during the replay the algorithm needs to mimic the event \emph{Notify\_start} but the transition associated to the event is not enabled in the marking reached by the net, so the algorithm creates artificial tokens for completing the trace replay.
-This situation is possible only if the invisible transition $t2$ is not enabled.
-Now, if $t2$ is not enabled that means that there are no tokens in the places $p8$ or $p10$.
-But notice that there are 21 remaining tokens in the place $p10$, so the transition $t2$ is not enabled for those $37$ instances because the place $p8$ does not have any token.
-Moreover, the Petri net in Figure \ref{replayResult} shows 37 remaining tokens in place $p3$, so we can deduce that in those instances financial activities were not performed.
-From these facts, we can conclude that the conformance problems presented in  37 instances are due to a missing execution of the financial activity checking.
-On the other hand, the decision tree reports that the instances with conformance problems are made by consolidate clients.
-
-From the analysis done for the sales process based on both classification and log replay results, a new scenario of the sale procedure has emerged: orders done by the consolidate clients of the organization are not checked from a financial point of view.
-This fact can be accepted as a possible and reasonable behavior in a sale procedure, consequently an extension of the business process model is needed in order to include the new scenario (\emph{model extension activity}).
-Alternatively, the new scenario can be considered as an anomaly in the sales procedure.
-In this case, since the analysis done provides accurate information about the conformance error, corrective measures can be taken in order to avoid errors in the next orders.
-
-The classifier constructed with this approach can be used also in a predictive way.
-Given a new trace recording a process execution, in order to perform the conformance checking to the process model, the decision tree can be used to predict the conformance result.
-This bring an advantage in terms of the time required by the analysis since the log replay algorithm takes more time than the one needed by a decision tree to classify a new instance.
-Moreover, it is worth noting that whenever the set of attributes needed for the classification is known before the completion of a process execution, it can be interesting to predict the conformance result of the execution and avoid the errors that could possibly happen.
+The classifier constructed with this approach can be used also in a
+predictive way. Given a trace recording a process execution the
+decision tree can be used to predict the conformance result. This
+bring an advantage in terms of the time required by the analysis since
+the log replay algorithm takes more time than execution of a decision
+tree. Moreover, it is worth noting that whenever the set of attributes
+needed for the classification is known before the completion of a
+process execution, the non-conformances predicted by the decision tree can
+be used to promptly alert the stakeholders and activate proper
+countermeasures.
 

### 2013-09-20 paper/paper2.tex
Versione finale

@@
 Process Mining provides a set of techniques for process discovery and analysis.
-Through the knowledge extracted from event logs, process discovery allows the construction of a process model.
+Through the knowledge extracted from event logs, process discovery allows one to construct a process model.
 Process analysis instead assumes the existence of a model and it consists of checking the conformance and performance of the process executions with respect to it.
 
-The motivation of this paper is based on the observation that existing event logs are rich of data, and this data is used only in part by the existing process analysis techniques.
-Since data is considered a valid source of knowledge, developing a technique based on data analysis can provide advantages for  process analysis.
-Therefore, our goal consists of finding a way for transforming the data recorded during the process activities in knowledge useful for the analysis.
-To this aim, one possible approach is to try to understand how data may influence the process behavior.
-In particular, it is interesting to understand how may the data influence the conformance and performance of the process executions.
-Exploring this influence can provide qualitative information about the causes of possible anomalies in the process behavior, facilitating the task of taking corrective measures.
+The work presented in this paper is motivated by the observation that existing event logs are rich of data, but such data is used only in part by the existing process analysis techniques.
+Since data is considered as a valid source of knowledge, developing techniques based on data analysis can provide advantages for  process discovery and analysis.
+This fact has already been observed several times in the literature, and it brought to the development of various approaches where machine learning techniques are applied to workflow or process discovery (see e.g.~\cite{DBLP:conf/ecml/Herbst00,Maruster-PHD,DBLP:conf/bpm/RozinatA06,DBLP:conf/sac/LeoniA13}).
 
-In order to analyze data arising from event logs, in this paper we present an approach based on \emph{data mining} \cite{tan2005introduction}.
-In particular, we exploit \emph{classification} techniques to find patterns on data in presence of which conformance errors occur.
+Here instead, we consider more specifically process analysis.
+Our goal is to find a way to transform the data recorded during the process activities in knowledge useful for the analysis.
+In particular,  we try to determine how data may affect the conformance and performance of process executions.
+Identifying this influence can provide qualitative information about the causes of possible anomalies in the process behaviors, facilitating the task of taking corrective measures.
+
+In order to analyze data contained in event logs, we propose an approach based on {machine learning}~\cite{tan2005introduction}, where we exploit \emph{classification} techniques to find patterns on data in presence of which conformance errors occur.
 The same approach has been extended to evaluate how data influences  process efficiency, by combining classification problems with performance evaluation: this is just summarized briefly in the paper due to size constraints.
-Classification is a well-known \emph{machine learning} technique, and it consists of identifying to which category, among a given set, a new observation belongs.
+Classification is a well-known {machine learning} technique which consists of identifying to which category, among a given set, a new observation belongs.
 This is done on the basis of a training set of data containing observations whose category membership is known.
 
-%% Classification techniques in the Process Mining field have already been explored \cite{DBLP:conf/bpm/RozinatA06}\cite{DBLP:conf/sac/LeoniA13}.
-In \cite{DBLP:conf/bpm/RozinatA06} the idea is exploring how data influences the case routing of process flow execution.
+%% Classification techniques in the Process Mining field have already been explored~\cite{DBLP:conf/bpm/RozinatA06}\cite{DBLP:conf/sac/LeoniA13}.
+In~\cite{DBLP:conf/bpm/RozinatA06} the idea is exploring how data influences the case routing of process flow execution.
 In this contribution the goal is to strengthen the use of data analysis techniques for the benefit of process analysis.
@@
 
-After the presentation of a case study in Section \ref{example}, Section \ref{Background} introduces some background concepts about process analysis.
-First the algorithm log replay is briefly described, then conformance  analysis is presented.
-This analysis is illustrated through the business process of Section \ref{example}.
-In Section \ref{ClassifApproach} our approach to process analysis based on classification is presented.
-First a few basic concepts about the classification technique are described, next the conformance analysis based on classification is presented in detail through an illustration with our case study, moreover a possible extension of the approach for performance analysis is presented.
-Finally, Section \ref{implementation} describes the implementation of our approach provided as \emph{ProM6}\cite{DBLP:conf/bpm/AalstDGRVW09} plug-ins, as well the methodology to follow in the process analysis based on classification.
+After the presentation of a case study in Section~\ref{example}, Section~\ref{Background} introduces some background concepts about process analysis.
+First the log replay algorithm is briefly described, then conformance  analysis is introduced using the business process of Section~\ref{example}.
+In Section~\ref{ClassifApproach} our approach to process analysis based on classification is presented.
+First a few basic concepts about the classification technique are described, next the conformance analysis based on classification is presented in detail through an illustration with our case study.
+A possible extension of the approach for performance analysis is briefly discussed presented.
+Finally, Section~\ref{implementation} describes the implementation of our approach provided as \emph{ProM6}\cite{DBLP:conf/bpm/AalstDGRVW09} plug-ins, as well the methodology to follow in order to analyse processes using our approach.
 
@@
 There are several formalisms for representing business processes.
-In this paper we represent them using Petri nets \cite{DBLP:conf/ifip/Petri62}\cite{DBLP:journals/jcsc/Aalst98} since this make possible exploiting some existing process analysis techniques that are based on Petri nets.
-However, in the business management context, Petri nets can be not expressive enough and usually process models are presented as flowcharts, in particular using the standard for the business process modeling: BPMN (Business Process Model and Notation) \cite{BPMN}.
-Transforming a BPMN model into a Petri net is possible thanks to a mapping technique presented in \cite{DBLP:conf/wsfm/BruniCFFGS11}.
+In this paper we represent them using Petri nets~\cite{DBLP:conf/ifip/Petri62,DBLP:journals/jcsc/Aalst98} since this makes possible to exploit some existing process analysis techniques that are based on such nets.
+However, in a business management context, Petri nets can be not expressive enough and usually process models are presented as workflows, for example using a standard notation for the business process modeling: BPMN (Business Process Model and Notation)~\cite{BPMN}.
+Transforming a BPMN model into a Petri net is possible thanks to a mapping technique presented in~\cite{DBLP:conf/wsfm/BruniCFFGS11}.
 \begin{figure}[h]
@@
 {./items/Sales_PN.pdf}
-\caption{Petri net for a a sale process}
+\caption{Petri net for a sale process}
 \label{pnet}
 \end{figure}
-
-Figure \ref{pnet} presents the Petri net model for a sale
-process
-\footnote{\small{
-The Petri net of Figure \ref{pnet} is obtained by
-transforming a BPMN model using the algorithm of \cite{DBLP:conf/wsfm/BruniCFFGS11}, and mapping
+%
+Figure~\ref{pnet} presents the Petri net model for a sale
+process.\footnote{\small{
+The Petri net of Figure~\ref{pnet} is obtained by
+transforming a BPMN model using the algorithm of~\cite{DBLP:conf/wsfm/BruniCFFGS11}, and mapping
 each activity to a pair of start/end transitions. The resulting net
 has been simplified by removing the unnecessary \emph{invisible
-  transitions} used to encode the \emph{Join} or \emph{Fork} gateway \cite{DBLP:conf/wsfm/BruniCFFGS11}.
-}}.
+  transitions} used to encode the \emph{Join} or \emph{Fork} gateway~\cite{DBLP:conf/wsfm/BruniCFFGS11}.
+}} 
 This business process abstractly represents a fragment of the
@@
 process continues with some activities that can be done in parallel,
-since they do not present any dependencies. The \emph{FinancialCheck}
+since they do not present any dependency. The \emph{FinancialCheck}
 activity represents the financial analysis of the order  (i.e
@@
 
-%Note that the Petri net of Figure \ref{pnet} is obtained by transforming a BPMN model using the algorithm described in \cite{DBLP:conf/wsfm/BruniCFFGS11}, and mapping each activity to a pair of start/end transitions.
+%Note that the Petri net of Figure~\ref{pnet} is obtained by transforming a BPMN model using the algorithm described in~\cite{DBLP:conf/wsfm/BruniCFFGS11}, and mapping each activity to a pair of start/end transitions.
 The resulting net has been simplified by removing the unnecessary \emph{invisible transitions} used to encode e.g.
-the \emph{Join} or \emph{Fork} gateway \cite{DBLP:conf/wsfm/BruniCFFGS11}.
+the \emph{Join} or \emph{Fork} gateway~\cite{DBLP:conf/wsfm/BruniCFFGS11}.
 
@@
 Process analysis is performed using the Petri net representing the process and an event log recording the related executions, also called \emph{process instances}.
-The basic building blocks of event logs are events.
-An event $e$ can be seen as a pair $e = (a,t)$ representing an action $a$ recorded and the corresponding timestamp $t$.
-Events that belong to the same process are grouped into {\itshape traces}.
-A trace $T$ is a finite sequence of events ordered based on timestamp.
-A log $L$ is a set of traces, recording the activities performed by a system during a finite number of process executions.
-In this paper the following assumptions are made:
+The basic building blocks of event logs are \emph{events}, that record the execution of the process actions.
+Therefore an event $e$ can be seen as a tuple including the action recorded, a corresponding timestamp, and possibly other attributes describing relevant information about the state of the process.
+%in gen pair $e = (a,t)$ representing an action $a$ recorded and the corresponding timestamp $t$.  
+Events that belong to the same process instance are grouped in a {\itshape trace}, where events are ordered 
+% A trace $T$ is a finite sequence of events ordered 
+by timestamp. An \emph{event log} is a set of traces, recording the activities performed by a system during a finite number of process executions.  In this paper the following assumptions are made:
 \begin{itemize}
 \item All traces are instances of the same process.
-\item For each action there exist two corresponding transitions in the net that will be denoted, for simplicity, by the same name of the action.
+\item For each action there exists a corresponding transition in the net that will be denoted, for simplicity, by the same name of the action.
 \end{itemize}
@@
 \subsection{Log replay algorithm}\label{logreplayAlg}
-The key algorithm exploited to analyze a Petri net model with respect to the log is the {\itshape log replay} algorithm \cite{DBLP:journals/widm/AalstAD12}\cite{DBLP:journals/is/RozinatA08}\cite{DBLP:conf/bpm/Munoz-GamaC10}.
-Given a Petri net model and an event log as input to the algorithm, the output results can be used to check the conformance of traces and to evaluate some performance metrics.
-For each trace in the log, the algorithm starts by placing one token in the start place of the net.
-For each event in the trace the corresponding transition is fired assuming a {\itshape non-blocking} behavior of the algorithm, then the marking of the net is updated.
-Non-blocking replay means that whenever a firing of a disabled transition is needed, the algorithm enables the transition either creating artificial tokens in the pre-set or, if possible, firing some invisible transitions; the non-determinism in this procedure is resolved with a suitable cost function.
-%The output of a log replay execution on a trace can be represented as an ordered list $R$ of pairs $(tr, i)$, representing that the transition $tr$ has been fired to mimic the event $T[i]$.
+The key algorithm exploited to analyze a Petri net model with respect to an event log is the {\itshape log replay} algorithm (see e.g.~\cite{DBLP:journals/is/RozinatA08}).
+%DBLP:journals/widm/AalstAD12,
+%,DBLP:conf/bpm/Munoz-GamaC10
+Given a Petri net model and an event log as input to the algorithm, for each trace in the log, the algorithm starts by placing one token in the start place of the net.
+For each event in the trace the corresponding transition is fired assuming a  non-blocking behavior, then the marking of the net is updated.
+\emph{Non-blocking replay} means that whenever a firing of a disabled transition is needed, the algorithm enables the transition either by creating artificial tokens in the pre-set or, if possible, by firing some invisible transitions; the non-determinism in this procedure is resolved with a suitable cost function.
+For each trace, the result of the log replay execution is a map indicating which transitions of the model have been fired for each event in the trace, and additionally which tokens were artificially created and which remained  upon completion of the trace.
+%% sequence on a trace can be represented as an ordered list $R$ of pairs $(tr, i)$, representing that the transition $tr$ has been fired to mimic the event $T[i]$.
+The output of the log replay algorithm  can be used to check the conformance of the traces.
+% and to evaluate some performance metrics.
 
@@
 Conformance problems can be discovered by analyzing tokens artificially created during the replay {\itshape (the missing tokens)} and tokens not consumed {\itshape (the remaining tokens)}.
-Figure \ref{ConfLog}  presents two different traces, $T$ and $T'$, of the business process presented in Section \ref{example}.
-The log replay of the trace $T$, which is compliant with the Petri net, terminates with a marking containing one token in the end place \{${p13 \rightarrow 1}$\} and without reporting any missing tokens.
+Figure~\ref{ConfLog}  presents two different traces, $T$ and $T'$, of the business process presented in Section~\ref{example}, where only the actions and the timestamps are reported.
+The log replay of trace $T$, which is compliant with the Petri net, terminates with a single remaining
+%marking containing one 
+token in the end place \{${p13 \rightarrow 1}$\} and without reporting  missing tokens.
 
@@
 {./items/logConforme.pdf}
-\caption{Conform ($T$) and non-conform ($T'$) traces}
+\caption{Conforming ($T$) and non-conforming ($T'$) traces}
 \label{ConfLog}
@@
 The log replay of the trace $T'$ terminates with remaining tokens \{$p5 \rightarrow 1$, $p13 \rightarrow 1$\} and one missing token \{${p8 \rightarrow 1}$\}.
-The missing token is created artificially and this fact witness a wrong execution of the event $Notify\_start$.
+The missing token is created artificially and this fact witnesses a wrong execution of the event \emph{Notify\_start}.
 In fact, in $T'$ this event is executed before the termination of the activity \emph{FinancialCheck}: this is interpreted as a non-conformance to the process model.
@@
 be used to compute performance measures of the process my measuring the elapsed time between production and consumption of tokens.
-Performance analysis can be applied only to traces that do not generate missing tokens during the replay since such tokens cannot have time information \cite{DBLP:conf/wsfm/BruniCFFGS11} .
+Performance analysis can be applied only to traces that do not generate missing tokens during the replay since such tokens cannot have time information~\cite{DBLP:conf/wsfm/BruniCFFGS11} .
 During the log replay the following metrics can be computed for each trace and each place:
@@
 \begin{comment}
-To clarify the metrics evaluated by this technique, as done with conformance analysis, we exploit the Petri net  presented in Section \ref{example} and the trace T presented in Figure \ref{ConfLog}.
+To clarify the metrics evaluated by this technique, as done with conformance analysis, we exploit the Petri net  presented in Section~\ref{example} and the trace T presented in Figure~\ref{ConfLog}.
 The replay starts with firing the transition $Order\_start$ at time $0s$, so a token arrives at $p2$ at time $0s$.
@@
 $p10$. For example, the synchronization time is greater than zero for
-$p8$ in the trace $T$ in Figure \ref{ConfLog}, representing that, for
+$p8$ in the trace $T$ in Figure~\ref{ConfLog}, representing that, for
 this particular process instance, the branch performing the financial
@@
 
-\section{An approach based on classification for Process Analysis}\label{ClassifApproach}
-Employing data mining techniques for process analysis is encouraged by the presence of a huge amount of data recorded in event logs under \emph{attribute} format.
+%\section{An approach based on classification for Process Analysis}\label{ClassifApproach}
+\section{Classification for Process Analysis}\label{ClassifApproach}
+Exploiting machine learning techniques for process analysis is encouraged by the presence of a huge amount of data recorded as attributes in event logs.
 The implicit information contained in those data could be significant for the process behavior analysis.
@@
 For this reason, we are interested in discovering how the process data may influence the conformance and the performance of the process executions.
-This is done with an approach based on \emph{classification}, a classical data mining technique able to detect patterns on data in correspondence to which the process assumes a specific behavior.
+This is done with an approach based on \emph{classification}, a classical machine learning technique able to detect patterns on data in correspondence to which the process assumes a specific behavior.
 
@@
 
-In a classification problem \cite{tan2005introduction}, data is represented by a collection of records (called instances), and each of them is characterized by a tuple $(\mathbf{x},y)$, where $\mathbf{x}$ is a set of attributes and $y$ is a special attribute called \emph{target attribute}.
-The target denotes the class to which a record belongs.
+In a classification problem~\cite{tan2005introduction}, data is represented by a collection of records (called instances), and each of them is characterized by a tuple $(\mathbf{x},y)$, where $\mathbf{x}$ is a set of attributes and $y$ is a special one called \emph{target attribute}.
+The value of the target attribute is a label identifying   the class to which the record belongs.
 The goal is to learn a classification model that maps each attribute set $\mathbf{x}$ to one of the predefined class labels $y$.
 
-A classification model can be useful for a descriptive purpose: it provides an explanatory tool to distinguish between objects of different classes, and for a predictive purpose: the model can be used to predict the class label of new records.
+A classification model can be useful for both a descriptive and a predictive purpose.
+In fact it provides an explanatory tool to distinguish between objects of different classes, and at the same time it can be used to predict the class label of new records.
 In this way the model acts as a black box that automatically assigns a class label to an attribute set of a new record.
 
-There is a general approach to solve a classification problem from a dataset, that is independent from the specific classification model.
-Each classification technique employs a learning algorithm to identify a model that best fit the relationship between the attribute set and class label of the input data.
-The model generated should both fit the input data well, and correctly predict the class labels of new records.
-Finding the right trade-off between this two purposes is the most delicate part of a classification technique, and in general of any machine learning task.
+%purpose: it provides an explanatory tool to distinguish between objects of different classes, and for a predictive purpose: the model can be used to predict the class label of new records.
+In this way the model acts as a black box that automatically assigns a class label to an attribute set of a new record.
+
+There is a general approach to solve a classification problem from a given dataset that is independent from the specific classification model.
+Each classification technique uses a learning algorithm to identify a model that best fits the relationship between the attribute sets and the class labels of the input data.
+The generated model should both fit the input data well, and correctly predict the class labels of new records.
+Finding the right trade-off between this two goals is the most delicate part of a classification technique, and in general of any machine learning task.
 
@@
 
-Figure \ref{figclassApproach} summarizes the general approach for
-solving a classification problem \cite{tan2005introduction}. A \emph{Training set}
+Figure~\ref{figclassApproach} summarizes the general approach for
+solving a classification problem~\cite{tan2005introduction}. A \emph{training set}
 consisting of records whose class labels are known must be provided to
-a learning algorithm, and it is used to build a classification
+the learning algorithm, and it is used to build the classification
 model. This model is then applied to a \emph{test set}, which consists
@@
 This step allows to  estimate the \emph{model accuracy} using one of
-the standard approaches: e.g. the accuracy can be represented by the
-fraction of number the records correctly predicted and the total number of records in the test set.
+the standard approaches. For example, the accuracy can be represented by the
+percentage 
+%fraction of number the 
+of records correctly predicted 
+with respect to the total number of records in the test set.
 
@@
 Furthermore, this choice is encouraged by the simplicity of decision trees.
-In a decision tree, a class label (possible value of the attribute target) is assigned to each leaf node.
+In a decision tree, a class label (possible value of the target attribute) is assigned to each leaf node.
 The non-terminal nodes, which include the root and the internal nodes, contain attribute test conditions to filter records that have different characteristics.
@@
 While some of the trees are more accurate than others, finding the optimal tree is computationally infeasible because of the exponential size of the search space.
-However, efficient algoritms have been developed to induce a reasonably accurate decision tree in a reasonable amount of time.
-These algorithms usually employ a greedy strategy that grows a decision tree by making a series of decisions about which attribute to use for partioning the data.
+However, efficient algorithms have been developed to induce a reasonably accurate decision tree in a reasonable amount of time.
+These algorithms usually employ a greedy strategy that grows a decision tree by making a series of decisions about which attribute to use for partitioning the data.
 For instance Hunt's algorithm is one of them and it represents the basis of many existing decision tree induction algorithms, including $C4.5$ algorithm used for our purposes.
@@
 is extracted from the process data: for each process instance,
-relevant information is  extract and included as a record in the
+relevant information is  extracted and included as a record in the
 dataset.
-For the process presented in Section \ref{example}, a record is
-characterized by the following attributes: an instance identifier, a
+For the process presented in Section~\ref{example}, a record includes 
+the following attributes: an instance identifier, a 
 client identifier, the client typology (new or consolidated client),
@@
 in case of a provision, and finally the result communicated to the
-client for the order issued. The conformance result for the process
-instance provides an additional and important data, which takes the
-role of the target attribute. 
+client for the order issued. Furthemore, as target attribute 
+we associate with each record the conformance result for that instance, 
+which is a boolean value that is extracted from the output of the log replay.
+% process
+%instance provides an additional and important data, 
 All these attributes are discrete and contribute to build the dataset
-presented in table \ref{tab:SaleData} of the classification
-problem. Notice that, since our goal is to illustrate the application
-of existing data mining techniques to process analysis rather then the
-evaluation of a new data mining algorithm, we use synthetic data that
-are generated to emulate a realistic process executions.
+presented in Table~\ref{tab:SaleData} of the classification
+problem. In this preliminary work, 
+in order to 
+%Notice that, since our goal is to 
+explore the applicability
+of existing machine learning  techniques to process analysis,
+%rather then the
+%evaluation of a new data mining algorithm, 
+we just used synthetic data
+generated to emulate process executions.
 
@@
 computed using the conformance checking algorithm.
-The Petri net in Figure \ref{replayResult}  summarizes the
-conformance analysis executed on the event log:
-edges are labeled with the number of activations done,
+The Petri net of Figure~\ref{replayResult}  shows the
+output of the conformance analysis executed on the event log:
+edges are labeled with the number of times they were activated,
 places are labeled with the number of remaining and missing tokens.
@@
 executions with conformance errors.
-The presence of 37 missing tokens in $p11$ represents a forced
-execution of  transition \emph{Notify\_start}.
-In order to provide an interpretation for this results, we have to
-take into account the non-blocking behavior of the log replay
-algorithm. In this case, the 37 missing tokens in $p11$ represents the
-number of times that that the replay the algorithm failed to mimic the
-event \emph{Notify\_start} since the transition associated to the
-event was not enabled. 
-Moreover, 37 remaining tokens in place $p3$ indicates the number of
-times that the instances financial activities were not performed.
+The presence of 37 missing tokens in $p11$ represents 
+the
+number of times that the log replay  algorithm forced the execution of transition
+\emph{Notify\_start}, even if it was not enabled, in order to mimic a corresponding event 
+in the trace.
+
+Moreover, 37 remaining tokens in place $p3$ indicate the number of
+times that the financial check activities were not performed.
 Taking into account the application scenario, we can conclude that 
-37 instances of the process are not compliant with sale policy since
-they did not executed the financial activity.
+37 instances of the process are not compliant with the sale policy since
+they did not execute the financial check activity.
 
@@
 But notice that there are 21 remaining tokens in the place $p10$, so the transition $t2$ is not enabled for those $37$ instances because the place $p8$ does not have any token.
-Moreover, the Petri net in Figure \ref{replayResult} shows 37 remaining tokens in place $p3$, so we can deduce that in those instances financial activities were not performed.
+Moreover, the Petri net in Figure~\ref{replayResult} shows 37 remaining tokens in place $p3$, so we can deduce that in those instances financial activities were not performed.
 From these facts, we can conclude that the conformance problems presented in  37 instances are due to a missing execution of the financial activity checking.
@@
 
-The results of the conformance checking enable the usage of the
-existing data mining tools, allowing us to mine the decision tree in Figure \ref{salesDecTree} as classification model for the sale process.
-The resulting decision tree describes a data pattern in correspondence to which a process instance could present conformance errors: order managed by the sales manager \emph{Mary} and received from consolidate clients of the organization may not respect the standard sales procedure.
-To get more significant information for our analysis,
-From the analysis done for the sales process based on both
+The results of the conformance checking enable the use of 
+existing machine learning  tools, allowing us to mine the decision tree of Figure~\ref{salesDecTree} as classification model for the sale process.
+The resulting decision tree describes a data pattern in correspondence to which a process instance could present conformance errors: order managed by the sales manager \emph{Mary} and received from consolidate clients of the organization may not respect the standard sale procedure.
+%To get more significant information for our analysis,
+
+From the analysis done for the sale process based on both
 classification and log replay results, a new scenario of the sale
@@
 organization are not checked from a financial point of view. This fact
-can be accepted as a valid  behavior in a sale procedure, consequently
+can be accepted as a valid  behavior in a sale procedure, and consequently
 an extension of the business process model is needed to include the
 new scenario (\emph{model extension activity}). Alternatively, the
-misbehaving scenario can be considered as an anomaly in the sales
-procedure. In this case, since the analysis provided accurate
+misbehaving scenario can be considered as an anomaly in the sale
+procedure. In this case, since the analysis provides accurate
 information about the conformance error, corrective measures can be
-taken in order to avoid errors in the subsequent orders.
+taken in order to prevent such errors in the future.
 
@@
 decision tree can be used to predict the conformance result. This
-bring an advantage in terms of the time required by the analysis since
-the log replay algorithm takes more time than execution of a decision
+brings an advantage in terms of the time required by the analysis, since
+the replay of a process instance  takes more time than its 
+classification using a decision
 tree. Moreover, it is worth noting that whenever the set of attributes
@@
 
-\subsection{Extension of the approach for Performance Analysis}\label{ClassPerf}
+\subsection{Extension of the approach to Performance Analysis}\label{ClassPerf}
 Performance analysis of a process can be carried exploiting log replay.
-Since a log contains timestamps, during the replay of a trace it is possible to compute (for each place of the Petri net model) some performance measures such as: \emph{synchronization time} (i.e.
+Since an event log contains timestamps, during the replay of a trace it is possible to compute (for each place of the Petri net model) some performance measures such as: \emph{synchronization time} (i.e.
 time interval between arrival of a token in the place and enabling of a transition in the post-set of the place), \emph{sejour time} (i.e.
@@
 A process with parallel branches and synchronizations can present bottleneck activities that lead to increase the execution time of other activities and consequently the completion time of the entire process.
-To find out the possible data influences on synchronization time, the approach presented in Section \ref{ClassConformance} for conformance analysis through classification can be easily extended to performance analysis.
+To find out the possible data influences on synchronization time, the approach presented in Section~\ref{ClassConformance} for conformance analysis through classification can be easily extended to performance analysis.
 For example, a classification problem can be formulated for each synchronization point of the process.
@@
 In this way, the model obtained in correspondence to a synchronization point classifies the process instances based on their attribute value and regarding the synchronization time of the point taken into exam.
-For the sales process presented in Section \ref{example}, a single point of synchronization is identified.
-In fact, the Petri net in Figure \ref{pnet} has two parallel branches, one for the financial activity and the other for the warehouse activities.
+For the sales process presented in Section~\ref{example}, a single point of synchronization is identified.
+In fact, the Petri net in Figure~\ref{pnet} has two parallel branches, one for the financial activity and the other for the warehouse activities.
 We are interested in analyzing the synchronization between these branches.
-Indeed, the synchronization time of interest is measured in correspondence to the places $p(8)$ and $p(10)$: $tsc(p8)$, $tsc(p10)$ which are computed by the log replay algorithm as explained in Section \ref{Background}.
+Indeed, the synchronization time of interest is measured in correspondence to the places $p(8)$ and $p(10)$: $tsc(p8)$, $tsc(p10)$ which are computed by the log replay algorithm as explained in Section~\ref{Background}.
 
@@
 It should be noted that classification deals with discrete or categorical target attribute.
-Since the performance metrics, the synchronization time in this case, are expressed as continous values, a discretization phase is required  during the formulation of the classification problem.
-Let us choose the attribute target as a categorical attribute: given a treshold $S$, if $|tsc(p8)- tsc(p10)|$ exceeds the value $S$, then the target attribute indicates the branch with the high synchronization time, otherwise the target indicates that the instance presents a normal synchronization time.
+Since the performance metrics, the synchronization time in this case, are expressed as continuous values, a discretization phase is required  during the formulation of the classification problem.
+Let us choose the attribute target as a categorical attribute: given a threshold $S$, if $|tsc(p8)- tsc(p10)|$ exceeds the value $S$, then the target attribute indicates the branch with the high synchronization time, otherwise the target indicates that the instance presents a normal synchronization time.
 Therefore, the target attribute can assume three possible values: \emph{high\_financial, high\_warehouse} and \emph{normal}.
-Based on the process data extracted from the event log and target attribute computed by a discretization of the log replay results, the dataset for the classification problem is built and presented in Table \ref{tab:SaleDataPerf}.
-As done for conformance analysis, the dataset is used as input of the classification algorithm that returns the decision tree presented in Figure \ref{salesPerfDecTree}.
+Based on the process data extracted from the event log and target attribute computed by a discretization of the log replay results, the dataset for the classification problem is built and presented in Table~\ref{tab:SaleDataPerf}.
+As done for conformance analysis, the dataset is used as input of the classification algorithm that returns the decision tree presented in Figure~\ref{salesPerfDecTree}.
 
@@
 \section{Implementation with ProM}\label{implementation}
-In order to experiment with some business process prototypes, the approach presented in this paper has been implemented as a set of plug-ins that integrates \emph{ProM6} \cite{DBLP:conf/bpm/AalstDGRVW09}, an open-source framework implementing Process Mining tools, and \emph{Weka} \cite{DBLP:journals/sigkdd/HallFHPRW09}, a data mining framework providing machine learning tools.
+In order to experiment with some business process prototypes, the approach presented in this paper has been implemented as a set of plug-ins that integrates \emph{ProM6}~\cite{DBLP:conf/bpm/AalstDGRVW09}, an open-source framework implementing Process Mining tools, and \emph{Weka}~\cite{DBLP:journals/sigkdd/HallFHPRW09}, a data mining framework providing machine learning tools.
 We can classify the plug-ins developed in three categories. The first one includes plug-ins for building the dataset needed for the classification:
 \begin{itemize}
-\item \texttt{Generate Instances With Conformance}: this plug-in takes as input an event log and its conformance results computed by log replay.
-It returns as output a training set as shown in Table \ref{tab:SaleData} in the format needed by Weka classification tool.
+\item \texttt{Generate Instances with Conformance}: this plug-in takes as input an event log and its conformance results computed by log replay.
+It returns as output a training set as shown in Table~\ref{tab:SaleData} in the format needed by the Weka classification tool.
 Note that the plug-in extracts \emph{all} the attributes data recorded in the event log.
-In case, the dataset resulting can be subject to some \emph{feature selection} techniques before using it as a training set for the classification task.
+If necessary, the dataset resulting can be subject to some \emph{feature selection} techniques before using it as a training set for the classification task.
 
 \begin{comment}
-\item \texttt{Generate Instances With Performance}: the plug-in takes as input an event log, an indication of the synchronization point to examine and the performance result computed by the log replay.
-The output is a training set as presented in Table \ref{tab:SaleDataPerf}.
+\item \texttt{Generate Instances with Performance}: the plug-in takes as input an event log, an indication of the synchronization point to examine and the performance result computed by the log replay.
+The output is a training set as presented in Table~\ref{tab:SaleDataPerf}.
 \end{comment}
@@
 \begin{itemize}
-\item \texttt{Generate Classifier}: given a training set, the plug-in generates a decision tree model using $J48$ algorithm, an implementation of the algorithm $C4.5$ developed by Weka.
-\item \texttt{Classify Instances}: given a set of non-classified instances and a classifier model, this plug-in simply classify the instances according to the classifier.
+\item \texttt{Generate Classifier}: given a training set, the plug-in generates a decision tree model using the $J48$ algorithm, an implementation of the algorithm $C4.5$ provided by Weka.
+\item \texttt{Classify Instances}: given a set of non-classified instances and a classifier model, this plug-in simply classifies the instances according to the classifier.
 \end{itemize}
@@
 
-It should be clear that the analysis presented on the previous Section assumes the presence of a Petri net model of the business process to be analyzed.
-The model can be expressed in BPMN language as well, but in this case a transformation into a Petri net can be done through the techniques described in \cite{DBLP:conf/wsfm/BruniCFFGS11}.
+It should be clear that the analysis presented in the previous sections assumes the presence of a Petri net model of the business process to be analyzed.
+The model can be expressed in BPMN  as well, but in this case a transformation into a Petri net can be done through the techniques described in~\cite{DBLP:conf/wsfm/BruniCFFGS11}.
 The classification analysis  needs also an event log tracing the process behavior over multiple instances.
-Given a process model $M$ and an event log $L$, the methodology to be followed for the conformance analysis with the approach presented in Section \ref{ClassConformance} can be summarized by these steps:
+Given a process model $M$ and an event log $L$, the methodology to be followed for the conformance analysis with the approach presented in Section~\ref{ClassConformance} can be summarized by these steps:
 \begin{enumerate}
 
-\item \label{step1} \textbf{Log replay analysis}: this phase consists in an execution of the log replay with  the model $M$ and the log $L$ as inputs.
-The results consists in two structures characterizing both the conformance and the performance of the traces in $L$.
-Let us call them $ConfResult$, and $PerfResult$.
+\item \label{step1} \textbf{Log replay analysis}: this phase consists of an execution of the log replay with  the model $M$ and the log $L$ as inputs.
+The results consist of two structures characterizing both the conformance and the performance of the traces in $L$.
+Let us call them \emph{ConfResult} and \emph{PerfResult}.
 
-\item \label{step2} \textbf{Training set construction}: this phase aims to build the training set for the classification technique.
-This is done through an execution of the plug-in \texttt{Generate Instances With Conformance} with the event log $L$ and the conformance result $ConfResult$ as inputs.
-Let us call the training set obtaind $TS$.
+\item \label{step2} \textbf{Training set construction}: this phase aims at building the training set for the classification technique.
+This is done through an execution of the plug-in \texttt{Generate Instances with Conformance} with the event log $L$ and the conformance result \emph{ConfResult} as inputs.
+Let us call the obtained training set $TS$.
 
 \item \textbf{Classifier building}: in this step the classification model (decision tree) for the conformance analysis is built using the plug-in \texttt{Generate Classifier} with the training set $TS$ as input.
-Let us call the decision tree obtained $DT$.
+Let us call the obtained decision tree  $DT$.
 
-\item \textbf{Using the classifier}: in this step the goal is to exploit $DT$ and the log replay result $ConfResult$ to establish the corrective measures in case of anomalies.
-Note that both $DT$ and $ConfResult$ can be visualized as in Figures \ref{salesDecTree} and \ref{replayResult} into ProM6 thanks to the plug-ins for resources managing that integrate the framework.
+\item \textbf{Using the classifier}: in this step the goal is to exploit $DT$ and the log replay result \emph{ConfResult} to establish the corrective measures in case of anomalies.
+Note that both $DT$ and \emph{ConfResult} can be visualized as in Figures~\ref{salesDecTree} and~\ref{replayResult} into ProM6 thanks to the plug-ins for resources managing that integrate the framework.
 $DT$ can also be used in a predictive way to establish the conformance of new process instances without needing a log replay execution.
@@
 
-To perform a performance analysis based on the extension presented in Section \ref{ClassPerf}, the same methodology can be followed modifying the step \ref{step2}.
-The construction of the training set is done through a plug-in (analogous to \texttt{Generate Instances With Conformance}) that takes as input $PerfResult$ obtained by the step \ref{step1} and an indication about the synchronization point chosen for the analysis.
+For a  performance analysis based on the extension summarized in Section~\ref{ClassPerf}, the same methodology can be followed modifying step~\ref{step2}.
+The construction of the training set is done through a plug-in (analogous to \texttt{Generate Instances with Conformance}) that takes as input \emph{PerfResult} obtained by step~\ref{step1} and an indication about the synchronization point chosen for the analysis.
 
@@
 
-We presented some preliminary results of a research activity aimed at applying machine learning techniques, i.e.~ classification algorithms, in the realm of process analysis.
-The idea is to try to identify some correlation between the data value of certain variables in the individual cases of event logs, and the results of conformance checking.
-A successful identification of such values could allow to predict the conformance result  of a case during execution.
+We presented some preliminary results of a research activity aimed at applying machine learning techniques, i.e.~classification algorithms, in the realm of process analysis.
+The idea is to try to identify some correlation between the data value of certain variables in the process instances of event logs, and the results of conformance checking.
+A successful identification of such values could allow to predict the conformance result  of an instance during execution.
 The same technique was applied to performance checking as well, but only briefly summarized here because of space constraints.
-The approach is implemented with a combination of plug-ins of the Process Mining framework ProM and of the classification engine WEKA.
+The approach is implemented with a combination of plug-ins of the Process Mining framework ProM and of the classification engine Weka.
 Some guidelines about their use are presented in a concluding section.
@@
 
-Classification techniques in the Process Mining field have already been explored in \cite{DBLP:conf/bpm/RozinatA06} and more recently in \cite{DBLP:conf/sac/LeoniA13}.
+Classification techniques in the Process Mining field have already been explored in~\cite{DBLP:conf/bpm/RozinatA06} and more recently in~\cite{DBLP:conf/sac/LeoniA13}.
 In those papers the  idea is to  explore how data influences the case routing of process flow execution, by assigning a classification problem with each decision point in the model.
-In particular, \cite{DBLP:conf/sac/LeoniA13} improves significantly the results of \cite{DBLP:conf/bpm/RozinatA06} by exploiting some recently developed \emph{alignment techniques} between event log cases and the process model.
+In particular,~\cite{DBLP:conf/sac/LeoniA13} improves significantly the results of~\cite{DBLP:conf/bpm/RozinatA06} by exploiting some recently developed \emph{alignment techniques} between process instances and the process model.
 We intend to explore how far these techniques can be applied fruitfully in our framework as well.

### 2013-11-09 paper/Paper_aibp/camera-ready-paper2.tex
camera-ready copy added

--- 2013-09-20 paper/paper2.tex
@@
 \usepackage{marginnote}
-%\usepackage[outer=5cm, inner=2cm, heightrounded, marginparwidth=4cm, marginparsep=0.5cm]{geometry}
 
@@
 
-\pagestyle{plain}
-
 \title{Classification Techniques for Conformance and Performance Checking in Process Analysis}
 \author{Hind Chfouka\inst{1}, Andrea Corradini\inst{1} and Roberto Guanciale\inst{2}}
-\institute{Department of Computer Science, University of Pisa, Italy\and School of Computer Science and Communication,
-KTH Royal Institute of Technology, Stockholm, Sweden}
+\institute{Department of Computer Science, University of Pisa, Italy \\ 
+\email{\{chfouka,andrea\}@di.unipi.it}\and School of Computer Science and Communication,
+KTH Royal Institute of Technology, Stockholm, Sweden \\
+\email{robertog@kth.se}}	
 \date{today}
@@
 
-%% Classification techniques in the Process Mining field have already been explored~\cite{DBLP:conf/bpm/RozinatA06}\cite{DBLP:conf/sac/LeoniA13}.
-In~\cite{DBLP:conf/bpm/RozinatA06} the idea is exploring how data influences the case routing of process flow execution.
-In this contribution the goal is to strengthen the use of data analysis techniques for the benefit of process analysis.
-This is done by studying how data may influence, in general, the overall process behavior.
-
 After the presentation of a case study in Section~\ref{example}, Section~\ref{Background} introduces some background concepts about process analysis.
@@
 
-\section{The case study: a sale business process}\label{example}
+\section{The Case Study: a Sale Business Process}\label{example}
 A business process can be seen as a collection of activities occurring within an organization that lead to a specific goal.
@@
 }} 
-This business process abstractly represents a fragment of the
-procedure followed in a commercial organization for managing orders
-received from clients. In general this procedure includes several
-activities and each of them involves a specific department of the
-organization. We assume that the sale process begins with the
-notification of an order from a client, so the first activity is
-simply called \emph{Order}. After the order is received, the sale
-process continues with some activities that can be done in parallel,
-since they do not present any dependency. The \emph{FinancialCheck}
-activity represents the financial analysis of the order  (i.e
-verifying the financial situation of the client). Concurrently, the
-\emph{WarehouseCheck} activity checks the availability of merchandise
-requested by the order and, in case of a shortage, starts a supplying procedure.
+This business process abstractly represents a fragment of the procedure followed in a commercial organization for managing orders received from clients.
+In general this procedure includes several activities and each of them involves a specific department of the organization.
+We assume that the sale process begins with the notification of an order from a client, so the first activity is simply called \emph{Order}.
+After the order is received, the sale process continues with some activities that can be done in parallel, since they do not present any dependency.
+The \emph{FinancialCheck} activity represents the financial analysis of the order (i.e verifying the financial situation of the client).
+Concurrently, the \emph{WarehouseCheck} activity checks the availability of merchandise requested by the order and, in case of a shortage, starts a supplying procedure.
 This is done through the activity \emph{ExternalWarehouse}.
@@
 
-%Note that the Petri net of Figure~\ref{pnet} is obtained by transforming a BPMN model using the algorithm described in~\cite{DBLP:conf/wsfm/BruniCFFGS11}, and mapping each activity to a pair of start/end transitions.
-The resulting net has been simplified by removing the unnecessary \emph{invisible transitions} used to encode e.g.
-the \emph{Join} or \emph{Fork} gateway~\cite{DBLP:conf/wsfm/BruniCFFGS11}.
-
 \section{Process Analysis}\label{Background}
-%Process analysis is performed using the Petri net representing the process and an event log recording the related executions, also called \emph{process instances}.
-The basic building blocks of event logs are events.
-An event {\itshape e} can be seen as a pair {\itshape e = (a,t) } representing an action {\itshape a } recorded and the corresponding timestamp {\itshape t}.
-Action and timestamp are denoted respectively by $\alpha(e)$ and $\phi(e)$.
-Events that belong to the same process are grouped into {\itshape traces}.
-Formally, a trace $T$ is a finite sequence of events $T[1],..., T[n]$, such that $\phi(T[i]) \leq \phi(T[i+1])$ for all  $i \in [1,n)$.
-A log $L$ is a set of traces, recording the activities performed by a system during a finite number of process executions.
 Process analysis is performed using the Petri net representing the process and an event log recording the related executions, also called \emph{process instances}.
@@
 Therefore an event $e$ can be seen as a tuple including the action recorded, a corresponding timestamp, and possibly other attributes describing relevant information about the state of the process.
-%in gen pair $e = (a,t)$ representing an action $a$ recorded and the corresponding timestamp $t$.  
-Events that belong to the same process instance are grouped in a {\itshape trace}, where events are ordered 
-% A trace $T$ is a finite sequence of events ordered 
-by timestamp. An \emph{event log} is a set of traces, recording the activities performed by a system during a finite number of process executions.  In this paper the following assumptions are made:
+Events that belong to the same process instance are grouped in a {\itshape trace}, where events are ordered  by timestamp.
+An \emph{event log} is a set of traces, recording the activities performed by a system during a finite number of process executions.
+In this paper the following assumptions are made:
 \begin{itemize}
@@
 The key algorithm exploited to analyze a Petri net model with respect to an event log is the {\itshape log replay} algorithm (see e.g.~\cite{DBLP:journals/is/RozinatA08}).
-%DBLP:journals/widm/AalstAD12,
-%,DBLP:conf/bpm/Munoz-GamaC10
 Given a Petri net model and an event log as input to the algorithm, for each trace in the log, the algorithm starts by placing one token in the start place of the net.
@@
 For each trace, the result of the log replay execution is a map indicating which transitions of the model have been fired for each event in the trace, and additionally which tokens were artificially created and which remained  upon completion of the trace.
-%% sequence on a trace can be represented as an ordered list $R$ of pairs $(tr, i)$, representing that the transition $tr$ has been fired to mimic the event $T[i]$.
 The output of the log replay algorithm  can be used to check the conformance of the traces.
-% and to evaluate some performance metrics.
-
-% The log replay result can be used to evaluate both the conformance and performance of a business process.
 
@@
 Figure~\ref{ConfLog}  presents two different traces, $T$ and $T'$, of the business process presented in Section~\ref{example}, where only the actions and the timestamps are reported.
-The log replay of trace $T$, which is compliant with the Petri net, terminates with a single remaining
-%marking containing one 
-token in the end place \{${p13 \rightarrow 1}$\} and without reporting  missing tokens.
+The log replay of trace $T$, which is compliant with the Petri net, terminates with a single remaining token in the end place \{${p13 \rightarrow 1}$\} and without reporting  missing tokens.
 
@@
 \centering
-\includegraphics[width=0.9\linewidth]
+\includegraphics[width=345pt]
 {./items/logConforme.pdf}
@@
 
-\begin{comment}
-The sequence returned by log replay is:
-\begin{equation}
-\begin{split}
-R'&=\{(Order\_start,1), (Order\_complete,2), (WarehouseCheck\_start,3), \\
-& (FinancialCheck\_start,4), (WarehouseCheck\_complete,5),(t1,6),(t2,6),\\
-& (Notify\_start,6), (Notify\_complete,7)\}
-\end{split}
-\end{equation}
-\end{comment}
-
-\begin{comment}
-\subsection{Performance analysis}\label{PerformanceAnalysis}
-The performance of a process can be analyzed by exploiting the log
-replay results. Since the logs contain timestamps, the log replay can
-be used to compute performance measures of the process my measuring the elapsed time between production and consumption of tokens.
-Performance analysis can be applied only to traces that do not generate missing tokens during the replay since such tokens cannot have time information~\cite{DBLP:conf/wsfm/BruniCFFGS11} .
-During the log replay the following metrics can be computed for each trace and each place:
-\begin{itemize}
-\item sojourn time $(tsj)$ : the time interval between arrival and departure of tokens;
-\item synchronization time $(tsc)$: the time interval between arrival of a token in the place and enabling of a transition in the post-set of the place;
-\item waiting time $(tw)$:  the time interval between enabling of a transition in the post-set of the place and token departure (thus $tsj=tsc+tw $).
-\end{itemize}
-
-The duration of an activity is an important performance measure as well, and it is given by the waiting time of the only place between the start and the complete transition of the activity.
-\end{comment}
-\begin{comment}
-To clarify the metrics evaluated by this technique, as done with conformance analysis, we exploit the Petri net  presented in Section~\ref{example} and the trace T presented in Figure~\ref{ConfLog}.
-The replay starts with firing the transition $Order\_start$ at time $0s$, so a token arrives at $p2$ at time $0s$.
-After $45 min$ the token in $p2$ is consumed and the transition $Order\_complete$ is fired, so $tw(p2)=45min$, and a token arrives at $p3$ and $p4$.
-At the time $1h:5min:25s$ the transition $WarhouseCheck\_start$ is fired, thus $tw(p4)=20min+25sec$ and a token is produced at $p6$.
-At the time $1h:10min:20s$ the transition $FinancialCheck\_start$ is fired and a token is produced at $p5$, so $tw(p3)=25min+20s$.
-After $2h:30min:20s$ from the replay start the transition $Financial\_complete$ is fired, hence we have $tw(p5)=1h+20min$ and a token is produced at $p8$.
-At time $3h:10min:20s$ the firing of the transition $WarehouseCheck\_complete$ is executed, so $tw(p6)=2h+4min+55s$ and a token is produced at $p7$.
-After $15 min$ the invisible transition $t1$ is fired and a token is produced at $p10$.
-At this point the transition $t2$ is enabled, so its firing is now possible and a token is produced at $p11$.
-Notice that $tsc(p8)=40min, tsc(p10)=0s$.
-At the time $3h:25min:20s$ the transition $Notify\_start$ is fired and a token is produced in $p12$, after $4min+4sec$ the transition $Notify\_complete$ is fired and $tw(p12) = 4min+40s$.
-
-It is worth noting that the synchronization time is greater than zero
-only for places which contain in their post-set transitions depending
-from other places. For the Petri net in Figure~\ref{pnet}, the only
-places which can have a non-null synchronization time are $p8$ and
-$p10$. For example, the synchronization time is greater than zero for
-$p8$ in the trace $T$ in Figure~\ref{ConfLog}, representing that, for
-this particular process instance, the branch performing the financial
-checking is faster than the warehouse branch.
-
-\end{comment}
-
-%\section{An approach based on classification for Process Analysis}\label{ClassifApproach}
 \section{Classification for Process Analysis}\label{ClassifApproach}
@@
 
-\subsection{Classification: basic concepts}\label{classifBasicConcept}
+\subsection{Classification: Basic Concepts}\label{classifBasicConcept}
 
@@
 
-A classification model can be useful for both a descriptive and a predictive purpose.
-In fact it provides an explanatory tool to distinguish between objects of different classes, and at the same time it can be used to predict the class label of new records.
-In this way the model acts as a black box that automatically assigns a class label to an attribute set of a new record.
+The general approach for solving a classification problem~\cite{tan2005introduction} consists of providing a \emph{training set} of records, whose class labels are already known, to the learning algorithm in order to build a classification model.
+This model is then applied to a \emph{test set}, which consists of records with unknown class labels.
+This step allows to  estimate the \emph{model accuracy} using one of the standard approaches.
+For example, the accuracy can be represented by the percentage  of records correctly predicted  with respect to the total number of records in the test set.
 
-%purpose: it provides an explanatory tool to distinguish between objects of different classes, and for a predictive purpose: the model can be used to predict the class label of new records.
-In this way the model acts as a black box that automatically assigns a class label to an attribute set of a new record.
-
-There is a general approach to solve a classification problem from a given dataset that is independent from the specific classification model.
-Each classification technique uses a learning algorithm to identify a model that best fits the relationship between the attribute sets and the class labels of the input data.
-The generated model should both fit the input data well, and correctly predict the class labels of new records.
-Finding the right trade-off between this two goals is the most delicate part of a classification technique, and in general of any machine learning task.
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=200pt]
-{./items/classifApproach.pdf}
-\caption{General approach for solving a classification problem.}
-\label{figclassApproach}
-\end{figure}
-
-Figure~\ref{figclassApproach} summarizes the general approach for
-solving a classification problem~\cite{tan2005introduction}. A \emph{training set}
-consisting of records whose class labels are known must be provided to
-the learning algorithm, and it is used to build the classification
-model. This model is then applied to a \emph{test set}, which consists
-of records with unknown class labels.
-This step allows to  estimate the \emph{model accuracy} using one of
-the standard approaches. For example, the accuracy can be represented by the
-percentage 
-%fraction of number the 
-of records correctly predicted 
-with respect to the total number of records in the test set.
-
-Many classification models can be applied to a classification
-problem. For our purpose we choose the \emph{decision tree} model
-since the rules detected by the classification are shown explicitly,
-providing an overview of how process data affects the
-process behavior.
-Furthermore, this choice is encouraged by the simplicity of decision trees.
+Many classification models can be applied to a classification problem.
+For our purpose we choose the \emph{decision tree} model since the rules detected by the classification are shown explicitly, providing an overview of how process data affects the process behavior.
 In a decision tree, a class label (possible value of the target attribute) is assigned to each leaf node.
 The non-terminal nodes, which include the root and the internal nodes, contain attribute test conditions to filter records that have different characteristics.
-\begin{comment}
-In principle, there are exponentially many decision trees that can be constructed from a given set of attributes.
-While some of the trees are more accurate than others, finding the optimal tree is computationally infeasible because of the exponential size of the search space.
-However, efficient algorithms have been developed to induce a reasonably accurate decision tree in a reasonable amount of time.
-These algorithms usually employ a greedy strategy that grows a decision tree by making a series of decisions about which attribute to use for partitioning the data.
-For instance Hunt's algorithm is one of them and it represents the basis of many existing decision tree induction algorithms, including $C4.5$ algorithm used for our purposes.
-\end{comment}
+
 \subsection{Classification for Conformance Checking}\label{ClassConformance}
-In order to identify the possible causes of non-conformance, a
-classification problem can be formulated. The classification dataset
-is extracted from the process data: for each process instance,
-relevant information is  extracted and included as a record in the
-dataset.
-For the process presented in Section~\ref{example}, a record includes 
-the following attributes: an instance identifier, a 
-client identifier, the client typology (new or consolidated client),
-the sales manager responsible for the order, the financial officer who
-conducts the financial evaluation activities, the warehouseman
-responsible for the warehouse checking, the supplying responsible name
-in case of a provision, and finally the result communicated to the
-client for the order issued. Furthemore, as target attribute 
-we associate with each record the conformance result for that instance, 
-which is a boolean value that is extracted from the output of the log replay.
-% process
-%instance provides an additional and important data, 
+In order to identify the possible causes of non-conformance, a classification problem can be formulated.
+The classification dataset is extracted from the process data: for each process instance, relevant information is  extracted and included as a record in the dataset.
+For the process presented in Section~\ref{example}, a record includes  the following attributes: an instance identifier, a  client identifier, the client typology (new or consolidated client), the sales manager responsible for the order, the financial officer who conducts the financial evaluation activities, the warehouseman responsible for the warehouse checking, the supplying responsible name in case of a provision, and finally the result communicated to the client for the order issued.
+Furthemore, as target attribute we associate with each record the conformance result for that instance,  which is a boolean value that is extracted from the output of the log replay.
 All these attributes are discrete and contribute to build the dataset
-presented in Table~\ref{tab:SaleData} of the classification
-problem. In this preliminary work, 
-in order to 
-%Notice that, since our goal is to 
-explore the applicability
-of existing machine learning  techniques to process analysis,
-%rather then the
-%evaluation of a new data mining algorithm, 
-we just used synthetic data
+presented in Table~\ref{tab:SaleData} of the classification problem. In this preliminary work,  in order to 
+explore the applicability of existing machine learning  techniques to process analysis, we just used synthetic data
 generated to emulate process executions.
+Note that for this  running example, attributes selected for the classification task are all focused on the actors of the process activities.
+In general, an attribute can characterize several aspects of an activity.
 
-In order to build the classification problem, the attributes characterizing the
-process are extracted from the log, while the conformance attribute is
-computed using the conformance checking algorithm.
-The Petri net of Figure~\ref{replayResult}  shows the
-output of the conformance analysis executed on the event log:
-edges are labeled with the number of times they were activated,
-places are labeled with the number of remaining and missing tokens.
+In order to build the classification problem, the attributes characterizing the process are extracted from the log, while the conformance attribute is computed using the conformance checking algorithm.
+The Petri net of Figure~\ref{replayResult}  shows the output of the conformance analysis executed on the event log: edges are labeled with the number of times they were activated, places are labeled with the number of remaining and missing tokens.
+
 \begin{figure}[h]
@@
 
-The resulting data highlight  that there are 37 out of 250 process
-executions with conformance errors.
-The presence of 37 missing tokens in $p11$ represents 
-the
-number of times that the log replay  algorithm forced the execution of transition
-\emph{Notify\_start}, even if it was not enabled, in order to mimic a corresponding event 
-in the trace.
+The resulting data highlight  that there are 37 out of 250 process executions with conformance errors.
+The presence of 37 missing tokens in $p11$ represents  the number of times that the log replay  algorithm forced the execution of transition \emph{Notify\_start}, even if it was not enabled, in order to mimic a corresponding event in the trace.
 
@@
 they did not execute the financial check activity.
-
-% This situation is possible only if the invisible transition $t2$ is not enabled.
-Now, if $t2$ is not enabled that means that there are no tokens in the places $p8$ or $p10$.
-But notice that there are 21 remaining tokens in the place $p10$, so the transition $t2$ is not enabled for those $37$ instances because the place $p8$ does not have any token.
-Moreover, the Petri net in Figure~\ref{replayResult} shows 37 remaining tokens in place $p3$, so we can deduce that in those instances financial activities were not performed.
-From these facts, we can conclude that the conformance problems presented in  37 instances are due to a missing execution of the financial activity checking.
-On the other hand, the decision tree reports that the instances with conformance problems are made by consolidate clients.
-
+{
+\renewcommand{\arraystretch}{1.2}
 \begin{table}[!h]
 \scriptsize{
-\centering
-\begin{tabular}{|p{1cm}|p{1cm}|p{1,5cm}|p{1,2cm}|p{1,4cm}|p{1,2cm}|p{1,5cm}|p{1,4cm}|p{0,8cm}|}
+\caption{Dataset for the conformance analysis.}
+
+\begin{tabular}{|p{1cm}|p{1cm}|p{1,5cm}|p{1,3cm}|p{1,3cm}|p{1,2cm}|p{1,5cm}|p{1,4cm}|p{0,8cm}|}
 \hline OrdIde & CltIde & CltType & SalMan & FinOff & WrhsMan & SupplyResp & OrdResut & Conf\\
@@
 }
-\caption{Dataset for the conformance analysis.}
 \label{tab:SaleData}
@@
 The resulting decision tree describes a data pattern in correspondence to which a process instance could present conformance errors: order managed by the sales manager \emph{Mary} and received from consolidate clients of the organization may not respect the standard sale procedure.
-%To get more significant information for our analysis,
 
-From the analysis done for the sale process based on both
-classification and log replay results, a new scenario of the sale
-procedure has emerged: orders done by the consolidate clients of the
-organization are not checked from a financial point of view. This fact
-can be accepted as a valid  behavior in a sale procedure, and consequently
-an extension of the business process model is needed to include the
-new scenario (\emph{model extension activity}). Alternatively, the
-misbehaving scenario can be considered as an anomaly in the sale
-procedure. In this case, since the analysis provides accurate
-information about the conformance error, corrective measures can be
-taken in order to prevent such errors in the future.
+From the analysis done for the sale process based on both classification and log replay results, a new scenario of the  sale procedure has emerged: orders done by the consolidate clients of the organization are not checked from a financial point of view.
+This fact can be accepted as a valid  behavior in a sale procedure, and consequently an extension of the business process model is needed to include the new scenario (\emph{model extension activity}).
+Alternatively, the misbehaving scenario can be considered as an anomaly in the sale procedure.
+In this case, since the analysis provides accurate information about the conformance error, corrective measures can be taken in order to prevent such errors in the future.
 
-The classifier constructed with this approach can be used also in a
-predictive way. Given a trace recording a process execution the
-decision tree can be used to predict the conformance result. This
-brings an advantage in terms of the time required by the analysis, since
-the replay of a process instance  takes more time than its 
-classification using a decision
-tree. Moreover, it is worth noting that whenever the set of attributes
-needed for the classification is known before the completion of a
-process execution, the non-conformances predicted by the decision tree can
-be used to promptly alert the stakeholders and activate proper
-countermeasures.
+The classifier constructed with this approach can be used also in a predictive way.
+Given a trace recording a process execution the decision tree can be used to predict the conformance result.
+This brings an advantage in terms of the time required by the analysis, since the replay of a process instance  takes more time than its  classification using a decision tree.
+Moreover, it is worth noting that whenever the set of attributes needed for the classification is  known before the completion of a process execution, the non-conformances predicted by the decision tree can be used to  promptly alert the stakeholders and activate proper countermeasures.
 
@@
 
-\begin{comment}
-\item \texttt{Generate Instances with Performance}: the plug-in takes as input an event log, an indication of the synchronization point to examine and the performance result computed by the log replay.
-The output is a training set as presented in Table~\ref{tab:SaleDataPerf}.
-\end{comment}
-
 \item \texttt{Generate Instances to Classify}: given an event log, the plug-in generates a dataset of instances that can be classified using a classification model.
@@
 
-\subsection{Methodology}
-This last paragraph is dedicated to a description of the methodology to be followed to analyze a business process under the approach presented in this paper.
+Figure \ref{dataflow} represents a data flow diagram that summarizes data dependencies between plug-ins discussed in this section.
 
-It should be clear that the analysis presented in the previous sections assumes the presence of a Petri net model of the business process to be analyzed.
-The model can be expressed in BPMN  as well, but in this case a transformation into a Petri net can be done through the techniques described in~\cite{DBLP:conf/wsfm/BruniCFFGS11}.
-The classification analysis  needs also an event log tracing the process behavior over multiple instances.
-Given a process model $M$ and an event log $L$, the methodology to be followed for the conformance analysis with the approach presented in Section~\ref{ClassConformance} can be summarized by these steps:
-\begin{enumerate}
-
-\item \label{step1} \textbf{Log replay analysis}: this phase consists of an execution of the log replay with  the model $M$ and the log $L$ as inputs.
-The results consist of two structures characterizing both the conformance and the performance of the traces in $L$.
-Let us call them \emph{ConfResult} and \emph{PerfResult}.
-
-\item \label{step2} \textbf{Training set construction}: this phase aims at building the training set for the classification technique.
-This is done through an execution of the plug-in \texttt{Generate Instances with Conformance} with the event log $L$ and the conformance result \emph{ConfResult} as inputs.
-Let us call the obtained training set $TS$.
-
-\item \textbf{Classifier building}: in this step the classification model (decision tree) for the conformance analysis is built using the plug-in \texttt{Generate Classifier} with the training set $TS$ as input.
-Let us call the obtained decision tree  $DT$.
-
-\item \textbf{Using the classifier}: in this step the goal is to exploit $DT$ and the log replay result \emph{ConfResult} to establish the corrective measures in case of anomalies.
-Note that both $DT$ and \emph{ConfResult} can be visualized as in Figures~\ref{salesDecTree} and~\ref{replayResult} into ProM6 thanks to the plug-ins for resources managing that integrate the framework.
-$DT$ can also be used in a predictive way to establish the conformance of new process instances without needing a log replay execution.
-Given an event log containing new process instances, a dataset can be built through the plug-in \texttt{Generate Instances to Classify} and then classified by the plug-in \texttt{Classify Instances} according to the model $DT$.
-\end{enumerate}
-
-For a  performance analysis based on the extension summarized in Section~\ref{ClassPerf}, the same methodology can be followed modifying step~\ref{step2}.
-The construction of the training set is done through a plug-in (analogous to \texttt{Generate Instances with Conformance}) that takes as input \emph{PerfResult} obtained by step~\ref{step1} and an indication about the synchronization point chosen for the analysis.
+\begin{figure}[h]
+\includegraphics[width=360pt]
+{./items/dataflow.pdf}
+\caption{Data flow diagram}
+\label{dataflow}
+\end{figure}
 

### 2013-12-02 paper/Paper_aibp/camera-ready-paper2.tex
aggiunta versione finale paper e presentazione

@@
 First a few basic concepts about the classification technique are described, next the conformance analysis based on classification is presented in detail through an illustration with our case study.
-A possible extension of the approach for performance analysis is briefly discussed presented.
-Finally, Section~\ref{implementation} describes the implementation of our approach provided as \emph{ProM6}\cite{DBLP:conf/bpm/AalstDGRVW09} plug-ins, as well the methodology to follow in order to analyse processes using our approach.
-
+A possible extension of the approach for performance analysis is briefly discussed.
+Finally, Section~\ref{implementation} describes the implementation of our approach provided as \emph{ProM6}\cite{DBLP:conf/bpm/AalstDGRVW09} plug-ins.
 \section{The Case Study: a Sale Business Process}\label{example}
@@
 
-The general approach for solving a classification problem~\cite{tan2005introduction} consists of providing a \emph{training set} of records, whose class labels are already known, to the learning algorithm in order to build a classification model.
+The general approach for solving a classification problem consists of providing a \emph{training set} of records, whose class labels are already known, to the learning algorithm in order to build a classification model.
 This model is then applied to a \emph{test set}, which consists of records with unknown class labels.
@@
 generated to emulate process executions.
-Note that for this  running example, attributes selected for the classification task are all focused on the actors of the process activities.
-In general, an attribute can characterize several aspects of an activity.
+Note that for this  running example, attributes selected for the classification task are all focused on the actors of the process activities, but in general attributes can characterize several other aspects of activities.
 
@@
 In order to experiment with some business process prototypes, the approach presented in this paper has been implemented as a set of plug-ins that integrates \emph{ProM6}~\cite{DBLP:conf/bpm/AalstDGRVW09}, an open-source framework implementing Process Mining tools, and \emph{Weka}~\cite{DBLP:journals/sigkdd/HallFHPRW09}, a data mining framework providing machine learning tools.
-We can classify the plug-ins developed in three categories. The first one includes plug-ins for building the dataset needed for the classification:
+We can classify the developed plug-ins in three categories. The first one includes plug-ins for building the dataset needed for the classification:
 \begin{itemize}
@@
 Note that the plug-in extracts \emph{all} the attributes data recorded in the event log.
-If necessary, the dataset resulting can be subject to some \emph{feature selection} techniques before using it as a training set for the classification task.
+If necessary, the resulting dataset can be subject to some \emph{feature selection} techniques before using it as a training set for the classification task.
 
@@
 \end{itemize}
-The second category includes plug-ins for the classifier generation and utilization as a predictive model:
+The second category includes plug-ins for generating the classifier and for using it as a predictive model:
 \begin{itemize}
@@
 
-Figure \ref{dataflow} represents a data flow diagram that summarizes data dependencies between plug-ins discussed in this section.
+Figure \ref{dataflow} represents a data flow diagram that summarizes the data dependencies among the plug-ins discussed in this section.
 
@@
 {./items/dataflow.pdf}
-\caption{Data flow diagram}
+\caption{Data dependencies among the implemented plug-ins.}
 \label{dataflow}
@@
 The approach is implemented with a combination of plug-ins of the Process Mining framework ProM and of the classification engine Weka.
-Some guidelines about their use are presented in a concluding section.
 The approach has been successfully tested on synthetic event logs only.

### 2013-12-02 paper/Paper_aibp/presentation/presentazione_aibp.tex
aggiunta versione finale paper e presentazione

\documentclass[10pt]{beamer}
\usetheme{Frankfurt}
\usepackage[british,english]{babel}
\usepackage[latin1,utf8x]{inputenc}
\setbeamercovered{transparent}

\title{Classification Techniques for Process Analysis}
\author{Hind Chfouka\inst{1}, Andrea Corradini\inst{1} and Roberto Guanciale\inst{2}}
\institute{\inst{1} Department of Computer Science, University of Pisa, Italy
\and \inst{2} KTH Royal Institute of Technology, Stockholm, Sweden }	
\makeatletter\date{December 6 2013, Turin}\let\Date\@date
\makeatother

\begin{document}
\begin{frame}
\maketitle
\center{
AI meets Business Processes Workshop \\ 
XIII Conference of the Italian Association for Artificial Intelligence}

\end{frame}

\section{Introduction}

\begin{frame}
\frametitle{Context: Process Mining}

\begin{figure}
\includegraphics[width=280pt]{./item/fig1.pdf}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Focus: Process Analysis}
We focus on: Process Analysis
\begin{figure}
\includegraphics[width=280pt]{./item/fig1Annotated.pdf}
\caption{Process mining}
\end{figure}

\end{frame}

\section{Running example}
\begin{frame}
\frametitle{An example: Sale Business Process}
\begin{figure}
\includegraphics[width=280pt]{./item/SaleProcessModels.pdf}
\caption{Sale Business Process}
\end{figure}
\end{frame}

\section{Process Analysis}
\begin{frame}
\frametitle{ Process Analysis }
\begin{figure}
\includegraphics[width=300pt]{./item/analysis.pdf}
\end{figure}

\end{frame}

\begin{frame}
\frametitle{Log Replay Algorithm}
	\begin{block}{Assumptions}
		\begin{itemize}
		\item Event $e =(a,t,atts)$: the event log building block
		\item Trace: a finite sequence of events $T[1],..., T[n]$ ordered by timestamp. A trace represents a process instance
		\item Event log: a finite sequence of traces
		\item Each event of a trace can be mapped into a transition of the Petri net model
		\end{itemize}
	\end{block}
	\begin{block}{{Algorithm}}
		\begin{itemize}
			\item \alert{Log replay}: executes traces of an event log in a non-blocking way
					\begin{enumerate}
					\item Starts with a token in the start place
					\item Extracts the top event of the log
					\item Fires the corresponding transition in the current marking of the net
						\begin{itemize}
						\item If the transition is not enabled creates the missing tokens artificially
						\end{itemize}
				\end{enumerate}
		\item Log replay results are used in conformance and performance checking
		\end{itemize}
	\end{block}
\end{frame}

\begin{frame}
\frametitle{Conformance Analysis with log replay}
\begin{block}{}
	Conformance checking = check if a trace is compliant with the Petri net model
	\end{block}
	\smallskip
	\smallskip
	\begin{block}{Based on the log replay results...}
		\begin{itemize}
			\item \alert{missing tokens} are generated to mimic an event with a corresponding transition not enabled: this indicates a non-compliance to the model
		\end{itemize}
	\end{block}
\smallskip
\smallskip
\begin{figure}
\includegraphics[width=300pt]{./item/conformanceAnalysis.pdf}
\end{figure}
\end{frame}

\section{Machine Learning for Process Analysis}

\begin{frame}
\frametitle{ML for Process Analysis}
\begin{block}{}
Event logs are huge and rich of data: this encourages use of ML techniques\\
\smallskip
Several approaches exploiting Machine Learning techniques for the Business Process understanding:
\begin{itemize}
\item To extract the process model
\item To find rules associated with a decision point
\item To extract implicit information from the data process instances with Data Mining tools
\item ....
\end{itemize} 
\end{block}

\smallskip
\smallskip
\smallskip
\smallskip
\begin{block}{{Our idea}}
Exploiting ML techniques to discover how the {\color{red} process instance data} may influence its {\color{red} conformance}.\\
\end{block}
\end{frame}

\begin{frame}
\frametitle{Classification for Conformance Checking}
Conformance checking problem can be seen as a Classification problem:

\begin{figure}
\includegraphics[width=250pt]{./item/classification.pdf}
\end{figure}

Why?
\begin{itemize}
\item To find out patterns of data in correspondence of which conformance errors occur
\item To Predict conformance result at run-time. 
\end{itemize}
How?
\begin{itemize}
\item Learning from previous analysis.
\item Using an explicit classifier: Decision Tree.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{The approach}
\begin{itemize}
\item {\color{red} Step 1}: collecting a dataset based on event logs
\item {\color{blue} Step 2}: dataset preprocessing including feature selection
\item {\color{green} Step 3}: building a decision tree model using ML algorithm
\item {\color{purple}Step 4}: using the classifier to predict conformance result
\end{itemize}
\smallskip
\smallskip\smallskip
\begin{figure}
\includegraphics[width=300pt]{./item/methodology.pdf}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Example: classification for the sale process}
A Petri net summarizing the results of the log replay execution on an event log $L$:
\begin{figure}
\includegraphics[width=280pt]{./item/Sales_PN_resultMod.pdf}
\end{figure}
37 instances of the process are not compliant with the sale policy since they did not execute the financial check
activity.
\end{frame}

\begin{frame}
\frametitle{Example: classification for the sale process (cont.)}
Log replay conformance results and process data extracted from the event log $L$ enable:
\begin{itemize}
\item the construction of a dataset.
\item the mining of a decision tree.
\end{itemize}
\begin{figure}
\includegraphics[width=290pt]{./item/casestudy.pdf}
\end{figure}

%Orders issued by consolidated clients and handled by a particular financial officer may not respects the standard sale procedure.
\end{frame}

\section{Framework for the analysis}
\begin{frame}
\frametitle{Framework ProM6}
\begin{figure}
\includegraphics[width=150pt]{./item/prom.pdf}
\end{figure}
\begin{block}{ProM}
\begin{itemize}
\item ProM is an extensible framework that supports a large variety of process mining and analysis techniques.
\item It is a modular software implemented in Java and distributed under GNU Public License (GPL).
\item ProM is a project of Process Mining Group of Eindhoven Technical University, Netherlands.
\end{itemize}
\end{block}
\end{frame}

\begin{frame}
\frametitle{Framework for the analysis}
\begin{figure}
\includegraphics[width=300pt]{./item/worklow.pdf}
\end{figure}
\end{frame}

\section{Conclusion and future works}
\begin{frame}
\frametitle{Conclusion and future works}
Conclusions:
\begin{itemize}
\item Preliminary research aimed at applying ML techniques in the Process Analysis.
\item Extension to performance checking.
\item Experimentation done only with synthetic data.
\end{itemize}
\smallskip
\smallskip
\smallskip
Future Work:
\begin{itemize}
\item Experimentations with real events logs.
\item Exploration of a new technique of conformance checking based on event log alignment.
\end{itemize}

\end{frame}

\end{document}

\begin{comment}
\begin{frame}
\frametitle{Classification Technique}

Few words about the well known machine learning task: classification
\end{frame}
\end{comment}

\begin{frame}

\begin{frame}
\frametitle{Classification for Performance Analysis: an extension}

the approach could be extended to performance analysis:\\
- each join gateway identifies a classification problem
- metrics to be considered as target can be: synchronization time supposing a previous discretization step.

\end{frame}
